---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# svyplan

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/svyplan)](https://CRAN.R-project.org/package=svyplan)
[![R-CMD-check](https://gitlab.com/dickoa/svyplan/badges/main/pipeline.svg)](https://gitlab.com/dickoa/svyplan/-/pipelines)
[![Codecov test coverage](https://codecov.io/gl/dickoa/svyplan/branch/main/graph/badge.svg)](https://app.codecov.io/gl/dickoa/svyplan?branch=main)
<!-- badges: end -->

Survey sample size determination, precision analysis, optimal and joint
multivariate/multidomain allocation, stratification, and power analysis for R.

## Installation

```r
# From GitLab
pak::pkg_install("gitlab::dickoa/svyplan")
```

## Sample sizes

```{r}
library(svyplan)

# Proportion with margin of error
n_prop(p = 0.3, moe = 0.05)

# Mean with finite population and design effect
n_mean(var = 100, moe = 2, N = 5000, deff = 1.5)
```

### Response rate adjustment

Most sizing and precision functions accept `resp_rate`. In sample-size
mode, the required sample is inflated by `1 / resp_rate` to account for
expected non-response:

```{r}
n_prop(p = 0.3, moe = 0.05, deff = 1.5, resp_rate = 0.8)
```

### Survey plan profiles

When the same design parameters apply across many calls, bundle them into
a `svyplan()` profile:

```{r}
plan <- svyplan(deff = 1.5, resp_rate = 0.85, N = 50000)

# Pass as argument
n_prop(p = 0.3, moe = 0.05, plan = plan)

# Or pipe (positional or named args)
plan |> n_mean(100, moe = 2)
plan |> n_mean(var = 100, moe = 2)

# Explicit args always override plan defaults
n_prop(p = 0.3, moe = 0.05, plan = plan, deff = 2.0)
```

## Precision analysis

Given a sample size, how precise will your estimates be? The `prec_*()`
functions are the inverse of `n_*()`:

```{r}
prec_prop(p = 0.3, n = 400)

prec_mean(var = 100, n = 400, mu = 50)
```

### Round-trip between size and precision

All `n_*()` and `prec_*()` functions are S3 generics. Pass a precision
result to `n_*()` to recover the sample size, or pass a sample size result
to `prec_*()` to compute the achieved precision:

```{r}
# Start with a precision target
s <- n_prop(p = 0.3, moe = 0.05, deff = 1.5)

# What precision does this n achieve?
p <- prec_prop(s)
p

# Recover the original n
n_prop(p)
```

## Multi-indicator surveys

Household surveys track many indicators at once. `n_multi()` finds the
sample size that satisfies all precision targets simultaneously.

```{r}
targets <- data.frame(
  name = c("stunting", "vaccination", "anemia"),
  p    = c(0.25, 0.70, 0.12),
  moe  = c(0.05, 0.05, 0.03),
  deff = c(2.0, 1.5, 2.5)
)

n_multi(targets)
```

Per-domain optimization works by specifying domain columns via the
`domains` parameter.

### MICS/DHS-style relative margin of error

Programmes like UNICEF MICS and DHS express precision as a **relative
margin of error** (RME = MOE / p). To use this with svyplan, convert to
an absolute margin of error: `moe = RME * p`.

```{r}
# RME = 12% for each indicator
rme <- 0.12
targets_rme <- data.frame(
  name = c("stunting", "vaccination", "anemia"),
  p    = c(0.25, 0.70, 0.12),
  deff = c(2.0, 1.5, 2.5)
)
targets_rme$moe <- rme * targets_rme$p

n_multi(targets_rme)
```

The MICS template reports sample size in **households**. svyplan returns
the number of **individuals** in the target population. To convert,
divide by the expected number of eligible individuals per household:
`n_hh = ceiling(n / (pb * hh_size))`, where `pb` is the share of the
target population and `hh_size` is the average household size.

## Multistage cluster designs

```{r}
# Optimal 2-stage allocation within a budget
n_cluster(stage_cost = c(500, 50), icc = 0.05, budget = 100000)

# Precision for a given allocation
prec_cluster(n = c(50, 12), icc = 0.05)
```

Variance components can be estimated from frame data and passed
directly to `n_cluster()`:

```{r}
set.seed(104)
frame <- data.frame(
  district = rep(1:40, each = 20),
  income = rep(rnorm(40, 500, 100), each = 20) + rnorm(800, 0, 50)
)

vc <- varcomp(income ~ district, data = frame)
vc
as.data.frame(vc)

n_cluster(stage_cost = c(500, 50), icc = vc, cv = 0.05)
```

`icc` is the survey-planning measure of homogeneity used by
`varcomp()`, `n_cluster()`, and `design_effect()`. It is not the same as
a generic mixed-model ICC, and values near 0 or 1 correspond to
degenerate boundary cases for the closed-form cluster optimizer.

## Sensitivity analysis

`predict()` evaluates a result at new parameter combinations, returning
a data frame suitable for plotting:

```{r}
x <- n_prop(p = 0.3, moe = 0.05, deff = 1.5)
predict(x, expand.grid(
  deff = c(1, 1.5, 2, 2.5),
  resp_rate = c(0.7, 0.8, 0.9, 1.0)
))
```

Sensitivity analysis is available for single-indicator sample-size and
precision results, cluster designs, power analyses, and strata boundaries.
Multi-indicator results are not currently supported by `predict()`.

## Strata boundaries

`strata_bound()` constructs candidate boundaries for a continuous
stratification variable.

```{r}
set.seed(905)
x <- rlnorm(5000, meanlog = 6, sdlog = 1.2)

strata_bound(x, n_strata = 4, n = 300, method = "cumrootf")
```

Four methods are available: Dalenius-Hodges (`"cumrootf"`), geometric
(`"geo"`), LH-inspired coordinate optimization (`"lh"`), and
Kozak-inspired random-restart local search (`"kozak"`). The latter two are
heuristics and do not claim global optimality or exact implementation of the
published algorithms.

## Two-phase designs

A large cheap phase 1, then a subsample measured on the expensive
variable. `n_twophase()` allocates both at once. The frame is one row per
phase-2 stratum, in the `n_alloc()` column vocabulary but on a narrower
contract: `sd` is required where `n_alloc()` also accepts `var`.

```{r twophase}
frame <- data.frame(
  stratum   = c("A", "B", "C", "D"),
  N         = c(3500, 2500, 2500, 1500),
  sd        = c(12, 25, 8, 40),
  mean      = c(40, 70, 35, 90),
  unit_cost = c(2, 5, 1, 9)
)

n_twophase(frame, phase1_cost = 1, budget = 50000)
```

Nonresponse follow-up is the same problem with two strata: respondents
are already measured, so they cost nothing more and are all kept, and
only the nonrespondents are subsampled.

```{r twophase-nrfu}
theta <- 0.5

nrfu <- data.frame(
  stratum   = c("respondents", "nonrespondents"),
  N         = c(theta, 1 - theta),
  sd        = c(1, 1),
  unit_cost = c(0, 200),
  take_all  = c(TRUE, FALSE)
)

n_twophase(nrfu, phase1_cost = 50, budget = 100000,
           mu = 1, single_cost = 50 / theta)
```

Neither phase has to be a simple random sample, and both can lose sample
to nonresponse. `phase1_deff` and `resp_rate` act on the between-stratum
component; `deff` and `resp_rate` frame columns act on each stratum's
residual; `single_deff` and `single_resp_rate` describe the comparator.
Each divides its own component, so a single factor on the combined
variance would be the wrong model. Issued and expected-responding counts
are reported separately.

Every result carries the single-phase comparison, because two-phase
sampling is not always an improvement.

## Power analysis

Solve for sample size, power, or minimum detectable effect. Supports
design effects, finite population correction, response rate adjustment,
panel overlap, unequal groups, and allocation ratios. Arcsine and
log-odds methods available for rare proportions.

```{r}
# Sample size to detect a 5pp change from 70% with deff = 2
power_prop(p1 = 0.70, p2 = 0.75, deff = 2.0)

# MDE with n = 1500 per group
power_prop(p1 = 0.70, n = 1500, deff = 2.0)

# Means
power_mean(200, effect = 5)

# Arcsine method for rare proportions
power_prop(p1 = 0.15, p2 = 0.18, alternative = "one.sided",
           method = "arcsine")

# Difference-in-differences
power_did(treat = c(0.50, 0.55), control = c(0.50, 0.48),
          outcome = "prop", effect = 0.07)
```

`plot()` draws the power-vs-sample-size curve with reference lines at the solved point:

```{r power-plot, fig.width=7, fig.height=4, fig.alt="Power increases with total sample size. Dashed reference lines mark 80 percent power at the required sample size for detecting a change from 70 to 75 percent with a design effect of 2."}
pw <- power_prop(p1 = 0.70, p2 = 0.75, power = 0.80, deff = 2.0)
plot(pw)
```


## Stratified allocation

Given a sampling frame with stratum sizes and variabilities, `n_alloc()`
distributes the total sample across strata:

```{r}
frame <- data.frame(
  N = c(4000, 3000, 3000),
  sd = c(10, 15, 8),
  mean = c(50, 60, 55)
)

n_alloc(frame, n = 600, alloc = "neyman")
```

Constraints and alternative solve modes are also supported:

```{r}
frame_constraints <- transform(
  frame,
  unit_cost = c(1, 1.5, 1),
  max_weight = c(25, 20, NA),
  take_all = c(FALSE, FALSE, TRUE)
)

# Budget-constrained allocation with weight and take-all constraints
n_alloc(frame_constraints, budget = 3500, alloc = "optimal", min_n_stratum = 40)
```

Domain-level CV targets can be enforced via the `domains` parameter:

```{r}
frame_domains <- data.frame(
  province = c("North", "North", "South", "South"),
  stratum = c("Urban", "Rural", "Urban", "Rural"),
  N = c(2000, 3000, 1800, 3200),
  sd = c(12, 18, 10, 16),
  mean = c(55, 48, 58, 50)
)

# Minimum total n such that each province meets the CV target
n_alloc(frame_domains, domains = "province",
        cv = 0.04, alloc = "power", alloc_q = 0.3)
```

For several indicators and overlapping domains, pass long `measures` and
`targets` tables. The allocation rule is then determined jointly by the
precision requirements:

```{r joint-allocation}
joint_frame <- data.frame(
  stratum = c("North urban", "North rural", "South urban", "South rural"),
  region = rep(c("North", "South"), each = 2),
  residence = rep(c("Urban", "Rural"), 2),
  N = c(1000, 1800, 1200, 1600),
  unit_cost = c(1, 1.2, 1.4, 1.1)
)
joint_measures <- data.frame(
  stratum = rep(joint_frame$stratum, 2),
  name = rep(c("coverage", "income"), each = 4),
  p = c(0.50, 0.40, 0.60, 0.35, rep(NA, 4)),
  mean = c(rep(NA, 4), 50, 55, 60, 48),
  sd = c(rep(NA, 4), 10, 12, 15, 9)
)
joint_targets <- data.frame(
  name = c("coverage", "coverage", "income"),
  domain = c(".overall", "region", "residence"),
  level = c(NA, "North", "Urban"),
  cv = c(0.05, 0.08, NA),
  moe = c(NA, NA, 2)
)

joint_fit <- n_alloc(
  joint_frame, measures = joint_measures, targets = joint_targets
)
joint_fit
```

The fitted object distinguishes the certified continuous optimum from a
feasible integer operational recommendation. Use `prec_alloc(joint_fit)` or
`prec_alloc(joint_fit, n = joint_fit$detail$n_int)` to inspect either design.

The same API handles fixed-take multistage designs. Stage populations, fixed
takes, and costs belong in `frame`; indicator-specific homogeneity parameters
belong in `measures`. Here only the PSU counts are optimized:

```{r joint-allocation-two-stage}
cluster_frame <- within(joint_frame, {
  unit_cost <- NULL
  N_psu <- c(100, 150, 100, 120)
  n_per_psu <- c(8, 10, 12, 8)
  cost_psu <- c(300, 400, 450, 350)
  cost_ssu <- c(25, 30, 35, 28)
})
cluster_measures <- transform(
  joint_measures,
  icc_psu = rep(c(0.03, 0.05, 0.08, 0.04), 2),
  var_ratio_psu = 1
)
cluster_fit <- n_alloc(
  cluster_frame, measures = cluster_measures, targets = joint_targets
)
cluster_fit$detail[, c("stratum", "n_psu_int", "n_per_psu", "n_int")]
```

Public `n` remains the ultimate-unit sample size. Thus the field design obeys
`n_int = n_psu_int * n_per_psu`; three-stage designs additionally multiply by
the fixed `n_per_ssu`.

`prec_alloc()` computes the precision for a given allocation (inverse of
`n_alloc()`).

Adding a `icc_psu` column to the frame (for example from `varcomp()`
with its `strata` argument) turns the allocation into a stratified
two-stage design, with cost-optimal or fixed cluster takes per stratum.
See `?n_alloc` and the vignette for the full workflow.

## Design effects

`design_effect()` anticipates the design effect of a plan by combining the
design features you are choosing. Components multiply, and each is selected
by the arguments you supply.

```{r}
# Clustering alone: 20 households per cluster
design_effect(icc = 0.05, n_per_psu = 20)

# Clustering, unequal weighting, and the stratification gain together
frame <- data.frame(
  N = c(50000, 120000), n = c(600, 400), sd = c(12, 20), mean = c(55, 48)
)
deff <- design_effect(icc = 0.05, n_per_psu = 20, strata = frame)
deff

# Use it wherever a deff is expected
n_prop(p = 0.3, moe = 0.05, deff = deff)
effective_n(deff, n = 1000)

# Or read the features off a plan you already built
design_effect(n_cluster(stage_cost = c(500, 50), icc = 0.05, cv = 0.05))
```

This is a planning tool. To measure the design effect a *collected* sample
actually achieved, use `survey::svymean(..., deff = TRUE)`, which computes
it from the realized weights, strata, and clusters.

## References

Cochran, W. G. (1977). *Sampling Techniques* (3rd ed.). Wiley.

Kish, L. (1965). *Survey Sampling*. Wiley.

Valliant, R., Dever, J. A., and Kreuter, F. (2018).
*Practical Tools for Designing and Weighting Survey Samples*
(2nd ed.). Springer.
