---
title: "Getting started with svyplan"
description: >
  Calculate a survey sample size, check its expected precision, and compare
  planning assumptions.
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Getting started with svyplan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
```

`svyplan` helps you plan survey sample sizes, allocate a sample across a
design, and assess expected precision or power. You supply anticipated outcome
variability, design features, and response assumptions. This vignette shows
how to calculate a sample size, check its precision, and compare planning
assumptions.

The calculations describe a design before data collection. They do not select
sample records or replace design-based analysis of the collected data.

## Plan a coverage estimate

Suppose a survey must estimate vaccination coverage. A previous survey
suggests that 70% of eligible people are vaccinated. The new survey should
have a 95% confidence interval with a margin of error no greater than 5
percentage points.

Load the package and calculate the sample size:

```{r first-size}
library(svyplan)

coverage <- n_prop(p = 0.70, moe = 0.05)
coverage
```

Here `p` is the anticipated proportion and `moe` is the desired interval
half-width. The result keeps a continuous calculation in `$n`. Printing and
`as.integer()` give the whole-unit sample to field:

```{r exact-and-operational}
c(
  continuous = as.double(coverage),
  fielded = as.integer(coverage)
)
```

Use the whole-unit value for fieldwork. Keeping the continuous value inside
the object avoids accumulating rounding differences when the result is passed
to another calculation.

## Add design and response assumptions

The first calculation assumes simple random sampling and complete response.
Suppose earlier rounds suggest a design effect of 1.5 and an 85% response
rate:

```{r design-assumptions}
coverage <- n_prop(
  p = 0.70,
  moe = 0.05,
  deff = 1.5,
  resp_rate = 0.85
)
coverage
```

The design effect increases the variance relative to simple random sampling.
The response-rate adjustment increases the issued sample so the expected
number of respondents meets the precision target. It assumes response is
ignorable under the planned adjustment. It does not remove nonresponse bias
or account for variation from response weights.

A finite population size can also be supplied with `N`. The finite population
correction matters when the planned sampling fraction is appreciable:

```{r finite-population}
n_prop(
  p = 0.70,
  moe = 0.05,
  N = 5000,
  deff = 1.5,
  resp_rate = 0.85
)
```

The single-stage `n_*()` and `prec_*()` functions share these arguments:

| Argument | Meaning | Default |
|:---|:---|:---|
| `deff` | Design effect multiplying the simple random sampling variance | 1 |
| `N` | Finite population size, which applies the finite population correction | `Inf` |
| `alpha` | Significance level of `moe` and `rmoe` targets and of intervals | 0.05 |
| `resp_rate` | Expected response rate. The issued sample is the respondent sample divided by it | 1 |

Any `deff > 0` is accepted. A design effect below 1 describes a design more
efficient than simple random sampling, such as a well-stratified sample.
`n_cluster()` takes `icc` and stage costs in place of `deff` and `N`, and
accepts a response rate for each stage.

## Check expected precision

Every core `n_*()` function has a paired `prec_*()` function. Passing the
sample-size result carries its assumptions into the precision calculation:

```{r precision-check}
coverage_precision <- prec_prop(coverage)
coverage_precision
confint(coverage)
```

The interval is the interval expected at the planning value of 70%. It is not
an interval estimated from survey data. Under the same method and continuous
design assumptions, passing the precision result back to `n_prop()` recovers
the original calculation:

```{r round-trip}
n_prop(coverage_precision)
```

Operational rounding and constrained allocations can make a fielded design
slightly more precise than its continuous target. In those cases, assess the
reported operational allocation rather than expecting exact numerical
reversal.

## Compare uncertain assumptions

Response rates and design effects are usually estimates. `predict()` evaluates
supported planning results at new parameter combinations while holding the
other assumptions fixed:

```{r sensitivity}
predict(
  coverage,
  expand.grid(
    deff = c(1.2, 1.5, 2.0),
    resp_rate = c(0.70, 0.85, 1.00)
  )
)
```

This grid shows the sample-size consequences of optimistic and conservative
assumptions. See `?predict.svyplan` for the parameters supported by each result
class. For example, prediction from a `strata_bound()` result assigns new
observations to its existing strata. It does not rerun a sensitivity grid.

## Reuse shared assumptions

When several calculations use the same population and fieldwork assumptions,
store them in a plan profile:

```{r plan-profile}
design <- svyplan(
  deff = 1.5,
  resp_rate = 0.85,
  N = 5000
)

n_prop(p = 0.70, moe = 0.05, plan = design)
n_mean(var = 100, moe = 2, plan = design)
```

An explicit function argument overrides the corresponding profile value. This
lets a common profile hold the defaults while an indicator supplies its own
design effect or response rate:

```{r plan-override}
n_prop(p = 0.70, moe = 0.05, plan = design, deff = 2)
```

Profiles can also be piped into functions whose first argument accepts them:

```{r plan-pipe}
design |> n_prop(p = 0.70, moe = 0.05)
```

A profile can also carry cluster context. Stage costs and the intraclass
correlation stored in the profile are read by `n_cluster()`:

```{r plan-cluster}
cluster_design <- svyplan(
  stage_cost = c(500, 50),
  icc = 0.05,
  resp_rate = 0.85
)

n_cluster(cv = 0.05, plan = cluster_design)
```

## Which function

Each row names a planning problem and the functions that address it. The
paired `prec_` function evaluates precision under the same sizing model.

| **I want to size, or evaluate** | **Functions** |
|:---|:---|
| a proportion, a mean or a ratio of two totals | `n_prop()`/`prec_prop()`, `n_mean()`/`prec_mean()`, `n_ratio()`/`prec_ratio()` |
| several indicators at once, taking the most demanding | `n_multi()`/`prec_multi()` |
| a two- or three-stage cluster design, for one indicator or a table of them | `n_cluster()`/`prec_cluster()` |
| an allocation across strata or domains, under a budget or a CV target | `n_alloc()`/`prec_alloc()`, `strata_bound()` |
| a two-phase design that screens or follows up | `n_twophase()`/`prec_twophase()` |
| a change between two occasions of a repeated survey | `n_change()`/`prec_change()` |
| the average of several occasions of a repeated survey | `n_pooled()`/`prec_pooled()` |
| a panel that must still deliver a sample after attrition | `n_panel()`/`prec_panel()` |
| the overlap and field schedule a rotation pattern produces | `design_rotation()`, `design_overlap()`, `design_schedule()` |
| the power of a two-group comparison or a difference-in-differences | `power_prop()`, `power_mean()`, `power_did()` |
| the design effect, effective size or degrees of freedom of a plan | `design_effect()`, `effective_n()`, `design_df()`, `varcomp()` |

Sizing for a regression coefficient is not implemented, and there is no
`n_reg()`. The survey-specific part of that calculation, the variance
inflation from a complex design, is the `deff` argument.

## Choose the next function

The package supports several estimands and design structures. Start from the
question your plan must answer:

| Planning question | Continue with |
|:---|:---|
| Which precision target or interval method should I use? | `?n_prop`, `?n_mean`, `?n_ratio`, `?prec_prop` |
| How do several indicators or domains determine one sample? | `?n_multi`, `?prec_multi` |
| How should I construct strata or allocate a sample? | `?strata_bound`, `?n_alloc`, `?prec_alloc` |
| How do I satisfy several allocation requirements together? | `?n_alloc` |
| How many clusters and units per cluster do I need? | `?varcomp`, `?n_cluster`, `?design_effect` |
| How do I plan screening or nonresponse follow-up? | `?n_twophase`, `?prec_twophase` |
| How do overlap, attrition, and rotation affect repeated surveys? | `?n_change`, `?n_pooled`, `?n_panel`, `?design_rotation`, `?design_schedule` |
| Can the design detect a group difference or change? | `?power_prop`, `?power_mean`, `?power_did` |

Sample-size planning depends on anticipated variances and on the sample design
[@cochran1977; @valliant2018]. Treat those inputs as documented assumptions and
revisit the plan when better information becomes available.

## References
