---
title: "Survey sample size planning with svyplan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survey sample size planning with svyplan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(svyplan)
```

`svyplan` provides a toolkit for survey sample size determination. It covers sample sizes for proportions and means, precision analysis, multistage cluster allocation, multi-indicator optimization, strata boundary optimization, and power analysis. The core sizing and precision functions use paired S3 interfaces for round-trip conversions. Sensitivity analysis via `predict()` is available for supported result classes.

## Function ecosystem

The package exports 22 functions organized into six families. The diagram below shows how they relate to each other.

```{r ecosystem-diagram, echo = FALSE, fig.height = 6.5, fig.width = 7}
op <- par(mar = c(0, 0, 0, 0), family = "sans")
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 9.5))

draw_box <- function(x, y, w, h, label, sublabel = NULL,
                     col = "white", border = "black") {
  rect(x - w / 2, y - h / 2, x + w / 2, y + h / 2,
       col = col, border = border, lwd = 1.5)
  text(x, y + if (!is.null(sublabel)) 0.2 else 0,
       label, cex = 0.85, font = 2)
  if (!is.null(sublabel))
    text(x, y - 0.25, sublabel, cex = 0.65, col = "grey30")
}

draw_arrow <- function(x0, y0, x1, y1, both = FALSE, ...) {
  arrows(x0, y0, x1, y1, length = 0.10, lwd = 1.5,
         col = "grey40", code = if (both) 3L else 2L, ...)
}

draw_box(5, 8.5, 3.8, 1.0, "Plan Profile",
         "svyplan")

draw_box(2.5, 6.5, 3.8, 1.4, "Sample Size",
         paste0("n_prop \u00b7 n_mean \u00b7 n_cluster\n",
                "n_multi \u00b7 n_multi_cluster\nn_alloc \u00b7 n_twophase"))
draw_box(7.5, 6.5, 3.8, 1.4, "Precision",
         paste0("prec_prop \u00b7 prec_mean \u00b7 prec_cluster\n",
                "prec_multi \u00b7 prec_multi_cluster\n",
                "prec_alloc \u00b7 prec_twophase"))
draw_box(2.5, 3.5, 3.8, 1.2, "Design Components",
         "varcomp \u00b7 design_effect \u00b7 effective_n")
draw_box(7.5, 3.5, 3.8, 1.2, "Power Analysis",
         "power_prop \u00b7 power_mean \u00b7 power_did")
draw_box(5, 1, 3.8, 1.2, "Stratification",
         "strata_bound")

draw_arrow(3.5, 8.0, 2.5, 7.1)
draw_arrow(6.5, 8.0, 7.5, 7.1)

draw_arrow(4.4, 6.7, 5.6, 6.7, both = TRUE)
text(5, 7.15, "round-trip", cex = 0.6, col = "grey40", font = 3)

draw_arrow(2.5, 4.1, 2.5, 5.9)
draw_arrow(4.5, 3.5, 5.5, 3.5)
draw_arrow(7.5, 4.1, 7.5, 5.9)

text(1.6, 5.0, "feeds into", cex = 0.55, col = "grey40", font = 3, srt = 90)
text(5.0, 3.8, "evaluate", cex = 0.55, col = "grey40", font = 3)
text(8.3, 5.0, "informs", cex = 0.55, col = "grey40", font = 3, srt = 90)

par(op)
```

| Family      | Functions                                                        | Purpose                                |
|:------------|:-----------------------------------------------------------------|:---------------------------------------|
| Plan        | `svyplan`                                                        | Reusable profile for design defaults   |
| Sample size | `n_prop`, `n_mean`, `n_cluster`, `n_multi`, `n_multi_cluster`, `n_alloc`, `n_twophase` | "How many units do I need?"        |
| Precision   | `prec_prop`, `prec_mean`, `prec_cluster`, `prec_multi`, `prec_multi_cluster`, `prec_alloc`, `prec_twophase` | "What precision does *n* achieve?" |
| Power       | `power_prop`, `power_mean`, `power_did`                          | "Can I detect a change?"               |
| Design      | `varcomp`, `design_effect`, `effective_n`                        | Variance components and design effects |
| Strata      | `strata_bound`                                                   | Optimal stratification boundaries      |

Most planning functions return structured S3 objects. Available methods depend on the result class; see the relevant help page for details.

## The n_\*/prec_\* duality

The most distinctive feature of svyplan is the bidirectional link between sample size and precision functions. Every `n_*` function has a `prec_*` counterpart, and each pair supports round-trip conversion. Under the same formula and continuous design assumptions, converting a result to its counterpart and back reproduces the original target. Changing methods or using an operational integer allocation can change the result.

```{r decision-diagram, echo = FALSE, fig.height = 4, fig.width = 7}
op <- par(mar = c(0, 0, 0, 0), family = "sans")
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 6))

rrect <- function(x, y, w, h, label, col = "white", border = "black",
                  cex = 0.75, font = 1) {
  rect(x - w / 2, y - h / 2, x + w / 2, y + h / 2,
       col = col, border = border, lwd = 1.5)
  text(x, y, label, cex = cex, font = font)
}

rrect(5, 5.2, 4, 0.8, "What is your question?",
      col = "white", border = "black", cex = 0.9, font = 2)

rrect(1.5, 3.2, 2.5, 0.7, '"How many units?"')
rrect(5.0, 3.2, 2.8, 0.7, '"What precision at n?"')
rrect(8.5, 3.2, 2.5, 0.7, '"Can I detect a change?"')

rrect(1.5, 1.5, 2.8, 0.9, "n_prop \u00b7 n_mean \u00b7 n_cluster\nn_multi \u00b7 n_multi_cluster\nn_alloc \u00b7 n_twophase",
      col = "white", border = "black", cex = 0.65)
rrect(5.0, 1.5, 2.8, 0.9, "prec_prop \u00b7 prec_mean \u00b7 prec_cluster\nprec_multi \u00b7 prec_multi_cluster\nprec_alloc \u00b7 prec_twophase",
      col = "white", border = "black", cex = 0.65)
rrect(8.5, 1.5, 2.8, 0.8, "power_prop \u00b7 power_mean\npower_did",
      col = "white", border = "black", cex = 0.65)

draw_arrow <- function(x0, y0, x1, y1)
  arrows(x0, y0, x1, y1, length = 0.08, lwd = 1.3, col = "grey40")

draw_arrow(3.5, 4.8, 1.5, 3.6)
draw_arrow(5.0, 4.8, 5.0, 3.6)
draw_arrow(6.5, 4.8, 8.5, 3.6)

draw_arrow(1.5, 2.85, 1.5, 1.95)
draw_arrow(5.0, 2.85, 5.0, 1.95)
draw_arrow(8.5, 2.85, 8.5, 1.95)

par(op)
```

Because `n_*` and `prec_*` are S3 generics, you can pass a result object directly to its inverse. All design parameters travel with the object:

```{r round-trip}
# Compute sample size for a proportion
s <- n_prop(p = 0.3, moe = 0.05, deff = 1.5)
s

# What precision does this n achieve?
p <- prec_prop(s)
p

# Recover the original sample size
n_prop(p)
```

You can also override a parameter on the way back:

```{r round-trip-override}
n_prop(p, cv = 0.08)
```

## Sample sizes for proportions and means

### Proportions

Estimate vaccination coverage (expected ~70%) with a 5 percentage point margin of error:

```{r n-prop-basic}
n_prop(p = 0.70, moe = 0.05)
```

Four method options control how the margin of error is computed:

- **`"wald"`** (default): the textbook normal approximation
- **`"wilson"`**: Wilson score interval, better coverage near 0 or 1
- **`"logodds"`**: log-odds transformation, the right choice when the plan will be reported as an odds ratio
- **`"beta"`**: Korn-Graubard interval, for a proportion whose expected number of positive counts is small. It stays inside $[0, 1]$ by construction, and a `df` argument widens it for a variance estimated from few clusters.

```{r n-prop-methods}
n_prop(p = 0.05, moe = 0.02, method = "wald")
n_prop(p = 0.05, moe = 0.02, method = "wilson")
n_prop(p = 0.05, moe = 0.02, method = "logodds")
n_prop(p = 0.05, moe = 0.02, method = "beta", df = 25)
```

The four differ materially only when the expected count is small. For sizing that difference is rarely worth acting on; for assessing an achieved design it often is. `?n_prop` has the full comparison.

Target a coefficient of variation instead of a margin of error:

```{r n-prop-cv}
n_prop(p = 0.70, cv = 0.05)
```

### Means

Estimate average household expenditure (variance = 40,000, mean = 800) with a margin of error of 30:

```{r n-mean}
n_mean(var = 40000, moe = 30)
```

### Common parameters

Most single-stage `n_*()` and `prec_*()` functions share these parameters:

| Parameter | Meaning |
|:----------|:--------|
| `deff` | Design effect multiplier (default 1) |
| `N` | Finite population size (enables FPC correction) |
| `alpha` | Significance level (default 0.05) |
| `resp_rate` | Expected response rate (inflates *n* by `1/resp_rate`) |

```{r common-params}
n_prop(p = 0.70, moe = 0.05, deff = 1.5, N = 50000, resp_rate = 0.85)
```

The design effect can be less than 1 for efficient designs (e.g. well-stratified or PPS samples). svyplan accepts any `deff > 0`.

`n_cluster()` / `prec_cluster()` use cluster-specific inputs instead (`stage_cost`, `icc`, `unit_relvar`, and optionally `resp_rate`), and `n_alloc()` / `prec_alloc()` have their own allocation-oriented interfaces.

### Survey plan profiles

When the same `deff`, `N`, `resp_rate`, or `alpha` apply across many single-stage calls, bundle them into a `svyplan()` profile to avoid repetition:

```{r svyplan-profile}
plan <- svyplan(deff = 1.5, resp_rate = 0.85, N = 50000)

# Pass as a named argument
n_prop(p = 0.3, moe = 0.05, plan = plan)
n_mean(var = 100, moe = 2, plan = plan)
power_prop(p1 = 0.30, p2 = 0.35, plan = plan)
```

The profile can also hold cluster context (`stage_cost`, `icc`) for `n_cluster()`:

```{r svyplan-cluster}
cl_plan <- svyplan(stage_cost = c(500, 50), icc = 0.05, resp_rate = 0.85)
n_cluster(cv = 0.05, plan = cl_plan)
```

Piping works with both positional and named arguments:

```{r svyplan-pipe}
plan |> n_prop(0.3, moe = 0.05)
plan |> n_prop(p = 0.3, moe = 0.05)
plan |> power_mean(200, effect = 5)
```

Explicit arguments always override plan defaults:

```{r svyplan-override}
# plan has deff = 1.5, but deff = 2.0 wins here
n_prop(p = 0.3, moe = 0.05, plan = plan, deff = 2.0)
```

## Evaluating precision

The `prec_*()` functions answer the reverse question: given a fixed sample size, what precision can you achieve?

```{r prec-basic}
prec_prop(p = 0.3, n = 400)

prec_mean(var = 40000, n = 300, mu = 800)
```

Design parameters work identically:

```{r prec-design}
prec_prop(p = 0.3, n = 400, deff = 1.5, resp_rate = 0.85)
```

### Confidence intervals

`confint()` extracts the expected confidence interval from a single-indicator proportion or mean result, whether it came from an `n_*` or a `prec_*` function. Cluster, allocation, power and multi-indicator results have no single interval to report and are not supported. A mean needs `mu`, since an interval has to be centred somewhere:

```{r confint}
s <- n_prop(p = 0.3, moe = 0.05)
confint(s)

pr <- prec_prop(p = 0.3, n = 400)
confint(pr)

confint(pr, level = 0.99)

# a mean carries no centre of its own, so supply mu
confint(n_mean(var = 100, moe = 2, mu = 50))
```

For proportions, the interval is clamped to [0, 1].

## Multi-indicator surveys

Household surveys like Demographic and Health Surveys (DHS) or Multiple Indicator Cluster Survey (MICS) track many indicators simultaneously. Each indicator has its own expected prevalence, precision target, and design effect. `n_multi()` finds the sample size that satisfies all targets at once.

```{r n-multi}
targets <- data.frame(
  name = c("stunting", "vaccination", "anemia"),
  p    = c(0.25, 0.70, 0.12),
  moe  = c(0.05, 0.05, 0.03),
  deff = c(2.0, 1.5, 2.5)
)

n_multi(targets)
```

The binding indicator (marked with `*`) drives the overall sample size. The other indicators are estimated with better precision than requested.

Proportion rows accept the same four methods as `n_prop()`, either for the whole table through `prop_method` or per row through a `prop_method` column. A `df` column carries the degrees of freedom for `"beta"` rows, so one table can mix an ordinary indicator with a rare one:

```{r n-multi-methods}
targets_rare <- data.frame(
  name = c("stunting", "cocaine_use"),
  p    = c(0.25, 0.02),
  moe  = c(0.05, 0.01),
  prop_method = c("wald", "beta"),
  df   = c(NA, 25)
)

n_multi(targets_rare)
```

### Achieved precision

`prec_multi()` computes the achieved precision for each indicator at the chosen *n*:

```{r prec-multi}
plan <- n_multi(targets)
prec_multi(plan)
```

### Domain targets

Surveys often need adequate precision within subpopulations (urban/rural, regions). Add domain columns to the targets data frame and `n_multi()` optimizes each domain separately:

```{r n-multi-domain}
targets_dom <- data.frame(
  name       = rep(c("stunting", "vaccination", "anemia"), each = 2),
  residence  = rep(c("urban", "rural"), 3),
  p          = c(0.18, 0.32, 0.80, 0.60, 0.08, 0.16),
  moe        = c(0.05, 0.05, 0.05, 0.05, 0.03, 0.03),
  deff       = c(1.5, 2.5, 1.2, 1.8, 2.0, 3.0)
)

n_multi(targets_dom, domains = "residence")
```

Domain columns are specified explicitly via the `domains` parameter. The returned `n` is the per-domain maximum (the binding domain's requirement), not the sum across domains. Use `$domains` to see each domain's sample size.

### MICS/DHS-style relative margin of error

Programmes like UNICEF MICS and DHS express precision as a **relative margin of error** (RME), defined as `MOE / p`. This is related to the coefficient of variation by `RME = z * CV` (at 95% confidence, `RME ≈ 1.96 * CV`). To use svyplan with an RME target, convert to an absolute margin of error: `moe = RME * p`.

```{r n-multi-rme}
# MICS-style: 12% RME per domain, domain-specific deff and prevalence
rme <- 0.12
targets_mics <- data.frame(
  name      = rep(c("stunting", "vaccination"), each = 3),
  region    = rep(c("North", "Central", "South"), 2),
  p         = c(0.35, 0.25, 0.40, 0.60, 0.75, 0.55),
  deff      = c(2.5, 2.0, 3.0, 1.5, 1.2, 1.8),
  resp_rate = c(0.90, 0.85, 0.95, 0.90, 0.85, 0.95)
)
targets_mics$moe <- rme * targets_mics$p

n_multi(targets_mics, domains = "region")
```

The MICS template reports sample size in **households**. svyplan returns the number of **individuals** in the target population. To convert, divide by the expected number of eligible individuals per household: `n_hh = ceiling(n / (pb * hh_size))`, where `pb` is the share of the target population in the total population and `hh_size` is the average household size.

### Minimum sample size per domain

The `min_n_domain` parameter sets a floor for the per-domain sample size (`n_alloc()` uses `min_n_stratum` for its per-stratum floor):

```{r n-multi-min-n}
n_multi(targets_dom, domains = "residence", min_n_domain = 300)
```

### Mixing proportions and means

Targets can mix proportion and mean indicators. Use `var` and `mu` columns for means (leave `p` as `NA`), and `p` for proportions (leave `var`/`mu` as `NA`):

```{r n-multi-mixed}
targets_mixed <- data.frame(
  name = c("vaccination", "expenditure"),
  p    = c(0.70, NA),
  var  = c(NA, 40000),
  mu   = c(NA, 800),
  cv   = c(0.05, 0.10),
  deff = c(1.5, 2.0)
)

n_multi(targets_mixed)
```

## Multistage cluster designs

For two-stage or three-stage designs, the question is not just "how many?" but "how many clusters and how many units per cluster?" The answer depends on the cost structure and the degree of clustering.

### Variance components from frame data

If frame data with cluster identifiers is available, `varcomp()` estimates between-cluster and within-cluster variance components:

```{r varcomp}
set.seed(1)
n_clust <- 80
n_hh <- 15
cluster_id <- rep(seq_len(n_clust), each = n_hh)
cluster_mean <- rnorm(n_clust, mean = 50, sd = 10)
expenditure <- cluster_mean[cluster_id] + rnorm(n_clust * n_hh, sd = 20)
frame <- data.frame(cluster = cluster_id, expenditure = expenditure)

vc <- varcomp(expenditure ~ cluster, data = frame)
vc
as.data.frame(vc)
```

The `icc` value is the survey-planning measure of homogeneity. It feeds directly into `n_cluster()` and the cluster-planning mode of `design_effect()`. This `icc` is not a generic mixed-model ICC: `varcomp()` returns the bounded planning quantity from Valliant, Dever, and Kreuter (2018), and that is the scale expected by the cluster-planning functions. When `icc` is near 0, most variation is within clusters, so the analytical cluster optimum tends to favour many interviews in very few PSUs. When `icc` is near 1, most variation is between clusters, so the optimum tends to favour very few interviews in many PSUs. Both are degenerate boundary cases for the closed-form optimizer, so `n_cluster()` rejects values numerically too close to 0 or 1.

### Optimal two-stage allocation

Given per-stage costs and the variance structure, `n_cluster()` finds the allocation that either minimizes cost for a target CV, or minimizes CV for a given budget:

```{r n-cluster-2stage}
# Minimize cost to achieve CV = 0.05
n_cluster(stage_cost = c(500, 50), icc = vc, cv = 0.05)

# Maximize precision within a budget
n_cluster(stage_cost = c(500, 50), icc = vc, budget = 100000)
```

Passing `icc = vc` (the varcomp object) automatically extracts `icc`, `unit_relvar`, and `var_ratio`.

### Three-stage designs

Add a third cost element and a second icc for three-stage designs:

```{r n-cluster-3stage}
n_cluster(
  stage_cost  = c(1000, 200, 20),
  icc = c(0.01, 0.05),
  cv    = 0.05
)
```

### Response rate in cluster designs

In budget mode, applying a response rate does not change the allocation because the budget constrains the design. Instead, the achieved CV worsens. In CV mode, the first-stage sample is inflated by `1 / resp_rate`:

```{r n-cluster-resp}
n_cluster(stage_cost = c(500, 50), icc = vc, cv = 0.05, resp_rate = 0.90)
```

### Fixed overhead costs

The standard cost model is purely linear: `C = c1*n_psu + c2*n_psu*n_per_psu`. Real surveys also have a fixed overhead (training, infrastructure, setup). The `fixed_cost` parameter adds this term: `C = C0 + c1*n_psu + c2*n_psu*n_per_psu`. In budget mode, only `budget - fixed_cost` is available for the variable component, reducing n_psu while leaving the optimal n_per_psu/n_psu ratio unchanged. In CV mode, all sample sizes stay the same and `fixed_cost` is simply added to the total cost:

```{r n-cluster-fixed}
# Budget mode: fixed overhead reduces the allocatable budget
n_cluster(stage_cost = c(500, 50), icc = vc, budget = 100000, fixed_cost = 5000)

# CV mode: same allocation, cost increases by fixed_cost
n_cluster(stage_cost = c(500, 50), icc = vc, cv = 0.05, fixed_cost = 5000)
```

The same parameter is available in `n_multi_cluster()` for multistage
multi-indicator designs.

### Evaluating cluster allocations

`prec_cluster()` computes the achieved CV for any allocation:

```{r prec-cluster}
alloc <- n_cluster(stage_cost = c(500, 50), icc = vc, budget = 100000)
prec_cluster(alloc)

# Arbitrary allocation
prec_cluster(n = c(50, 12), icc = 0.05)
```

### Design effects

`design_effect()` builds the design effect you expect a plan to produce, from the design features you are choosing. The clustering component uses the same variance model as `n_cluster()` and `prec_cluster()`, so the two always agree:

```{r design-effect}
design_effect(icc = 0.05, n_per_psu = 20)
```

Components are selected by the arguments you supply and multiply together. Unequal weighting comes from planned weights or from a planned allocation's stratum sizes and takes, and the stratification gain (the only component that can fall below 1) from the stratum standard deviations and means:

```{r design-effect-components}
frame <- data.frame(
  N = c(50000, 120000), n = c(600, 400), sd = c(12, 20), mean = c(55, 48)
)
deff <- design_effect(icc = 0.05, n_per_psu = 20, strata = frame)
deff
```

The result is a plain number wherever one is expected, and `effective_n()` is its mirror:

```{r design-effect-use}
n_prop(p = 0.3, moe = 0.05, deff = deff)
effective_n(deff, n = 1000)
```

You can also read the features off a plan you already built, including an `n_alloc()` allocation:

```{r design-effect-plan}
design_effect(n_cluster(stage_cost = c(500, 50), icc = 0.05, cv = 0.05))
```

These are planning quantities. After data collection, compute the design effect the sample actually achieved with `survey::svymean(..., deff = TRUE)`, which uses the realized weights, strata, and clusters.

### Multi-indicator multistage

For surveys with multiple indicators in a cluster design, provide
`icc_psu` columns and stage costs to `n_multi_cluster()`:

```{r n-multi-cluster}
targets_ms <- data.frame(
  name   = c("stunting", "vaccination"),
  p      = c(0.25, 0.70),
  cv     = c(0.10, 0.08),
  icc_psu = c(0.05, 0.02)
)

n_multi_cluster(targets_ms, stage_cost = c(500, 50), budget = 80000)
```

## Strata boundaries

When a continuous auxiliary variable is available on the frame,
`strata_bound()` constructs cut points intended to reduce the sample size or
CV for a given number of strata. The iterative choices below are local
heuristics, so their result is a candidate solution rather than a certified
global minimum.

```{r strata-bound}
set.seed(12345)
x <- rlnorm(5000, meanlog = 6, sdlog = 1.2)

strata_bound(x, n_strata = 4, n = 300, method = "cumrootf")
```

Four methods are available:

| Method       | Algorithm                                 | Speed    | Best for       |
|:-------------|:------------------------------------------|:---------|:---------------|
| `"cumrootf"` | Dalenius-Hodges cumulative root frequency | Fast     | General use    |
| `"geo"`      | Geometric progression                     | Fast     | Skewed data    |
| `"lh"`       | LH-inspired coordinate optimization       | Moderate | Fast local search |
| `"kozak"`    | Kozak-inspired random-restart local search| Slow     | More starting points |

The `"kozak"` heuristic explores multiple starts and can find a better local
solution on highly skewed distributions. Neither iterative method guarantees a
global optimum or claims to reproduce the published algorithm exactly:

```{r strata-kozak}
strata_bound(x, n_strata = 4, cv = 0.05, method = "kozak")
```

### Allocation methods

Allocation within strata is controlled by the `alloc` parameter:

- `"proportional"`: $n_h \propto N_h$
- `"neyman"` (default): $n_h \propto N_h \cdot S_h$ (minimizes national CV)
- `"optimal"`: $n_h \propto N_h \cdot S_h / \sqrt{c_h}$ (accounts for costs)
- `"power"`: Bankier (1988) compromise, $n_h \propto S_h \cdot N_h^{q}$,
  where the exponent `alloc_q` $\in [0, 1]$ controls the trade-off between
  national precision (`alloc_q` $= 1$, Neyman) and equal subnational CVs
  (`alloc_q` $= 0$)

```{r strata-power-alloc}
# Bankier power allocation (compromise, alloc_q = 0.5)
strata_bound(x, n_strata = 4, n = 300, alloc = "power", alloc_q = 0.5)
```

### Classifying new observations

`predict()` applies the learned boundaries to new data, returning a
factor:

```{r strata-predict}
sb <- strata_bound(x, n_strata = 4, n = 200, method = "cumrootf")
new_x <- c(100, 500, 1500, 5000)
predict(sb, newdata = new_x, labels = paste0("S", 1:4))
```

### Visualizing allocation

`plot()` shows the sampling fraction per stratum. Under Neyman allocation, high-variance strata are sampled more intensively. The dashed line marks the overall fraction:

```{r strata-plot, fig.width=7, fig.height=4}
plot(sb)
```

### Handing boundaries to the allocator

`$strata` carries the `N`, `sd`, and `mean` columns `n_alloc()` needs, so the
two halves of a stratified design connect directly:

```{r strata-handoff}
sb_cv <- strata_bound(x, n_strata = 4, cv = 0.05, method = "lh")
n_alloc(sb_cv$strata, cv = 0.05)
```

Both functions evaluate the same variance, so `cv` means the same thing to
each and `n` is a fielded sample in both. Design assumptions carry across
unchanged, whether passed directly or through a `svyplan()` profile:

```{r strata-deff}
plan <- svyplan(deff = 1.8, resp_rate = 0.85)
sb_deff <- strata_bound(x, n_strata = 4, cv = 0.05, plan = plan)
sb_deff$n
n_alloc(sb_deff$strata, cv = 0.05, plan = plan)$n
```

The remaining gap between the two totals is integer rounding, not a
difference in model: `strata_bound()` reports the whole-unit allocation in
`$strata$n`, while `n_alloc()` reports the continuous optimum.

A scalar `deff` does not move the cut points. It scales the variance of every
candidate boundary set equally, so what it changes is the `n` a target
requires, not where the strata divide.

## Stratified allocation

Given a sampling frame with stratum population sizes (`N`) and stratum standard deviations (`sd`), `n_alloc()` distributes a total sample across strata. This is the second half of a stratified design: `strata_bound()` finds the cut points, and `n_alloc()` allocates within strata.

```{r n-alloc-basic}
frame <- data.frame(
  N = c(4000, 3000, 3000),
  sd = c(10, 15, 8),
  mean = c(50, 60, 55)
)

# Fixed total n with Neyman allocation
n_alloc(frame, n = 600, alloc = "neyman")
```

Three solve modes are available: fixed total `n`, target `cv`, or `budget` constraint (when `unit_cost` is provided in the frame).

Constraints and domain-level targets are often the more practical use cases:

```{r n-alloc-constraints}
frame_constraints <- transform(
  frame,
  unit_cost = c(1, 1.5, 1),
  max_weight = c(25, 20, NA),
  take_all = c(FALSE, FALSE, TRUE)
)

n_alloc(frame_constraints, budget = 3500, alloc = "optimal", min_n_stratum = 40)
```

```{r n-alloc-domains}
frame_domains <- data.frame(
  province = c("North", "North", "South", "South"),
  stratum = c("Urban", "Rural", "Urban", "Rural"),
  N = c(2000, 3000, 1800, 3200),
  sd = c(12, 18, 10, 16),
  mean = c(55, 48, 58, 50)
)

# Minimum total n such that each province meets the CV target
n_alloc(frame_domains, domains = "province",
        cv = 0.04, alloc = "power", alloc_q = 0.3)
```

`prec_alloc()` computes the achieved precision for any allocation:

```{r prec-alloc}
prec_alloc(frame, n = c(200, 250, 150))
```

### Joint indicator and domain targets

When one allocation must satisfy several estimands, use an atomic frame plus
long `measures` and `targets` tables. Atomic means that every frame row has a
single value for every domain classification in use. The example below crosses
region and residence, allowing those two domain systems to overlap without
requiring separate public functions or an `alloc` selector.

```{r n-alloc-joint-data}
joint_frame <- data.frame(
  stratum = c("NU", "NR", "SU", "SR"),
  region = c("North", "North", "South", "South"),
  residence = c("Urban", "Rural", "Urban", "Rural"),
  N = c(1000, 2000, 1500, 1000),
  unit_cost = c(1, 1.2, 1.5, 1)
)

joint_measures <- data.frame(
  stratum = rep(joint_frame$stratum, 2),
  name = rep(c("vaccination", "income"), each = 4),
  p = c(0.50, 0.40, 0.60, 0.30, rep(NA, 4)),
  mean = c(rep(NA, 4), 50, 55, 60, 45),
  sd = c(rep(NA, 4), 10, 12, 15, 9)
)

joint_targets <- data.frame(
  name = c("vaccination", "vaccination", "income"),
  domain = c(".overall", "region", "residence"),
  level = c(NA, "North", "Urban"),
  cv = c(0.05, 0.08, NA),
  moe = c(NA, NA, 2)
)
```

```{r n-alloc-joint-fit}
joint_fit <- n_alloc(
  joint_frame,
  measures = joint_measures,
  targets = joint_targets,
  min_n_stratum = 2
)
joint_fit
joint_fit$constraints
```

The optimizer minimizes variable cost subject to all three requirements.
`joint_fit$detail$n` is the continuous optimum and
`joint_fit$detail$n_int` is the deterministic feasible operational
recommendation. The latter is locally cleaned but is not claimed to be a
globally optimal integer solution. Both precision tables are retained:

```{r n-alloc-joint-assess}
prec_alloc(joint_fit)
prec_alloc(joint_fit, n = joint_fit$detail$n_int)
```

A modified named allocation can be assessed in any order because names are
matched exactly to `frame$stratum`.

#### Fixed-take multistage joint allocation

The same public interface supports two- and three-stage designs when every
later-stage take is fixed. The frame owns stage populations, takes, and costs;
the measures table owns indicator-specific stage homogeneity parameters. Only
`n_psu` is optimized. Public `n`, planning bounds, and precision assessments
remain in ultimate-unit units.

This three-stage example samples PSUs, then a fixed number of SSUs per PSU,
then a fixed number of ultimate units per SSU:

```{r n-alloc-joint-three-stage}
joint_frame_3stage <- within(joint_frame, {
  unit_cost <- NULL
  N_psu <- c(100, 160, 120, 90)
  N_ssu <- c(600, 1200, 900, 500)
  n_per_psu <- c(8, 10, 12, 7)
  n_per_ssu <- c(4, 5, 3, 4)
  cost_psu <- c(300, 400, 450, 350)
  cost_ssu <- c(25, 30, 35, 28)
  cost_tsu <- c(5, 6, 7, 5)
})

joint_measures_3stage <- transform(
  joint_measures,
  icc_psu = rep(c(0.03, 0.05, 0.08, 0.04), 2),
  icc_ssu = rep(c(0.10, 0.08, 0.12, 0.06), 2),
  var_ratio_psu = 1,
  var_ratio_ssu = 1
)

joint_fit_3stage <- n_alloc(
  joint_frame_3stage,
  measures = joint_measures_3stage,
  targets = joint_targets,
  min_n_stratum = 20
)
joint_fit_3stage$detail[, c(
  "stratum", "n_psu_int", "n_per_psu", "n_per_ssu", "n_int"
)]
prec_alloc(joint_fit_3stage, n = joint_fit_3stage$detail$n_int)
```

The exact operational identity is
`n_int = n_psu_int * n_per_psu * n_per_ssu`. `N_psu` is required for both
multistage forms, while three-stage frames also require the stratum's total
`N_ssu`. Fixed takes must be whole numbers. `take_all` is unavailable because
taking every PSU alone does not establish a census of ultimate units.

### Stratified two-stage designs

Most household surveys stratify first and cluster within each stratum,
sampling enumeration areas and then households. Adding a `icc_psu`
column to the frame switches `n_alloc()` to this design. The per-stratum
homogeneity can come straight from `varcomp()` with `strata`, whose
output columns are named to match the frame:

```{r n-alloc-cluster-varcomp}
set.seed(3)
listing <- data.frame(
  region = rep(c("North", "South"), each = 600),
  ea = rep(1:60, each = 20),
  income = rnorm(1200, rep(c(50, 70), each = 600), 15) +
    rep(rnorm(60, 0, 6), each = 20)
)
vc_region <- varcomp(income ~ ea, data = listing, strata = ~region)
vc_region
```

```{r n-alloc-cluster}
frame_2stage <- merge(
  data.frame(stratum = c("North", "South"), N = c(40000, 60000)),
  as.data.frame(vc_region)[, c("stratum", "sd", "mean", "icc_psu", "var_ratio_psu")],
  by = "stratum"
)
frame_2stage$cost_psu <- c(400, 550)
frame_2stage$cost_ssu <- c(45, 60)

res_2stage <- n_alloc(frame_2stage, cv = 0.02)
res_2stage
res_2stage$detail[, c("stratum", "n_int", "n_per_psu", "n_psu_int")]
```

Each stratum gets its cost-optimal cluster take (fix it instead with a
`n_per_psu` column), and the PSU counts fall out of the element
allocation. Budget mode and every constraint from the previous section
work the same way here.

## Two-phase designs

In a two-phase (or double) sample, a large cheap phase 1 measures something that helps target the expensive measurement, and a subsample is carried into phase 2. `n_twophase()` allocates both phases at once, minimizing cost for a target CV or CV for a fixed budget.

The frame is one row per phase-2 stratum, similarly to what is used in `n_alloc()` :

```{r twophase-strat}
frame <- data.frame(
  stratum   = c("A", "B", "C", "D"),
  N         = c(3500, 2500, 2500, 1500),
  sd        = c(12, 25, 8, 40),
  mean      = c(40, 70, 35, 90),
  unit_cost = c(2, 5, 1, 9)
)

n_twophase(frame, phase1_cost = 1, budget = 50000)
```

The `nu` column is the fraction of each phase-1 stratum carried forward. Its shape is Neyman-like, proportional to $S_h/\sqrt{c_h}$, but the overall scale is set by the *between*-stratum variance: the weaker the stratification, the closer every fraction moves to 1, meaning keep everything phase 1 found.

### Two design effects, not one

Neither phase has to be a simple random sample, and the two carry *separate* design effects because they apply to different things. `phase1_deff` inflates the between-stratum component, the part phase 1 is responsible for. A `deff` column inflates each stratum's within-stratum residual, the part phase 2 has to measure. `single_deff` describes the comparator, which need not be fielded like phase 2.

```{r twophase-deff}
clustered <- transform(frame, deff = c(1.2, 2.5, 0.8, 1.7))

n_twophase(clustered, phase1_cost = 1, budget = 50000,
           phase1_deff = 2.2, single_deff = 2.9)
```

Inflating the combined variance by one design effect instead would be a different and wrong model. The two also pull against each other, and the trap runs in the direction that looks attractive: a stratifier built purely from between-cluster structure shrinks the residual, lowering the stratum `deff`, but it makes the fitted values nearly constant within clusters and so drives `phase1_deff` up towards the cluster size. Under a clustered phase 1 that first term can dominate, and the design loses to a plain sample even though the residual design effect looks favourable.

Every result carries the single-phase comparison, because two-phase sampling is not always an improvement. Skipping phase 1 and measuring directly can be both cheaper and more precise when the screener costs too much or the stratifier predicts too little, and the print output says so when that happens.

### Response rates

Both phases can lose sample to nonresponse, and the two enter separately. `resp_rate` is the phase-1 response or successful-classification rate; a `resp_rate` column is the phase-2 completion rate in each stratum. Response divides its own component, exactly as the design effects do, so a stratum that responds poorly is subsampled differently rather than the whole design being inflated by one factor.

```{r twophase-resp}
screened <- transform(frame, resp_rate = c(0.9, 0.7, 0.85, 0.6))

n_twophase(screened, phase1_cost = 1, budget = 50000, resp_rate = 0.8)
```

Everything here counts units **issued**, so a cost quoted per completed interview has to be converted first: `c_contact + resp_rate * c_complete`. Issued and expected-responding counts are reported separately, because they are different quantities and neither is an effective sample size. There is no single effective size for a two-phase design, since the two variance components carry different design effects.

Phase 2 can only draw from what phase 1 managed to classify, so every subsampling fraction is capped at the phase-1 response rate.

One caveat worth stating plainly: dividing by a response rate assumes response is ignorable *within the strata you supplied*. That is a substantive claim about the strata, not a property of the arithmetic, and no sample size removes nonresponse bias. Where the assumption is uncomfortable, modelling the nonresponse as a follow-up phase is the alternative the design itself offers.

### When the phase-1 sample is already fixed

Often it is. The screener has run, or phase 1 is an existing survey, a panel, or a listing operation sized by field capacity rather than by this design. Phone surveys built on a previous face-to-face round are the common case. Supply `n_phase1` and the only decision left is how deep to subsample.

```{r twophase-fixed}
n_twophase(frame, phase1_cost = 1, budget = 50000, n_phase1 = 20000)
```

The relative allocation is unchanged, because it is $S_h\sqrt{d_{2h}/c_h}$ in every mode. What moves is the overall scale, now pinned by the budget left after paying for phase 1 rather than by the variance-cost trade-off. Since the optimizing choice of `n_phase1` is the best member of this family, fixing it can only match or lose to leaving it free, and the difference is what a phase-1 size you did not choose is costing you.

Two things can go wrong, and both give a specific error. The budget may not reach phase 2 at all. Or a target CV may sit below the **phase-1 variance floor**: what that `n_phase1` leaves when every classified unit is carried into phase 2, which is

$$\frac{1}{\mu}\sqrt{\frac{1}{n_a}\left(\frac{d_1A}{r_1} + \sum_h \frac{d_{2h}W_hS_h^2}{r_{2h}r_1}\right) - \frac{d_1A + \sum_h d_{2h}W_hS_h^2}{N}}.$$

The between-stratum term is the part already paid for by the phase-1 size. It is not the whole floor, though, because subsampling stops at the classification rate rather than at 1, so the phase-2 residual survives at $\nu_h = r_1$ and is added to it. The subtracted term is the finite population correction, and it is what makes the floor vanish at a census: with $n_a = N$ and no losses the two halves cancel exactly. It carries the design effects, since those inflate an SRSWOR variance here as everywhere, but not the response rates, because measuring every unit removes all the variance while failing to reach some of them does not.

### Whole units and assurance

`$operational` carries the field design in whole units. Budget mode floors and then buys back units in order of variance reduction per unit cost, staying inside the budget; CV mode rounds up, overshooting precision rather than missing it.

Planning at the *expected* respondent count leaves roughly half of all designs short. `assurance` reports the issue for which the required respondents arrive with at least that probability, from the binomial distribution of respondents:

```{r twophase-assurance}
screened <- transform(frame, resp_rate = c(0.9, 0.7, 0.85, 0.6))

n_twophase(screened, phase1_cost = 1, budget = 50000,
           resp_rate = 0.8, assurance = 0.90)
```

If the assured phase-2 issue exceeds the pool phase 1 supplies, that is a signal to enlarge phase 1 rather than to over-issue.

### Nonresponse follow-up

Following up a subsample of nonrespondents is the same allocation problem with two strata. The respondents are already measured, so they cost nothing more and are all kept, which is `unit_cost = 0` with `take_all = TRUE`. Only the nonrespondents are subsampled:

```{r twophase-nrfu}
theta <- 0.5   # phase-1 response rate

nrfu <- data.frame(
  stratum   = c("respondents", "nonrespondents"),
  N         = c(theta, 1 - theta),
  sd        = c(1, 1),
  unit_cost = c(0, 200),
  take_all  = c(TRUE, FALSE)
)

n_twophase(nrfu, phase1_cost = 50, budget = 100000,
           mu = 1, single_cost = 50 / theta)
```

The optimal follow-up fraction reduces to $\sqrt{c_1/(c_2\theta)}$ here, the standard result. Two details matter in this configuration. `single_cost`: a `unit_cost` of 0 means "already measured", not "free", so the baseline for the comparison has to be named explicitly as the cost of one completed interview without follow-up. And `resp_rate` should stay at its default of 1, because here the strata *are* response status, so classification succeeds for every unit and setting a phase-1 rate as well would count the same loss twice.

## Power analysis

Precision analysis answers "how precisely can we estimate a level?" Power analysis answers a different question: "can we detect a change or difference between two groups or time points?"

### Sample size for a detectable change

A survey measured vaccination coverage at 70%. How many households are
needed in a follow-up to detect a 5 percentage point increase with 80%
power?

```{r power-n}
power_prop(p1 = 0.70, p2 = 0.75, power = 0.80)
```

With a design effect:

```{r power-n-deff}
power_prop(p1 = 0.70, p2 = 0.75, power = 0.80, deff = 2.0)
```

### Three solve modes

`power_prop()` and `power_mean()` each solve for whichever of `n`, `power`, or the effect size is left unspecified:

```{r power-modes}
# Solve for n (default when p2 and power given)
power_prop(p1 = 0.70, p2 = 0.75, power = 0.80, deff = 2.0)

# Solve for power (set power = NULL)
power_prop(p1 = 0.70, p2 = 0.75, n = 1500, power = NULL, deff = 2.0)

# Solve for MDE (omit p2)
power_prop(p1 = 0.70, n = 1500, deff = 2.0)
```

### Panel surveys

If the follow-up resamples a fraction of the original households and the between-round correlation is known, the required sample size drops:

```{r power-panel}
power_prop(p1 = 0.70, p2 = 0.75, power = 0.80, deff = 2.0,
           overlap = 0.50, overlap_cor = 0.6)
```

### Continuous outcomes

The same three solve modes work for means:

```{r power-mean}
# Sample size to detect a difference of 5 (within-group var = 200)
power_mean(200, effect = 5)

# MDE with n = 400 per group
power_mean(200, n = 400)
```

### Difference-in-differences

`power_did()` handles two-group, two-period DiD designs. Specify treated and control group outcomes as `c(baseline, endline)` vectors:

```{r power-did}
# Proportion outcome: treated improves, control stable
power_did(treat = c(0.50, 0.55), control = c(0.50, 0.48),
          outcome = "prop", effect = 0.07)

# Mean outcome with panel overlap
power_did(treat = c(50, 55), control = c(50, 52),
          outcome = "mean", var = 100, effect = 3,
          overlap = 0.5, overlap_cor = 0.6)
```

### Unequal groups and alternative methods

Both `power_prop()` and `power_mean()` support unequal group sizes via `ratio` or explicit `n = c(n1, n2)`:

```{r power-unequal}
# 2:1 allocation ratio
power_prop(p1 = 0.30, p2 = 0.35, ratio = 2)

# Unequal group variances
power_mean(c(80, 120), effect = 5)
```

For rare proportions, arcsine and log-odds transforms are more accurate:

```{r power-methods}
power_prop(p1 = 0.15, p2 = 0.18, alternative = "one.sided",
           method = "arcsine")

power_prop(p1 = 0.15, p2 = 0.18, alternative = "one.sided",
           method = "logodds")
```

### Power curve

`plot()` draws the power-vs-sample-size curve with reference lines at the solved point:

```{r power-plot, fig.width=7, fig.height=4}
pw <- power_prop(p1 = 0.70, p2 = 0.75, power = 0.80, deff = 2.0)
plot(pw)
```

## Sensitivity analysis

How sensitive is a sample size to assumptions about the design effect, response rate, or prevalence? The `predict()` method evaluates any `svyplan` result at new parameter combinations.

### Varying design parameters

```{r predict-n}
x <- n_prop(p = 0.3, moe = 0.05, deff = 1.5)
predict(x, expand.grid(
  deff = c(1.0, 1.5, 2.0, 2.5),
  resp_rate = c(0.8, 0.9, 1.0)
))
```

### Cluster budget scenarios

```{r predict-cluster}
cl <- n_cluster(stage_cost = c(500, 50), icc = 0.05, budget = 100000)
predict(cl, data.frame(budget = c(50000, 100000, 150000, 200000)))
```

### Power curves

```{r predict-power}
pw <- power_prop(p1 = 0.30, p2 = 0.35, n = 500, power = NULL)
predict(pw, data.frame(n = seq(200, 1000, 200)))
```

### Precision by sample size

```{r predict-prec}
pr <- prec_prop(p = 0.3, n = 400)
predict(pr, data.frame(n = c(100, 200, 400, 800, 1600)))
```

## Putting it all together

A typical planning sequence for a national household survey ties several `svyplan` functions together. A `svyplan()` profile captures the shared design assumptions:

1. Set shared design parameters once with `svyplan()`
2. Define indicators and precision targets per domain
3. Compute sample sizes with `n_multi()`
4. Evaluate achieved precision with `prec_multi()`
5. Run sensitivity analysis with `predict()`
6. If frame data is available, estimate variance components with `varcomp()`
7. Optimize the multistage allocation with `n_cluster()` or `n_multi_cluster()`
8. Determine strata boundaries with `strata_bound()`
9. Check that the design has adequate power with `power_prop()` / `power_mean()`

```{r workflow}
# 1: shared design assumptions
design <- svyplan(deff = 2.0, resp_rate = 0.85)

# 2-3: multi-indicator sample size
targets <- data.frame(
  name = c("stunting", "vaccination", "anemia"),
  p    = c(0.25, 0.70, 0.12),
  moe  = c(0.05, 0.05, 0.03),
  deff = c(2.0, 1.5, 2.5)
)
result <- n_multi(targets)
result

# 4: achieved precision for each indicator
prec_multi(result)

# 5: how sensitive is the binding indicator to the response rate?
predict(
  n_prop(p = 0.25, moe = 0.05, plan = design),
  data.frame(resp_rate = c(0.7, 0.8, 0.9, 1.0))
)

# 9: can we detect a 5pp decline in stunting with this sample?
power_prop(
  p1 = 0.25, p2 = 0.20,
  n = as.integer(result), power = NULL,
  plan = design
)
```

Sample-size, cluster, power, and strata-boundary results support `as.integer()` and `as.double()`. For sample-size results, these return the rounded-up and exact values of *n*, respectively, which makes it straightforward to pass results between functions or into other packages.

## Scope

`svyplan` answers one question in several forms: how many units does a planned design need, and what precision does a given design achieve. Some neighbouring topics are deliberately outside that boundary. They are listed here so their absence reads as a decision rather than an oversight.

**Totals.** There is no `n_total()`. A total is a mean rescaled by *N*, so the CV target is identical and a margin of error on the total is `moe / N` on the mean. `?n_mean` has a worked section on sizing for a total.

**Ratio and regression estimation.** Ratio estimators and regression estimators are analysis-stage techniques. Designs are not usually sized around the precision of a ratio, and sizing for a regression coefficient is a general statistical problem rather than a survey one, covered by packages such as `pwr`. The survey-specific part, the variance inflation from a complex design, is already the `deff` argument.

**Two-phase designs beyond stratification and follow-up.** `n_twophase()` allocates a two-phase sample: a large cheap phase 1, then a subsample measured on the expensive variable. Double sampling for stratification and nonresponse follow-up are the same problem in that function, differing only in which strata are marked `take_all`. What is *not* covered is the third member of the family, double sampling for a regression or ratio estimator, where the phase-1 variable is continuous rather than a stratifier. Its variance effect is expressible today as a design effect below 1, since a regression estimator multiplies the variance by $1 - \rho^2$, but the cost trade-off between the phases is not built in.

**More than three stages.** Two- and three-stage designs are supported. Surveys rarely go deeper, and the cost and variance models for a fourth stage would carry more assumptions than they would earn.

**Targeting nonresponse follow-up.** Deciding *how much* follow-up to do is answered by `n_twophase()`. Deciding *which* cases to pursue is not, and is out of scope: prioritizing by response propensity, or adapting effort as fieldwork proceeds, needs paradata from the field that no planning stage has.

**PPS selection inside joint allocation.** `varcomp()` estimates variance components under PPS first-stage selection, but the joint allocator (`n_alloc()` with `measures` and `targets`) optimizes a variance written for equal-probability PSU selection. Feeding it components estimated under PPS is therefore an approximation, and the more unequal the sizes, the rougher it is.

Certainty units are better supported than that might suggest. `strata_bound(take_all_above = )` cuts a census stratum at a size threshold, and a `take_all` frame column makes `n_alloc()` enumerate that stratum, in joint mode as much as in any other. What is genuinely unavailable is `take_all` in a **fixed-take multistage** joint design, where it is refused rather than approximated: taking every PSU still leaves a sample of units within each one, so it does not establish the ultimate-unit census that `take_all` asserts. A certainty *PSU* stratum in a multistage design is thus the case you construct yourself, by splitting the frame on the measure of size and planning that stratum separately.

**Domains not identifiable on the frame.** Targets can be set per domain when domain membership is known before fielding, as it is for region or urban/rural. For domains realized only after contact, such as age or sex subgroups, the realized subsample is random and has to be inflated by hand from the expected domain share.

**Converting between listing units and analysis units.** Sizing works in the units the precision target refers to. Converting a size in eligible persons to a number of households to list, given a mean household size and an eligibility rate, is arithmetic the user does explicitly. The MICS/DHS section above and the README both show the calculation.
