---
title: "Getting started with svyplan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with svyplan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(svyplan)
```

`svyplan` provides a toolkit for survey sample size determination. It covers sample sizes for proportions and means, precision analysis, multistage cluster allocation, multi-indicator optimization, strata boundary optimization, and power analysis. The core sizing and precision functions use paired `S3` interfaces for round-trip conversions. Sensitivity analysis via `predict()` is available for supported result classes.

## Function ecosystem

The package exports `r length(getNamespaceExports("svyplan"))` functions organized into six families. The diagram below shows how they relate to each other.

```{r ecosystem-diagram, echo = FALSE, fig.height = 6.5, fig.width = 7}
op <- par(mar = c(0, 0, 0, 0), family = "sans")
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 9.5))

draw_box <- function(x, y, w, h, label, sublabel = NULL,
                     col = "white", border = "black") {
  rect(x - w / 2, y - h / 2, x + w / 2, y + h / 2,
       col = col, border = border, lwd = 1.5)
  text(x, y + if (!is.null(sublabel)) h / 2 - 0.22 else 0,
       label, cex = 0.85, font = 2)
  if (!is.null(sublabel))
    text(x, y - 0.12, sublabel, cex = 0.65, col = "grey30")
}

draw_arrow <- function(x0, y0, x1, y1, both = FALSE, ...) {
  arrows(x0, y0, x1, y1, length = 0.10, lwd = 1.5,
         col = "grey40", code = if (both) 3L else 2L, ...)
}

draw_box(5, 8.5, 3.8, 1.0, "Plan Profile",
         "svyplan")

draw_box(2.5, 6.5, 3.8, 2.1, "Sample Size",
         paste0("n_prop \u00b7 n_mean \u00b7 n_ratio\n",
                "n_cluster \u00b7 n_multi \u00b7 n_multi_cluster\n",
                "n_alloc \u00b7 n_twophase\n",
                "n_change \u00b7 n_panel \u00b7 n_pooled"))
draw_box(7.5, 6.5, 3.8, 2.1, "Precision",
         paste0("prec_prop \u00b7 prec_mean \u00b7 prec_ratio\n",
                "prec_cluster \u00b7 prec_multi \u00b7 prec_multi_cluster\n",
                "prec_alloc \u00b7 prec_twophase\n",
                "prec_change \u00b7 prec_panel\nprec_pooled"))
draw_box(2.5, 3.5, 3.8, 1.7, "Design Components",
         paste0("varcomp \u00b7 design_effect\n",
                "effective_n \u00b7 design_df\n",
                "design_overlap \u00b7 design_schedule"))
draw_box(7.5, 3.5, 3.8, 1.4, "Power Analysis",
         "power_prop \u00b7 power_mean \u00b7 power_did")
draw_box(5, 1, 3.8, 1.2, "Stratification",
         "strata_bound")

draw_arrow(3.5, 8.0, 2.5, 7.5)
draw_arrow(6.5, 8.0, 7.5, 7.5)

draw_arrow(4.4, 6.9, 5.6, 6.9, both = TRUE)
text(5, 7.3, "round-trip", cex = 0.6, col = "grey40", font = 3)

draw_arrow(2.5, 4.25, 2.5, 5.5)
draw_arrow(4.5, 3.4, 5.5, 3.4)
draw_arrow(7.5, 4.25, 7.5, 5.5)

text(1.6, 4.9, "feeds into", cex = 0.55, col = "grey40", font = 3, srt = 90)
text(5.0, 3.7, "evaluate", cex = 0.55, col = "grey40", font = 3)
text(8.3, 4.8, "informs", cex = 0.55, col = "grey40", font = 3, srt = 90)

par(op)
```

| Family      | Functions                                                        | Purpose                                |
|:------------|:-----------------------------------------------------------------|:---------------------------------------|
| Plan        | `svyplan`                                                        | Reusable profile for design defaults   |
| Sample size | `n_prop`, `n_mean`, `n_ratio`, `n_cluster`, `n_multi`, `n_multi_cluster`, `n_alloc`, `n_twophase`, `n_change`, `n_panel`, `n_pooled` | "How many units do I need?"        |
| Precision   | `prec_prop`, `prec_mean`, `prec_ratio`, `prec_cluster`, `prec_multi`, `prec_multi_cluster`, `prec_alloc`, `prec_twophase`, `prec_change`, `prec_panel`, `prec_pooled` | "What precision does *n* achieve?" |
| Power       | `power_prop`, `power_mean`, `power_did`                          | "Can I detect a difference?"            |
| Design      | `varcomp`, `design_effect`, `effective_n`, `design_df`, `design_overlap`, `design_schedule` | Variance components, design effects, degrees of freedom, and rotation schedules |
| Strata      | `strata_bound`                                                   | Optimal stratification boundaries      |

The same table read by question rather than by family, since several questions cut across the rows:

| Question | Where |
|:---------|:------|
| How many units for a target precision on a level? | `n_prop`, `n_mean`, and the cluster, allocation and two-phase members |
| How many units for a ratio of two totals from the same sample? | `n_ratio` |
| How many for a target precision on a **change** between two occasions? | `n_change` |
| How many to **recruit** so that enough survive to a later wave? | `n_panel` |
| How many for a target precision on an **average across occasions**, an annual figure from quarterly rounds? | `n_pooled` |
| What precision does a design I already have achieve? | the `prec_*` counterpart of each |
| Can I detect a difference, and how large a one? | `power_prop`, `power_mean`, `power_did` |
| Can I detect a change over two occasions? | `power_prop` and `power_mean` with `overlap` set, which is the same variance `n_change` inverts |
| What overlap does my rotation schedule produce? | `design_overlap` |
| What should each startup or intake cohort issue, and when is it active? | `design_schedule` |
| What design effect, effective size or degrees of freedom will the design have? | `design_effect`, `effective_n`, `design_df`, `varcomp` |

Most planning functions return structured S3 objects. Available methods depend on the result class. See the relevant help page for details.

## The n_\*/prec_\* duality

The most distinctive feature of svyplan is the bidirectional link between sample size and precision functions. Every `n_*` function has a `prec_*` counterpart, and each pair supports round-trip conversion. Under the same formula and continuous design assumptions, converting a result to its counterpart and back reproduces the original target. Changing methods or using an operational integer allocation can change the result.

```{r decision-diagram, echo = FALSE, fig.height = 4, fig.width = 7}
op <- par(mar = c(0, 0, 0, 0), family = "sans")
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 6))

rrect <- function(x, y, w, h, label, col = "white", border = "black",
                  cex = 0.75, font = 1) {
  rect(x - w / 2, y - h / 2, x + w / 2, y + h / 2,
       col = col, border = border, lwd = 1.5)
  text(x, y, label, cex = cex, font = font)
}

rrect(5, 5.2, 4, 0.8, "What is your question?",
      col = "white", border = "black", cex = 0.9, font = 2)

rrect(1.5, 3.2, 3.0, 0.7, '"How many units?"')
rrect(5.0, 3.2, 3.0, 0.7, '"What precision at n?"')
rrect(8.5, 3.2, 3.0, 0.7, '"Can I detect a difference?"', cex = 0.72)

rrect(1.5, 1.2, 3.0, 1.7, paste0(
        "n_prop \u00b7 n_mean \u00b7 n_ratio\nn_cluster \u00b7 n_multi\n",
        "n_multi_cluster\nn_alloc \u00b7 n_twophase\n",
        "n_change \u00b7 n_panel\nn_pooled"),
      col = "white", border = "black", cex = 0.62)
rrect(5.0, 1.2, 3.0, 1.7, paste0(
        "prec_prop \u00b7 prec_mean \u00b7 prec_ratio\nprec_cluster \u00b7 prec_multi\n",
        "prec_multi_cluster\nprec_alloc \u00b7 prec_twophase\n",
        "prec_change \u00b7 prec_panel\nprec_pooled"),
      col = "white", border = "black", cex = 0.62)
rrect(8.5, 1.2, 3.0, 1.7, paste0(
        "power_prop \u00b7 power_mean\npower_did\n\n",
        "with overlap, the same\ntwo-occasion change"),
      col = "white", border = "black", cex = 0.62)

draw_arrow <- function(x0, y0, x1, y1)
  arrows(x0, y0, x1, y1, length = 0.08, lwd = 1.3, col = "grey40")

draw_arrow(3.5, 4.8, 1.5, 3.6)
draw_arrow(5.0, 4.8, 5.0, 3.6)
draw_arrow(6.5, 4.8, 8.5, 3.6)

draw_arrow(1.5, 2.85, 1.5, 2.1)
draw_arrow(5.0, 2.85, 5.0, 2.1)
draw_arrow(8.5, 2.85, 8.5, 2.1)
# the duality the section is about, drawn where the two families meet
arrows(3.0, 1.3, 3.5, 1.3, length = 0.08, lwd = 1.3, col = "grey40", code = 3)
text(3.25, 2.25, "round-trip", cex = 0.55, col = "grey40", font = 3)

par(op)
```

Because `n_*` and `prec_*` are S3 generics, you can pass a result object directly to its inverse. All design parameters travel with the object:

```{r round-trip}
# Compute sample size for a proportion
s <- n_prop(p = 0.3, moe = 0.05, deff = 1.5)
s

# What precision does this n achieve?
p <- prec_prop(s)
p

# Recover the original sample size
n_prop(p)
```

You can also override a parameter on the way back:

```{r round-trip-override}
n_prop(p, cv = 0.08)
```

## Sample sizes for proportions and means

### Proportions

Estimate vaccination coverage (expected ~70%) with a 5 percentage point margin of error:

```{r n-prop-basic}
n_prop(p = 0.70, moe = 0.05)
```

Four method options control how the margin of error is computed:

- **`"wald"`** (default): the textbook normal approximation
- **`"wilson"`**: Wilson score interval, better coverage near 0 or 1
- **`"logodds"`**: log-odds transformation. Its back-transformed limits bound the proportion, or equivalently its odds. They do not bound an odds ratio or size a logistic regression.
- **`"beta"`**: Korn-Graubard interval, for a proportion near a boundary where the expected count in the nearer category is small. It stays inside $[0, 1]$ by construction, and a `df` argument widens it for a variance estimated from few clusters.

```{r n-prop-methods}
n_prop(p = 0.05, moe = 0.02, method = "wald")
n_prop(p = 0.05, moe = 0.02, method = "wilson")
n_prop(p = 0.05, moe = 0.02, method = "logodds")
n_prop(p = 0.05, moe = 0.02, method = "beta", df = 25)
```

The four differ materially only when the expected count is small. For sizing that difference is rarely worth acting on, but for assessing an achieved design it often is. `?n_prop` has the full comparison.

Target a coefficient of variation instead of a margin of error:

```{r n-prop-cv}
n_prop(p = 0.70, cv = 0.05)
```

Or state the margin of error as a fraction of the proportion rather than in
percentage points. `rmoe = 0.07` at `p = 0.70` asks for the same interval as
`moe = 0.049`, and says it in the units UNICEF's Multiple Indicator Cluster
Surveys (MICS) state a requirement in:

```{r n-prop-rmoe}
n_prop(p = 0.70, rmoe = 0.07)
```

The three are targets on one design, not three designs. `moe` and `rmoe` fix
the interval the chosen `method` builds, `cv` fixes the sampling variance
underneath it, and every result reports all of them back whichever one you
supplied.

### Means

Estimate average household expenditure (variance = 40,000, mean = 800) with a margin of error of 30:

```{r n-mean}
n_mean(var = 40000, moe = 30)
```

### Common parameters

Most single-stage `n_*()` and `prec_*()` functions share these parameters:

| Parameter | Meaning |
|:----------|:--------|
| `deff` | Design effect multiplier (default 1) |
| `N` | Finite population size (enables FPC correction) |
| `alpha` | Significance level (default 0.05) |
| `resp_rate` | Expected response rate (inflates *n* by `1/resp_rate`) |

Dividing by `resp_rate` sets an expected respondent count and evaluates variance at that net size. It assumes response is ignorable under the planned adjustment. Inflation does not remove nonresponse bias or include variation from response weights. Use sensitivity analysis over plausible rates, and see the allocation article for explicit nonresponse follow-up.

```{r common-params}
n_prop(p = 0.70, moe = 0.05, deff = 1.5, N = 50000, resp_rate = 0.85)
```

The design effect can be less than 1 for efficient designs (e.g. well-stratified or PPS samples). svyplan accepts any `deff > 0`.

`n_cluster()` / `prec_cluster()` use cluster-specific inputs instead (`stage_cost`, `icc`, `unit_relvar`, and optionally `resp_rate_psu`), and `n_alloc()` / `prec_alloc()` have their own allocation-oriented interfaces.

### Survey plan profiles

When the same `deff`, `N`, `resp_rate`, or `alpha` apply across many single-stage calls, bundle them into a `svyplan()` profile to avoid repetition:

```{r svyplan-profile}
plan <- svyplan(deff = 1.5, resp_rate = 0.85, N = 50000)

# Pass as a named argument
n_prop(p = 0.3, moe = 0.05, plan = plan)
n_mean(var = 100, moe = 2, plan = plan)
power_prop(p1 = 0.30, p2 = 0.35, plan = plan)
```

The profile can also hold cluster context (`stage_cost`, `icc`) for `n_cluster()`:

```{r svyplan-cluster}
cl_plan <- svyplan(stage_cost = c(500, 50), icc = 0.05, resp_rate = 0.85)
n_cluster(cv = 0.05, plan = cl_plan)
```

Piping works with both positional and named arguments:

```{r svyplan-pipe}
plan |> n_prop(0.3, moe = 0.05)
plan |> n_prop(p = 0.3, moe = 0.05)
plan |> power_mean(200, effect = 5)
```

Explicit arguments always override plan defaults:

```{r svyplan-override}
# plan has deff = 1.5, but deff = 2.0 wins here
n_prop(p = 0.3, moe = 0.05, plan = plan, deff = 2.0)
```

## Ratios of two totals

Some headline indicators are neither a mean nor a proportion but a ratio of two totals estimated from the same sample: consumption per person, yield per hectare, spending per pupil. `n_ratio()` sizes for the ratio itself.

Give it the anticipated ratio, the coefficient of variation of each component, and the correlation between them across units:

```{r n-ratio-basic}
# Per-capita household consumption: total consumption over total members
n_ratio(r = 420, cv_num = 1.20, cv_den = 0.45, component_cor = 0.65, cv = 0.05)
```

The printed `unit_relvar` is the coefficient the size is driven by. It is the relative variance of the linearized variable $e = y - Rx$, and the correlation is what makes it small: a numerator that moves with its denominator produces a stable ratio even when both components are individually variable. Moving `component_cor` from 0.30 to 0.90, with everything else held, nearly halves the required sample:

```{r n-ratio-correlation}
predict(
  n_ratio(r = 420, cv_num = 1.20, cv_den = 0.45, component_cor = 0.65, cv = 0.05),
  data.frame(component_cor = c(0.30, 0.65, 0.90))
)
```

Targets work on the ratio's own scale. `moe = 20` asks for an interval of plus or minus 20 currency units per person, and `rmoe` states the same thing relative to `r`:

```{r n-ratio-targets}
n_ratio(r = 420, cv_num = 1.20, cv_den = 0.45, component_cor = 0.65,
        moe = 20, deff = 1.3, resp_rate = 0.85)
```

`prec_ratio()` is the inverse, and the pair round-trips like every other:

```{r n-ratio-roundtrip}
size <- n_ratio(r = 420, cv_num = 1.20, cv_den = 0.45,
                component_cor = 0.65, cv = 0.05)
prec_ratio(size)
```

What this is, and is not: the estimand is $R = T_y / T_x$, with numerator and denominator observed on the same units. It is not the mean of the unit ratios $y_i / x_i$, which is an ordinary mean and belongs in `n_mean()`. A rate whose denominator is a subpopulation count, such as coverage among an eligible group, is usually easier to plan as a domain proportion with `n_prop()`. `n_ratio()` earns its place when the denominator is a continuous per-unit quantity, where there is no domain share to inflate by.

One warning is worth knowing about. When the two coefficients of variation are close and the correlation is near its bound, the ratio is nearly constant and the size collapses toward zero. The result then rests entirely on a correlation almost nobody knows to three decimals, so the function says so:

```{r n-ratio-warning}
tryCatch(
  n_ratio(r = 420, cv_num = 1.20, cv_den = 1.19, component_cor = 0.9999,
          cv = 0.05),
  warning = function(w) conditionMessage(w)
)
```

### Clustered ratios

For a multistage design, the quantity that matters is the homogeneity of the linearized variable, not of either component. They can differ by an order of magnitude in the same population, so substituting the numerator's `icc` is not a conservative approximation but a different design.

Compute $e = y - Rx$ from pilot data and pass its components to `n_cluster()`:

```{r n-ratio-cluster}
set.seed(1)
psu <- rep(1:30, each = 20)
x <- rlnorm(600, 1, 0.4)
y <- 3 * x + rnorm(600, 0, 0.8)

R <- mean(y) / mean(x)
e <- y - R * x

vc <- suppressWarnings(varcomp(e, stage_id = list(psu)))
n_cluster(stage_cost = c(500, 50), icc = vc$icc,
          unit_relvar = var(e) / mean(y)^2,
          var_ratio = vc$var_ratio, cv = 0.05)
```

`e` has mean zero by construction, so `varcomp()` warns that its relative variances are infinite. Its `icc` and `var_ratio` do not depend on the mean and stay valid, which is why they are taken from the object while `unit_relvar` is supplied separately. Passing the whole `vc` object would carry the infinite value in with it.

## Evaluating precision

The `prec_*()` functions answer the reverse question: given a fixed sample size, what precision can you achieve?

```{r prec-basic}
prec_prop(p = 0.3, n = 400)

prec_mean(var = 40000, n = 300, mu = 800)
```

Design parameters work identically:

```{r prec-design}
prec_prop(p = 0.3, n = 400, deff = 1.5, resp_rate = 0.85)
```

### Solving for the level

`n_prop()` and `prec_prop()` are two readings of one equation relating the
proportion, the sample size, and the precision. Fixing any two returns the
third. Supplying `cv` in place of `p` takes the remaining direction and
returns the smallest proportion the design measures that precisely:

```{r prec-solve}
prec_prop(n = 1500, cv = 0.10, N = 2e6)
```

Because `se(p) / p` falls as `p` rises, the answer is a floor: every larger
proportion meets the target, and everything below it is too rare for this
design to report at that precision. It is the planning form of a publication
threshold, and `expected_cases` is printed beside it so the precision floor
and the `min_cases` floor `n_prop()` applies can be read together.

`predict()` traces the floor across designs, varying `cv` rather than the
proportion it now solves for:

```{r prec-solve-grid}
predict(prec_prop(n = 1500, cv = 0.10, N = 2e6),
        expand.grid(n = c(500, 1500, 5000)))
```

`prec_mean()` takes `cv` the same way. The standard error of a mean does not
involve the mean, so that inversion is a division rather than a search, and
it returns the magnitude:

```{r prec-solve-mean}
prec_mean(var = 25, n = 1500, cv = 0.10)$params$mu
```

Supplying `rmoe` asks the same question of the interval instead of the
variance: the smallest proportion the design can report at that relative
half-width, under the interval method it will actually use. Unlike the `cv`
solve, this one moves with `method`:

```{r prec-solve-rmoe}
prec_prop(n = 1500, rmoe = 0.20, method = "wald", N = 2e6)$params$p
prec_prop(n = 1500, rmoe = 0.20, method = "beta", N = 2e6)$params$p
```

Two properties bound that search, and both are reported rather than
approximated past. The Wilson and Korn-Graubard half-widths do not shrink to
zero as `p` approaches 1, so their relative margin of error has a positive
floor and a target below it is refused with the floor named. The
back-transformed log-odds half-width turns upward once the logit spread
outgrows `logit(p)`, near `p = 0.999` at `n = 1500` but as low as `p = 0.97`
at `n = 30`, so under `"logodds"` the search brackets at that turn and
returns its lower root.

### Confidence intervals

`confint()` extracts the expected confidence interval from sizing and precision results for proportions, means, ratios, changes, and pooled estimates. Cluster, allocation, power, and multi-indicator results have no single interval to report and are not supported. A mean needs `mu`, since an interval has to be centered somewhere:

```{r confint}
s <- n_prop(p = 0.3, moe = 0.05)
confint(s)

pr <- prec_prop(p = 0.3, n = 400)
confint(pr)

confint(pr, level = 0.99)

# a mean carries no center of its own, so supply mu
confint(n_mean(var = 100, moe = 2, mu = 50))

# a ratio always carries its center r
confint(n_ratio(r = 420, cv_num = 1.20, cv_den = 0.45,
                component_cor = 0.65, cv = 0.05))

# a repeated-survey result, the pooled level across two occasions
confint(prec_pooled(var = 100, n = 300, occasions = 2, mu = 50))
```

For proportions, Wald endpoints are truncated to [0, 1] when needed. The other methods stay within the parameter space by construction.

## Explore planning assumptions

A fitted result keeps its planning assumptions. `predict()` evaluates compatible parameter grids while holding the remaining assumptions, including the interval method, fixed:

```{r predict-taste}
baseline <- n_prop(p = 0.3, moe = 0.05, deff = 1.5)
predict(baseline, data.frame(resp_rate = c(0.7, 0.8, 0.9, 1.0)))
```

## Continue with the focused articles

The following detailed articles are available on the package website:

- [Allocation, domains, and two-phase designs](https://dickoa.gitlab.io/svyplan/articles/allocation.html)
- [Multistage survey planning](https://dickoa.gitlab.io/svyplan/articles/multistage.html)
- [Repeated surveys, panels, and rotation](https://dickoa.gitlab.io/svyplan/articles/repeated-surveys.html)
- [Power and sensitivity analysis](https://dickoa.gitlab.io/svyplan/articles/power.html)

## Scope

`svyplan` answers one question in several forms: how many units does a planned design need, and what precision does a given design achieve. Some neighbouring topics are deliberately outside that boundary. They are listed here so their absence reads as a decision rather than an oversight.

**Totals.** There is no `n_total()`. A total is a mean rescaled by *N*, so the CV target is identical and a margin of error on the total is `moe / N` on the mean. `?n_mean` has a worked section on sizing for a total.

**Ratio estimation as a variance-reduction technique.** Using an auxiliary variable to sharpen an estimate of a mean or total is analysis-stage work, and is not the same thing as planning for a ratio estimand, which `n_ratio()` covers. Its planning effect is a smaller residual variance, so supply it directly: fit the ratio model, take the residual variance, and pass it as `n_mean(var = )` with the mean of the analysis variable. A target stated as a CV then gives the same size as sizing for the ratio itself, because the two have the same relative variance.

**Regression coefficients.** Sizing for a regression slope is a general statistical problem rather than a survey one, covered by packages such as `pwr`. The survey-specific part, the variance inflation from a complex design, is already the `deff` argument. There is no `n_reg()`.

**Two-phase designs beyond stratification and follow-up.** `n_twophase()` allocates a two-phase sample, phase 1 measuring a variable the frame does not carry and phase 2 measuring the variable of interest on a subsample of the phase-1 units. Double sampling for stratification and nonresponse follow-up are the same problem in that function, differing only in which strata are marked `take_all`. What is *not* covered is the third member of the family, double sampling for a regression or ratio estimator, where the phase-1 variable is continuous rather than a stratifier. That is a two-phase allocation question and is separate from `n_ratio()`, which sizes a single-phase design for a ratio estimand. Its variance effect is expressible today as a design effect below 1, since a regression estimator multiplies the variance by $1 - \rho^2$, but the cost trade-off between the phases is not built in.

**More than three stages.** Two- and three-stage designs are supported. Surveys rarely go deeper, and the cost and variance models for a fourth stage would carry more assumptions than they would earn.

**Attrition, and choosing a rotation.** `n_panel()` is arithmetic on declared rates. It says what recruitment yields under a response rate and a retention curve, not where those rates come from. Attrition modelling, attrition weighting, and any adjustment for informative dropout are analysis-stage work and outside the boundary. Two consequences are worth stating. A unit lost at one wave is treated as lost for good, so a panel whose nonrespondents return later does better than planned here, and planning this way over-issues. Also, `design_overlap()` reads a schedule that has already been chosen. It does not solve the inverse problem of finding a schedule that delivers a target overlap while limiting how many times a unit may be interviewed.

**Targeting nonresponse follow-up.** Deciding *how much* follow-up to do is answered by `n_twophase()`. Deciding *which* cases to pursue is not, and is out of scope: prioritizing by response propensity, or adapting effort as fieldwork proceeds, needs paradata from the field that no planning stage has.

**PPS selection inside joint allocation.** `varcomp()` estimates variance components under PPS first-stage selection, but the joint allocator (`n_alloc()` with `measures` and `targets`) uses an anticipated variance model. Feeding it components estimated under PPS is therefore an approximation, and the more unequal the sizes, the rougher it is. A two-stage joint allocation can identify certainty PSUs from a register. It does not support `take_all` in a fixed-take multistage frame because a census of PSUs still leaves within-PSU sampling.

**Domains not identifiable on the frame.** Targets can be set per domain when domain membership is known before fielding, as it is for region or urban/rural. For domains realized only after contact, such as age or sex subgroups, the realized subsample is random and has to be inflated by hand from the expected domain share. The same line runs through `design_df()`: a domain that is a union of whole strata gets an exact degrees-of-freedom count in `$domains`, while one cutting across strata does not, because counting the PSUs that contain its members needs frame information a plan does not carry.

**Converting between listing units and analysis units.** Sizing works in the units the precision target refers to. Converting a size in eligible persons to a number of households to list, given a mean household size and an eligibility rate, is arithmetic the user does explicitly. The MICS and DHS section above and the README both show the calculation.
