---
title: "Survey sample size planning with svyplan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survey sample size planning with svyplan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(svyplan)
```

`svyplan` provides a toolkit for survey sample size determination. It covers sample sizes for proportions and means, precision analysis, multistage cluster allocation, multi-indicator optimization, strata boundary optimization, and power analysis. The core sizing and precision functions use paired S3 interfaces for round-trip conversions. Sensitivity analysis via `predict()` is available for supported result classes.

## Function ecosystem

The package exports 31 functions organized into six families. The diagram below shows how they relate to each other.

```{r ecosystem-diagram, echo = FALSE, fig.height = 6.5, fig.width = 7}
op <- par(mar = c(0, 0, 0, 0), family = "sans")
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 9.5))

draw_box <- function(x, y, w, h, label, sublabel = NULL,
                     col = "white", border = "black") {
  rect(x - w / 2, y - h / 2, x + w / 2, y + h / 2,
       col = col, border = border, lwd = 1.5)
  text(x, y + if (!is.null(sublabel)) h / 2 - 0.22 else 0,
       label, cex = 0.85, font = 2)
  if (!is.null(sublabel))
    text(x, y - 0.12, sublabel, cex = 0.65, col = "grey30")
}

draw_arrow <- function(x0, y0, x1, y1, both = FALSE, ...) {
  arrows(x0, y0, x1, y1, length = 0.10, lwd = 1.5,
         col = "grey40", code = if (both) 3L else 2L, ...)
}

draw_box(5, 8.5, 3.8, 1.0, "Plan Profile",
         "svyplan")

draw_box(2.5, 6.5, 3.8, 1.9, "Sample Size",
         paste0("n_prop \u00b7 n_mean \u00b7 n_cluster\n",
                "n_multi \u00b7 n_multi_cluster\nn_alloc \u00b7 n_twophase\n",
                "n_change \u00b7 n_panel \u00b7 n_pooled"))
draw_box(7.5, 6.5, 3.8, 1.9, "Precision",
         paste0("prec_prop \u00b7 prec_mean \u00b7 prec_cluster\n",
                "prec_multi \u00b7 prec_multi_cluster\n",
                "prec_alloc \u00b7 prec_twophase\n",
                "prec_change \u00b7 prec_panel\nprec_pooled"))
draw_box(2.5, 3.5, 3.8, 1.7, "Design Components",
         paste0("varcomp \u00b7 design_effect\n",
                "effective_n \u00b7 design_df\n",
                "design_overlap \u00b7 design_schedule"))
draw_box(7.5, 3.5, 3.8, 1.4, "Power Analysis",
         "power_prop \u00b7 power_mean \u00b7 power_did")
draw_box(5, 1, 3.8, 1.2, "Stratification",
         "strata_bound")

draw_arrow(3.5, 8.0, 2.5, 7.5)
draw_arrow(6.5, 8.0, 7.5, 7.5)

draw_arrow(4.4, 6.9, 5.6, 6.9, both = TRUE)
text(5, 7.3, "round-trip", cex = 0.6, col = "grey40", font = 3)

draw_arrow(2.5, 4.25, 2.5, 5.5)
draw_arrow(4.5, 3.4, 5.5, 3.4)
draw_arrow(7.5, 4.25, 7.5, 5.5)

text(1.6, 4.9, "feeds into", cex = 0.55, col = "grey40", font = 3, srt = 90)
text(5.0, 3.7, "evaluate", cex = 0.55, col = "grey40", font = 3)
text(8.3, 4.8, "informs", cex = 0.55, col = "grey40", font = 3, srt = 90)

par(op)
```

| Family      | Functions                                                        | Purpose                                |
|:------------|:-----------------------------------------------------------------|:---------------------------------------|
| Plan        | `svyplan`                                                        | Reusable profile for design defaults   |
| Sample size | `n_prop`, `n_mean`, `n_cluster`, `n_multi`, `n_multi_cluster`, `n_alloc`, `n_twophase`, `n_change`, `n_panel`, `n_pooled` | "How many units do I need?"        |
| Precision   | `prec_prop`, `prec_mean`, `prec_cluster`, `prec_multi`, `prec_multi_cluster`, `prec_alloc`, `prec_twophase`, `prec_change`, `prec_panel`, `prec_pooled` | "What precision does *n* achieve?" |
| Power       | `power_prop`, `power_mean`, `power_did`                          | "Can I detect a difference?"            |
| Design      | `varcomp`, `design_effect`, `effective_n`, `design_df`, `design_overlap`, `design_schedule` | Variance components, design effects, degrees of freedom, and rotation schedules |
| Strata      | `strata_bound`                                                   | Optimal stratification boundaries      |

The same table read by question rather than by family, since several questions cut across the rows:

| Question | Where |
|:---------|:------|
| How many units for a target precision on a level? | `n_prop`, `n_mean`, and the cluster, allocation and two-phase members |
| How many for a target precision on a **change** between two occasions? | `n_change` |
| How many to **recruit** so that enough survive to a later wave? | `n_panel` |
| How many for a target precision on an **average across occasions**, an annual figure from quarterly rounds? | `n_pooled` |
| What precision does a design I already have achieve? | the `prec_*` counterpart of each |
| Can I detect a difference, and how large a one? | `power_prop`, `power_mean`, `power_did` |
| Can I detect a change over two occasions? | `power_prop` and `power_mean` with `overlap` set, which is the same variance `n_change` inverts |
| What overlap does my rotation schedule produce? | `design_overlap` |
| What should each startup or intake cohort issue, and when is it active? | `design_schedule` |
| What design effect, effective size or degrees of freedom will the design have? | `design_effect`, `effective_n`, `design_df`, `varcomp` |

Most planning functions return structured S3 objects. Available methods depend on the result class. See the relevant help page for details.

## The n_\*/prec_\* duality

The most distinctive feature of svyplan is the bidirectional link between sample size and precision functions. Every `n_*` function has a `prec_*` counterpart, and each pair supports round-trip conversion. Under the same formula and continuous design assumptions, converting a result to its counterpart and back reproduces the original target. Changing methods or using an operational integer allocation can change the result.

```{r decision-diagram, echo = FALSE, fig.height = 4, fig.width = 7}
op <- par(mar = c(0, 0, 0, 0), family = "sans")
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 6))

rrect <- function(x, y, w, h, label, col = "white", border = "black",
                  cex = 0.75, font = 1) {
  rect(x - w / 2, y - h / 2, x + w / 2, y + h / 2,
       col = col, border = border, lwd = 1.5)
  text(x, y, label, cex = cex, font = font)
}

rrect(5, 5.2, 4, 0.8, "What is your question?",
      col = "white", border = "black", cex = 0.9, font = 2)

rrect(1.5, 3.2, 3.0, 0.7, '"How many units?"')
rrect(5.0, 3.2, 3.0, 0.7, '"What precision at n?"')
rrect(8.5, 3.2, 3.0, 0.7, '"Can I detect a difference?"', cex = 0.72)

rrect(1.5, 1.2, 3.0, 1.7, paste0(
        "n_prop \u00b7 n_mean\nn_cluster \u00b7 n_multi\n",
        "n_multi_cluster\nn_alloc \u00b7 n_twophase\n",
        "n_change \u00b7 n_panel\nn_pooled"),
      col = "white", border = "black", cex = 0.62)
rrect(5.0, 1.2, 3.0, 1.7, paste0(
        "prec_prop \u00b7 prec_mean\nprec_cluster \u00b7 prec_multi\n",
        "prec_multi_cluster\nprec_alloc \u00b7 prec_twophase\n",
        "prec_change \u00b7 prec_panel\nprec_pooled"),
      col = "white", border = "black", cex = 0.62)
rrect(8.5, 1.2, 3.0, 1.7, paste0(
        "power_prop \u00b7 power_mean\npower_did\n\n",
        "with overlap, the same\ntwo-occasion change"),
      col = "white", border = "black", cex = 0.62)

draw_arrow <- function(x0, y0, x1, y1)
  arrows(x0, y0, x1, y1, length = 0.08, lwd = 1.3, col = "grey40")

draw_arrow(3.5, 4.8, 1.5, 3.6)
draw_arrow(5.0, 4.8, 5.0, 3.6)
draw_arrow(6.5, 4.8, 8.5, 3.6)

draw_arrow(1.5, 2.85, 1.5, 2.1)
draw_arrow(5.0, 2.85, 5.0, 2.1)
draw_arrow(8.5, 2.85, 8.5, 2.1)
# the duality the section is about, drawn where the two families meet
arrows(3.0, 1.3, 3.5, 1.3, length = 0.08, lwd = 1.3, col = "grey40", code = 3)
text(3.25, 2.25, "round-trip", cex = 0.55, col = "grey40", font = 3)

par(op)
```

Because `n_*` and `prec_*` are S3 generics, you can pass a result object directly to its inverse. All design parameters travel with the object:

```{r round-trip}
# Compute sample size for a proportion
s <- n_prop(p = 0.3, moe = 0.05, deff = 1.5)
s

# What precision does this n achieve?
p <- prec_prop(s)
p

# Recover the original sample size
n_prop(p)
```

You can also override a parameter on the way back:

```{r round-trip-override}
n_prop(p, cv = 0.08)
```

## Sample sizes for proportions and means

### Proportions

Estimate vaccination coverage (expected ~70%) with a 5 percentage point margin of error:

```{r n-prop-basic}
n_prop(p = 0.70, moe = 0.05)
```

Four method options control how the margin of error is computed:

- **`"wald"`** (default): the textbook normal approximation
- **`"wilson"`**: Wilson score interval, better coverage near 0 or 1
- **`"logodds"`**: log-odds transformation, the right choice when the plan will be reported as an odds ratio
- **`"beta"`**: Korn-Graubard interval, for a proportion whose expected number of positive counts is small. It stays inside $[0, 1]$ by construction, and a `df` argument widens it for a variance estimated from few clusters.

```{r n-prop-methods}
n_prop(p = 0.05, moe = 0.02, method = "wald")
n_prop(p = 0.05, moe = 0.02, method = "wilson")
n_prop(p = 0.05, moe = 0.02, method = "logodds")
n_prop(p = 0.05, moe = 0.02, method = "beta", df = 25)
```

The four differ materially only when the expected count is small. For sizing that difference is rarely worth acting on, but for assessing an achieved design it often is. `?n_prop` has the full comparison.

Target a coefficient of variation instead of a margin of error:

```{r n-prop-cv}
n_prop(p = 0.70, cv = 0.05)
```

Or state the margin of error as a fraction of the proportion rather than in
percentage points. `rmoe = 0.07` at `p = 0.70` asks for the same interval as
`moe = 0.049`, and says it in the units MICS and DHS state a requirement in:

```{r n-prop-rmoe}
n_prop(p = 0.70, rmoe = 0.07)
```

The three are targets on one design, not three designs. `moe` and `rmoe` fix
the interval the chosen `method` builds, `cv` fixes the sampling variance
underneath it, and every result reports all of them back whichever one you
supplied.

### Means

Estimate average household expenditure (variance = 40,000, mean = 800) with a margin of error of 30:

```{r n-mean}
n_mean(var = 40000, moe = 30)
```

### Common parameters

Most single-stage `n_*()` and `prec_*()` functions share these parameters:

| Parameter | Meaning |
|:----------|:--------|
| `deff` | Design effect multiplier (default 1) |
| `N` | Finite population size (enables FPC correction) |
| `alpha` | Significance level (default 0.05) |
| `resp_rate` | Expected response rate (inflates *n* by `1/resp_rate`) |

```{r common-params}
n_prop(p = 0.70, moe = 0.05, deff = 1.5, N = 50000, resp_rate = 0.85)
```

The design effect can be less than 1 for efficient designs (e.g. well-stratified or PPS samples). svyplan accepts any `deff > 0`.

`n_cluster()` / `prec_cluster()` use cluster-specific inputs instead (`stage_cost`, `icc`, `unit_relvar`, and optionally `resp_rate_psu`), and `n_alloc()` / `prec_alloc()` have their own allocation-oriented interfaces.

### Survey plan profiles

When the same `deff`, `N`, `resp_rate`, or `alpha` apply across many single-stage calls, bundle them into a `svyplan()` profile to avoid repetition:

```{r svyplan-profile}
plan <- svyplan(deff = 1.5, resp_rate = 0.85, N = 50000)

# Pass as a named argument
n_prop(p = 0.3, moe = 0.05, plan = plan)
n_mean(var = 100, moe = 2, plan = plan)
power_prop(p1 = 0.30, p2 = 0.35, plan = plan)
```

The profile can also hold cluster context (`stage_cost`, `icc`) for `n_cluster()`:

```{r svyplan-cluster}
cl_plan <- svyplan(stage_cost = c(500, 50), icc = 0.05, resp_rate = 0.85)
n_cluster(cv = 0.05, plan = cl_plan)
```

Piping works with both positional and named arguments:

```{r svyplan-pipe}
plan |> n_prop(0.3, moe = 0.05)
plan |> n_prop(p = 0.3, moe = 0.05)
plan |> power_mean(200, effect = 5)
```

Explicit arguments always override plan defaults:

```{r svyplan-override}
# plan has deff = 1.5, but deff = 2.0 wins here
n_prop(p = 0.3, moe = 0.05, plan = plan, deff = 2.0)
```

## Evaluating precision

The `prec_*()` functions answer the reverse question: given a fixed sample size, what precision can you achieve?

```{r prec-basic}
prec_prop(p = 0.3, n = 400)

prec_mean(var = 40000, n = 300, mu = 800)
```

Design parameters work identically:

```{r prec-design}
prec_prop(p = 0.3, n = 400, deff = 1.5, resp_rate = 0.85)
```

### Solving for the level

`n_prop()` and `prec_prop()` are two readings of one equation relating the
proportion, the sample size, and the precision. Fixing any two returns the
third. Supplying `cv` in place of `p` takes the remaining direction and
returns the smallest proportion the design measures that precisely:

```{r prec-solve}
prec_prop(n = 1500, cv = 0.10, N = 2e6)
```

Because `se(p) / p` falls as `p` rises, the answer is a floor: every larger
proportion meets the target, and everything below it is too rare for this
design to report at that precision. It is the planning form of a publication
threshold, and `expected_cases` is printed beside it so the precision floor
and the `min_cases` floor `n_prop()` applies can be read together.

`predict()` traces the floor across designs, varying `cv` rather than the
proportion it now solves for:

```{r prec-solve-grid}
predict(prec_prop(n = 1500, cv = 0.10, N = 2e6),
        expand.grid(n = c(500, 1500, 5000)))
```

`prec_mean()` takes `cv` the same way. The standard error of a mean does not
involve the mean, so that inversion is a division rather than a search, and
it returns the magnitude:

```{r prec-solve-mean}
prec_mean(var = 25, n = 1500, cv = 0.10)$params$mu
```

Supplying `rmoe` asks the same question of the interval instead of the
variance: the smallest proportion the design can report at that relative
half-width, under the interval method it will actually use. Unlike the `cv`
solve, this one moves with `method`:

```{r prec-solve-rmoe}
prec_prop(n = 1500, rmoe = 0.20, method = "wald", N = 2e6)$params$p
prec_prop(n = 1500, rmoe = 0.20, method = "beta", N = 2e6)$params$p
```

Two properties bound that search, and both are reported rather than
approximated past. The Wilson and Korn-Graubard half-widths do not shrink to
zero as `p` approaches 1, so their relative margin of error has a positive
floor and a target below it is refused with the floor named. The
back-transformed log-odds half-width turns upward once the logit spread
outgrows `logit(p)`, near `p = 0.999` at `n = 1500` but as low as `p = 0.97`
at `n = 30`, so under `"logodds"` the search brackets at that turn and
returns its lower root.

### Confidence intervals

`confint()` extracts the expected confidence interval from a single-indicator proportion or mean result, whether it came from an `n_*` or a `prec_*` function. Cluster, allocation, power and multi-indicator results have no single interval to report and are not supported. A mean needs `mu`, since an interval has to be centered somewhere:

```{r confint}
s <- n_prop(p = 0.3, moe = 0.05)
confint(s)

pr <- prec_prop(p = 0.3, n = 400)
confint(pr)

confint(pr, level = 0.99)

# a mean carries no center of its own, so supply mu
confint(n_mean(var = 100, moe = 2, mu = 50))
```

For proportions, the interval is clamped to [0, 1].

## Multi-indicator surveys

Household surveys like Demographic and Health Surveys (DHS) or Multiple Indicator Cluster Survey (MICS) track many indicators simultaneously. Each indicator has its own expected prevalence, precision target, and design effect. `n_multi()` finds the sample size that satisfies all targets at once.

```{r n-multi}
targets <- data.frame(
  name = c("stunting", "vaccination", "anemia"),
  p    = c(0.25, 0.70, 0.12),
  moe  = c(0.05, 0.05, 0.03),
  deff = c(2.0, 1.5, 2.5)
)

n_multi(targets)
```

The binding indicator (marked with `*`) drives the overall sample size. The other indicators are estimated with better precision than requested.

Proportion rows accept the same four methods as `n_prop()`, either for the whole table through `prop_method` or per row through a `prop_method` column. A `df` column sets each row's interval quantile, so one table can mix an ordinary indicator with a rare one estimated from few clusters:

```{r n-multi-methods}
targets_rare <- data.frame(
  name = c("stunting", "cocaine_use"),
  p    = c(0.25, 0.02),
  moe  = c(0.05, 0.01),
  prop_method = c("wald", "beta"),
  df   = c(NA, 25)
)

n_multi(targets_rare)
```

### Achieved precision

`prec_multi()` computes the achieved precision for each indicator at the chosen *n*:

```{r prec-multi}
plan <- n_multi(targets)
prec_multi(plan)
```

### Domain targets

Surveys often need adequate precision within subpopulations (urban/rural, regions). Add domain columns to the targets data frame and `n_multi()` optimizes each domain separately:

```{r n-multi-domain}
targets_dom <- data.frame(
  name       = rep(c("stunting", "vaccination", "anemia"), each = 2),
  residence  = rep(c("urban", "rural"), 3),
  p          = c(0.18, 0.32, 0.80, 0.60, 0.08, 0.16),
  moe        = c(0.05, 0.05, 0.05, 0.05, 0.03, 0.03),
  deff       = c(1.5, 2.5, 1.2, 1.8, 2.0, 3.0)
)

n_multi(targets_dom, domains = "residence")
```

Domain columns are specified explicitly via the `domains` parameter, and `domain_sampling` decides how the per-domain requirements combine into the single `n` that is reported. The default `"separate"` treats the domains as disjoint quotas drawn independently, the usual case for regional domains in one national survey, and reports their sum: every quota has to be met, so nothing smaller delivers the design. `"natural"` treats them as analytic domains arising at their own rate inside one sample, requires a `share` column, and reports `max(.n / share)`, the size whose expected yield meets every quota.

Neither reading is the largest single domain requirement. That number is reported separately as `n_domain_max`, and it is what `binding` refers to. Use `$domains` to see each domain's own sample size.

### MICS/DHS-style relative margin of error

Programs like UNICEF MICS and DHS express precision as a **relative margin of error**, the margin of error as a fraction of the indicator. `rmoe` states it directly, on the scalar functions and as an indicator column.

`rmoe` is `moe / p`, so it fixes the same interval `moe` does while scaling with the estimate the way `cv` does. That places the package's four precision quantities on two axes: `se` and `cv` describe the sampling variance, while `moe` and `rmoe` describe the interval drawn around it.

|          | Absolute | Relative |
|:---------|:---------|:---------|
| Sampling | `se`     | `cv`     |
| Interval | `moe`    | `rmoe`   |

The sampling row is free of `alpha` and of the interval method, whereas the interval row depends on both. A rare outcome shows the split at its widest:

```{r rmoe-two-axes}
sapply(c("wald", "wilson", "logodds", "beta"), function(m) {
  r <- prec_prop(p = 0.02, n = 900, deff = 2, df = 25, method = m)
  c(se = r$se, cv = r$cv, moe = r$moe, rmoe = r$rmoe)
})
```

`se` and `cv` do not move across the four columns, but `moe` and `rmoe` do. So `rmoe = q * cv`, where `q` is the interval quantile the design uses (`qnorm(1 - alpha / 2)`, or the `t` quantile on `df` degrees of freedom when `df` is supplied, as here), holds under Wald alone. Under the other three, reading a relative margin of error off the CV that way overstates the precision by 5 to 10 percent in this design, which is why a relative margin of error is a target in its own right rather than a rescaled `cv`.

```{r n-multi-rme}
# MICS-style: 12% relative MOE per domain, domain-specific deff and prevalence
targets_mics <- data.frame(
  name      = rep(c("stunting", "vaccination"), each = 3),
  region    = rep(c("North", "Central", "South"), 2),
  p         = c(0.35, 0.25, 0.40, 0.60, 0.75, 0.55),
  rmoe      = 0.12,
  deff      = c(2.5, 2.0, 3.0, 1.5, 1.2, 1.8),
  resp_rate = c(0.90, 0.85, 0.95, 0.90, 0.85, 0.95)
)

n_multi(targets_mics, domains = "region")
```

The MICS template reports sample size in **households**. svyplan returns the number of **individuals** in the target population. To convert, divide by the expected number of eligible individuals per household: `n_hh = ceiling(n / (pb * hh_size))`, where `pb` is the share of the target population in the total population and `hh_size` is the average household size.

### Minimum sample size per domain

The `min_n_domain` parameter sets a floor for the per-domain sample size (`n_alloc()` uses `min_n_stratum` for its per-stratum floor):

```{r n-multi-min-n}
n_multi(targets_dom, domains = "residence", min_n_domain = 300)
```

### Mixing proportions and means

Targets can mix proportion and mean indicators. Use `var` and `mu` columns for means (leave `p` as `NA`), and `p` for proportions (leave `var`/`mu` as `NA`):

```{r n-multi-mixed}
targets_mixed <- data.frame(
  name = c("vaccination", "expenditure"),
  p    = c(0.70, NA),
  var  = c(NA, 40000),
  mu   = c(NA, 800),
  cv   = c(0.05, 0.10),
  deff = c(1.5, 2.0)
)

n_multi(targets_mixed)
```

## Multistage cluster designs

For two-stage or three-stage designs, the question is not just "how many?" but "how many clusters and how many units per cluster?" The answer depends on the cost structure and the degree of clustering.

### Variance components from frame data

If frame data with cluster identifiers is available, `varcomp()` estimates between-cluster and within-cluster variance components:

```{r varcomp}
set.seed(1)
n_clust <- 80
n_hh <- 15
cluster_id <- rep(seq_len(n_clust), each = n_hh)
cluster_mean <- rnorm(n_clust, mean = 50, sd = 10)
expenditure <- cluster_mean[cluster_id] + rnorm(n_clust * n_hh, sd = 20)
frame <- data.frame(cluster = cluster_id, expenditure = expenditure)

vc <- varcomp(expenditure ~ cluster, data = frame)
vc
as.data.frame(vc)
```

The `icc` value is the survey-planning measure of homogeneity. It feeds directly into `n_cluster()` and the cluster-planning mode of `design_effect()`. This `icc` is not a generic mixed-model ICC: `varcomp()` returns the bounded planning quantity from Valliant, Dever, and Kreuter (2018), and that is the scale expected by the cluster-planning functions. When `icc` is near 0, most variation is within clusters, so the analytical cluster optimum tends to favor many interviews in very few PSUs. When `icc` is near 1, most variation is between clusters, so the optimum tends to favor very few interviews in many PSUs. Both are degenerate boundary cases for the closed-form optimizer, so `n_cluster()` rejects values numerically too close to 0 or 1.

### Optimal two-stage allocation

Given per-stage costs and the variance structure, `n_cluster()` finds the allocation that either minimizes cost for a target CV, or minimizes CV for a given budget:

```{r n-cluster-2stage}
# Minimize cost to achieve CV = 0.05
n_cluster(stage_cost = c(500, 50), icc = vc, cv = 0.05)

# Maximize precision within a budget
n_cluster(stage_cost = c(500, 50), icc = vc, budget = 100000)
```

Passing `icc = vc` (the varcomp object) automatically extracts `icc`, `unit_relvar`, and `var_ratio`.

### Three-stage designs

Add a third cost element and a second icc for three-stage designs:

```{r n-cluster-3stage}
n_cluster(
  stage_cost  = c(1000, 200, 20),
  icc = c(0.01, 0.05),
  cv    = 0.05
)
```

### Response rate in cluster designs

A cluster design can lose units at more than one stage, and the two losses are not interchangeable, so they are named for the stage they act on. `resp_rate_psu` is the share of selected clusters that can be worked at all: whole PSUs lost to insecurity, flooding, or an inaccessible road. Losing a PSU removes its entire contribution, so it deflates every variance term alike and the first-stage sample is inflated by `1 / resp_rate_psu` to cover it.

In budget mode, applying a response rate does not change the allocation because the budget constrains the design. The achieved CV worsens instead. In CV mode:

```{r n-cluster-resp}
n_cluster(stage_cost = c(500, 50), icc = vc, cv = 0.05, resp_rate_psu = 0.90)
```

Household or individual nonresponse is a different quantity acting at a different stage, and each has its own argument: `resp_rate_ssu` for second-stage units in a three-stage design, and `resp_rate` for the ultimate units. Naming a stage the design does not have is an error rather than a silently ignored argument, because a dropped response rate plans a design with none.

The three are not interchangeable. Losing whole PSUs deflates every variance term alike, whereas losing ultimate units shrinks the realized cluster and so changes the clustering penalty too. That is why the later-stage rates move the cost-optimal take while the PSU rate does not:

```{r n-cluster-resp-stages}
# a DHS-shaped design: EAs, households within EAs, one person per household
n_cluster(stage_cost = c(500, 100, 50), icc = c(0.02, 0.05), cv = 0.05,
          resp_rate_psu = 0.95, resp_rate_ssu = 0.90, resp_rate = 0.85)
```

Sizes and costs stay gross: `$n` counts the units to issue and `$cost` pays for them, while the variance reads what they realize. Lumping the later rates into `resp_rate_psu` is conservative but wrong, and the gap grows with the between-PSU homogeneity.

### Fixed overhead costs

The standard cost model is purely linear: `C = c1*n_psu + c2*n_psu*n_per_psu`. Real surveys also have a fixed overhead (training, infrastructure, setup). The `fixed_cost` parameter adds this term: `C = C0 + c1*n_psu + c2*n_psu*n_per_psu`. In budget mode, only `budget - fixed_cost` is available for the variable component, reducing n_psu while leaving the optimal n_per_psu/n_psu ratio unchanged. In CV mode, all sample sizes stay the same and `fixed_cost` is simply added to the total cost:

```{r n-cluster-fixed}
# Budget mode: fixed overhead reduces the allocatable budget
n_cluster(stage_cost = c(500, 50), icc = vc, budget = 100000, fixed_cost = 5000)

# CV mode: same allocation, cost increases by fixed_cost
n_cluster(stage_cost = c(500, 50), icc = vc, cv = 0.05, fixed_cost = 5000)
```

The same parameter is available in `n_multi_cluster()` for multistage
multi-indicator designs.

### Evaluating cluster allocations

`prec_cluster()` computes the achieved CV for any allocation:

```{r prec-cluster}
alloc <- n_cluster(stage_cost = c(500, 50), icc = vc, budget = 100000)
prec_cluster(alloc)

# Arbitrary allocation
prec_cluster(n = c(50, 12), icc = 0.05)
```

### Design effects

`design_effect()` builds the design effect you expect a plan to produce, from the design features you are choosing. The clustering component uses the same variance model as `n_cluster()` and `prec_cluster()`, so the two always agree:

```{r design-effect}
design_effect(icc = 0.05, n_per_psu = 20)
```

Components are selected by the arguments you supply and multiply together. Unequal weighting comes from planned weights or from a planned allocation's stratum sizes and takes, and the stratification gain (the only component that can fall below 1) from the stratum standard deviations and means:

```{r design-effect-components}
frame <- data.frame(
  N = c(50000, 120000), n = c(600, 400), sd = c(12, 20), mean = c(55, 48)
)
deff <- design_effect(icc = 0.05, n_per_psu = 20, strata = frame)
deff
```

The result is a plain number wherever one is expected, and `effective_n()` is its mirror:

```{r design-effect-use}
n_prop(p = 0.3, moe = 0.05, deff = deff)
effective_n(deff, n = 1000)
```

You can also read the features off a plan you already built, including an `n_alloc()` allocation:

```{r design-effect-plan}
design_effect(n_cluster(stage_cost = c(500, 50), icc = 0.05, cv = 0.05))
```

These are planning quantities. After data collection, compute the design effect the sample actually achieved with `survey::svymean(..., deff = TRUE)`, which uses the realized weights, strata, and clusters.

### Multi-indicator multistage

For surveys with multiple indicators in a cluster design, provide
`icc_psu` columns and stage costs to `n_multi_cluster()`:

```{r n-multi-cluster}
targets_ms <- data.frame(
  name   = c("stunting", "vaccination"),
  p      = c(0.25, 0.70),
  cv     = c(0.10, 0.08),
  icc_psu = c(0.05, 0.02)
)

n_multi_cluster(targets_ms, stage_cost = c(500, 50), budget = 80000)
```

## Strata boundaries

When a continuous auxiliary variable is available on the frame,
`strata_bound()` constructs cut points intended to reduce the sample size or
CV for a given number of strata. The iterative choices below are local
heuristics, so their result is a candidate solution rather than a certified
global minimum.

The boundaries are built from the distribution of `x` alone, and the variance
they minimize is that of the total or mean of `x`. Nothing models the study
variable's conditional mean or variance given `x`, so for any `y` other than
`x` the result is a proxy, good in proportion to how closely `y` tracks `x`.
The reported CV is likewise the CV for `x`. Stratifying an expenditure frame
to measure a poverty rate is the case this serves well, whereas stratifying
on a variable unrelated to the indicator is not.

```{r strata-bound}
set.seed(12345)
x <- rlnorm(5000, meanlog = 6, sdlog = 1.2)

strata_bound(x, n_strata = 4, n = 300, method = "cumrootf")
```

Four methods are available:

| Method       | Algorithm                                 | Speed    | Best for       |
|:-------------|:------------------------------------------|:---------|:---------------|
| `"cumrootf"` | Dalenius-Hodges cumulative root frequency | Fast     | General use    |
| `"geo"`      | Geometric progression                     | Fast     | Skewed data    |
| `"lh"`       | LH-inspired coordinate optimization       | Moderate | Fast local search |
| `"kozak"`    | Kozak-inspired random-restart local search| Slow     | More starting points |

The `"kozak"` heuristic explores multiple starts and can find a better local
solution on highly skewed distributions. Neither iterative method guarantees a
global optimum or claims to reproduce the published algorithm exactly:

```{r strata-kozak}
strata_bound(x, n_strata = 4, cv = 0.05, method = "kozak")
```

### Allocation methods

Allocation within strata is controlled by the `alloc` parameter:

- `"proportional"`: $n_h \propto N_h$
- `"neyman"` (default): $n_h \propto N_h \cdot S_h$ (minimizes national CV)
- `"optimal"`: $n_h \propto N_h \cdot S_h / \sqrt{c_h}$ (accounts for costs)
- `"power"`: Bankier (1988) compromise, $n_h \propto S_h \cdot N_h^{q}$,
  where the exponent `alloc_q` $\in [0, 1]$ controls the trade-off between
  national precision (`alloc_q` $= 1$, Neyman) and equal subnational CVs
  (`alloc_q` $= 0$)

```{r strata-power-alloc}
# Bankier power allocation (compromise, alloc_q = 0.5)
strata_bound(x, n_strata = 4, n = 300, alloc = "power", alloc_q = 0.5)
```

### Classifying new observations

`predict()` applies the learned boundaries to new data, returning a
factor:

```{r strata-predict}
sb <- strata_bound(x, n_strata = 4, n = 200, method = "cumrootf")
new_x <- c(100, 500, 1500, 5000)
predict(sb, newdata = new_x, labels = paste0("S", 1:4))
```

### Visualizing allocation

`plot()` shows the sampling fraction per stratum. Under Neyman allocation, high-variance strata are sampled more intensively. The dashed line marks the overall fraction:

```{r strata-plot, fig.width=7, fig.height=4}
plot(sb)
```

### Handing boundaries to the allocator

`$strata` carries the `N`, `sd`, and `mean` columns `n_alloc()` needs, so the
two halves of a stratified design connect directly:

```{r strata-handoff}
sb_cv <- strata_bound(x, n_strata = 4, cv = 0.05, method = "lh")
n_alloc(sb_cv$strata, cv = 0.05)
```

Both functions evaluate the same variance, so `cv` means the same thing to
each and `n` is a fielded sample in both. Design assumptions carry across
unchanged, whether passed directly or through a `svyplan()` profile:

```{r strata-deff}
plan <- svyplan(deff = 1.8, resp_rate = 0.85)
sb_deff <- strata_bound(x, n_strata = 4, cv = 0.05, plan = plan)
sb_deff$n
n_alloc(sb_deff$strata, cv = 0.05, plan = plan)$n
```

The remaining gap between the two totals is integer rounding, not a
difference in model: `strata_bound()` reports the whole-unit allocation in
`$strata$n`, while `n_alloc()` reports the continuous optimum.

A scalar `deff` does not move the cut points. It scales the variance of every
candidate boundary set equally, so what it changes is the `n` a target
requires, not where the strata divide.

## Stratified allocation

Given a sampling frame with stratum population sizes (`N`) and stratum standard deviations (`sd`), `n_alloc()` distributes a total sample across strata. This is the second half of a stratified design: `strata_bound()` finds the cut points, and `n_alloc()` allocates within strata.

```{r n-alloc-basic}
frame <- data.frame(
  N = c(4000, 3000, 3000),
  sd = c(10, 15, 8),
  mean = c(50, 60, 55)
)

# Fixed total n with Neyman allocation
n_alloc(frame, n = 600, alloc = "neyman")
```

Three solve modes are available: fixed total `n`, target `cv`, or `budget` constraint (when `unit_cost` is provided in the frame).

Constraints and domain-level targets are often the more practical use cases:

```{r n-alloc-constraints}
frame_constraints <- transform(
  frame,
  unit_cost = c(1, 1.5, 1),
  max_weight = c(25, 20, NA),
  take_all = c(FALSE, FALSE, TRUE)
)

n_alloc(frame_constraints, budget = 3500, alloc = "optimal", min_n_stratum = 40)
```

```{r n-alloc-domains}
frame_domains <- data.frame(
  province = c("North", "North", "South", "South"),
  stratum = c("Urban", "Rural", "Urban", "Rural"),
  N = c(2000, 3000, 1800, 3200),
  sd = c(12, 18, 10, 16),
  mean = c(55, 48, 58, 50)
)

# Minimum total n such that each province meets the CV target
n_alloc(frame_domains, domains = "province",
        cv = 0.04, alloc = "power", alloc_q = 0.3)
```

`prec_alloc()` computes the achieved precision for any allocation:

```{r prec-alloc}
prec_alloc(frame, n = c(200, 250, 150))
```

`$detail` carries the same numbers per stratum, so you can see where the
precision is actually being bought. `.share` is the stratum's contribution to
the design variance of the overall mean and sums to 1:

```{r prec-alloc-detail}
prec_alloc(frame, n = c(200, 250, 150))$detail[, c("stratum", "n", ".se", ".cv", ".share")]
```

With `domains`, `$domains` reports the same table `n_alloc()` returns, so a
design and its assessment compare row for row.

### Fieldwork that differs across strata

`deff` and `resp_rate` take one value per stratum, either as vectors or as
frame columns, for designs whose fieldwork is not uniform. A response rate
that varies changes where the sample should go, not just how much of it comes
back:

```{r n-alloc-per-stratum}
uneven <- data.frame(
  stratum   = c("Urban", "Peri-urban", "Rural"),
  N         = c(4000, 3000, 3000),
  sd        = c(10, 15, 8),
  mean      = c(50, 60, 55),
  resp_rate = c(0.92, 0.80, 0.65)
)
n_alloc(uneven, n = 600, alloc = "neyman")$detail[, c("stratum", "n", "n_int")]
```

A value shared by every stratum is a constant factor that cancels out of a
proportional weighting, so scalar `deff` and `resp_rate` behave exactly as
before.

### Joint indicator and domain targets

When one allocation must satisfy several estimands, use an atomic frame plus
long `measures` and `targets` tables. Atomic means that every frame row has a
single value for every domain classification in use. The example below crosses
region and residence, allowing those two domain systems to overlap without
requiring separate public functions or an `alloc` selector.

```{r n-alloc-joint-data}
joint_frame <- data.frame(
  stratum = c("NU", "NR", "SU", "SR"),
  region = c("North", "North", "South", "South"),
  residence = c("Urban", "Rural", "Urban", "Rural"),
  N = c(1000, 2000, 1500, 1000),
  unit_cost = c(1, 1.2, 1.5, 1)
)

joint_measures <- data.frame(
  stratum = rep(joint_frame$stratum, 2),
  name = rep(c("vaccination", "income"), each = 4),
  p = c(0.50, 0.40, 0.60, 0.30, rep(NA, 4)),
  mean = c(rep(NA, 4), 50, 55, 60, 45),
  sd = c(rep(NA, 4), 10, 12, 15, 9)
)

joint_targets <- data.frame(
  name = c("vaccination", "vaccination", "income"),
  domain = c(".overall", "region", "residence"),
  level = c(NA, "North", "Urban"),
  cv = c(0.05, 0.08, NA),
  moe = c(NA, NA, 2)
)
```

```{r n-alloc-joint-fit}
joint_fit <- n_alloc(
  joint_frame,
  measures = joint_measures,
  targets = joint_targets,
  min_n_stratum = 2
)
joint_fit
joint_fit$constraints
```

The optimizer minimizes variable cost subject to all three requirements.
`joint_fit$detail$n` is the continuous optimum and
`joint_fit$detail$n_int` is the deterministic feasible operational
recommendation. The latter is locally cleaned but is not claimed to be a
globally optimal integer solution. Both precision tables are retained:

```{r n-alloc-joint-assess}
prec_alloc(joint_fit)
prec_alloc(joint_fit, n = joint_fit$detail$n_int)
```

A modified named allocation can be assessed in any order because names are
matched exactly to `frame$stratum`.

The precision a joint allocation plans for can be checked against the
precision it delivers, by executing the design repeatedly against a known
population and comparing empirical CVs with `$constraints$.achieved`. Two
things are worth fixing before reading such a check. Compare against
`$constraints$.achieved` rather than against the target, since integerization
leaves the realized design a little better than what was asked for, and set
the acceptance band in advance from the number of replicates, since an
empirical CV from a few hundred draws carries its own error. The recipe needs
a sampling engine to execute the design, so it belongs with one rather than
here.

#### Fixed-take multistage joint allocation

The same public interface supports two- and three-stage designs when every
later-stage take is fixed. The frame owns stage populations, takes, and costs.
The measures table owns indicator-specific stage homogeneity parameters. Only
`n_psu` is optimized. Public `n`, planning bounds, and precision assessments
remain in ultimate-unit units.

This three-stage example samples PSUs, then a fixed number of SSUs per PSU,
then a fixed number of ultimate units per SSU:

```{r n-alloc-joint-three-stage}
joint_frame_3stage <- within(joint_frame, {
  unit_cost <- NULL
  N_psu <- c(100, 160, 120, 90)
  N_ssu <- c(600, 1200, 900, 500)
  n_per_psu <- c(8, 10, 12, 7)
  n_per_ssu <- c(4, 5, 3, 4)
  cost_psu <- c(300, 400, 450, 350)
  cost_ssu <- c(25, 30, 35, 28)
  cost_tsu <- c(5, 6, 7, 5)
})

joint_measures_3stage <- transform(
  joint_measures,
  icc_psu = rep(c(0.03, 0.05, 0.08, 0.04), 2),
  icc_ssu = rep(c(0.10, 0.08, 0.12, 0.06), 2),
  var_ratio_psu = 1,
  var_ratio_ssu = 1
)

joint_fit_3stage <- n_alloc(
  joint_frame_3stage,
  measures = joint_measures_3stage,
  targets = joint_targets,
  min_n_stratum = 20
)
joint_fit_3stage$detail[, c(
  "stratum", "n_psu_int", "n_per_psu", "n_per_ssu", "n_int"
)]
prec_alloc(joint_fit_3stage, n = joint_fit_3stage$detail$n_int)
```

The exact operational identity is
`n_int = n_psu_int * n_per_psu * n_per_ssu`. The standard multistage forms
use `N_psu`, while three-stage frames also require the stratum's total `N_ssu`.
Fixed takes must be whole numbers. `take_all` is unavailable because taking
every PSU alone does not establish a census of ultimate units.

#### Certainty PSUs from a register

For a two-stage joint allocation, a PSU register can replace `N_psu`. It has
one row per PSU with `stratum` and `N`, and its totals must match `frame$N`.
The allocation identifies PSUs whose implied PPS inclusion probability reaches
one. A `certainty` column can add PSUs to that part of the design.

```{r n-alloc-joint-certainty}
certainty_frame <- joint_frame_3stage[, setdiff(names(joint_frame_3stage),
                                                  c("N_ssu", "n_per_ssu", "cost_tsu"))]
certainty_frame$N_psu <- NULL

psu_register <- do.call(rbind, lapply(seq_len(nrow(certainty_frame)), function(i) {
  N <- certainty_frame$N[i]
  size <- rep(10, 20)
  size[1] <- floor(N / 3)
  size[2] <- N - sum(size[-2])
  data.frame(stratum = certainty_frame$stratum[i], N = size)
}))

certainty_fit <- n_alloc(
  certainty_frame,
  measures = joint_measures_3stage[, setdiff(names(joint_measures_3stage),
                                              c("icc_ssu", "var_ratio_ssu"))],
  targets = joint_targets,
  psu = psu_register,
  min_n_stratum = 20
)
certainty_fit$detail[, c("stratum", "n_psu_certain", "n_psu_draw", "n_int")]
```

`$psu` carries the classification, threshold, and source for every PSU. The
operational allocation uses whole takes in certainty PSUs and whole PSUs in
the remainder. `prec_alloc(certainty_fit)` reproduces its precision exactly.

### Stratified two-stage designs

Most household surveys stratify first and cluster within each stratum,
sampling enumeration areas and then households. Adding a `icc_psu`
column to the frame switches `n_alloc()` to this design. The per-stratum
homogeneity can come straight from `varcomp()` with `strata`, whose
output columns are named to match the frame:

```{r n-alloc-cluster-varcomp}
set.seed(3)
listing <- data.frame(
  region = rep(c("North", "South"), each = 600),
  ea = rep(1:60, each = 20),
  income = rnorm(1200, rep(c(50, 70), each = 600), 15) +
    rep(rnorm(60, 0, 6), each = 20)
)
vc_region <- varcomp(income ~ ea, data = listing, strata = ~region)
vc_region
```

```{r n-alloc-cluster}
frame_2stage <- merge(
  data.frame(stratum = c("North", "South"), N = c(40000, 60000)),
  as.data.frame(vc_region)[, c("stratum", "sd", "mean", "icc_psu", "var_ratio_psu")],
  by = "stratum"
)
frame_2stage$cost_psu <- c(400, 550)
frame_2stage$cost_ssu <- c(45, 60)

res_2stage <- n_alloc(frame_2stage, cv = 0.02)
res_2stage
res_2stage$detail[, c("stratum", "n_int", "n_per_psu", "n_psu_int")]
```

Each stratum gets its cost-optimal cluster take (fix it instead with a
`n_per_psu` column), and the PSU counts fall out of the element
allocation. Budget mode, `min_n_stratum`, and `max_weight` work the same
way here. `take_all` does not apply: taking every PSU leaves the
within-PSU take in force, so it does not enumerate a stratum, and it is
refused rather than half-honoured.

An `N_psu` column caps each stratum at the PSUs it actually has, both in
the continuous allocation and in the whole-unit design. Without it the
allocation can ask for more clusters than a stratum contains:

```{r n-alloc-cluster-npsu}
frame_bounded <- frame_2stage
frame_bounded$N_psu <- c(40, 900)

res_bounded <- n_alloc(frame_bounded, n = 3000)
res_bounded$detail[, c("stratum", "n_psu", "n_psu_int", "N_psu",
                       ".psu_frac", ".bound_source")]
```

North holds only 40 enumeration areas, so its allocation stops there and
`.bound_source` names the constraint that stopped it. Asking for precision
the remaining PSUs cannot deliver is an error rather than a silent
overshoot, and it says the ceiling came from PSU availability under a
with-replacement first stage rather than from the population.

`N_psu` is a feasibility constraint only. It does not switch on a
first-stage finite population correction, so precision still uses a
with-replacement first stage and stays conservative when a design takes an
appreciable share of the available PSUs. `print()` says so when that
happens, and `.psu_frac` reports the share.

## Two-phase designs

In a two-phase (or double) sample, a large cheap phase 1 measures something that helps target the expensive measurement, and a subsample is carried into phase 2. `n_twophase()` allocates both phases at once, minimizing cost for a target CV or CV for a fixed budget.

The frame is one row per phase-2 stratum, similarly to what is used in `n_alloc()` :

```{r twophase-strat}
frame <- data.frame(
  stratum   = c("A", "B", "C", "D"),
  N         = c(3500, 2500, 2500, 1500),
  sd        = c(12, 25, 8, 40),
  mean      = c(40, 70, 35, 90),
  unit_cost = c(2, 5, 1, 9)
)

n_twophase(frame, phase1_cost = 1, budget = 50000)
```

The `nu` column is the fraction of each phase-1 stratum carried forward. Its shape is Neyman-like, proportional to $S_h/\sqrt{c_h}$, but the overall scale is set by the *between*-stratum variance: the weaker the stratification, the closer every fraction moves to 1, meaning keep everything phase 1 found.

### Two design effects, not one

Neither phase has to be a simple random sample, and the two carry *separate* design effects because they apply to different things. `phase1_deff` inflates the between-stratum component, the part phase 1 is responsible for. A `deff` column inflates each stratum's within-stratum residual, the part phase 2 has to measure. `single_deff` describes the comparator, which need not be fielded like phase 2.

```{r twophase-deff}
clustered <- transform(frame, deff = c(1.2, 2.5, 0.8, 1.7))

n_twophase(clustered, phase1_cost = 1, budget = 50000,
           phase1_deff = 2.2, single_deff = 2.9)
```

Inflating the combined variance by one design effect instead would be a different and wrong model. The two also pull against each other, and the trap runs in the direction that looks attractive: a stratifier built purely from between-cluster structure shrinks the residual, lowering the stratum `deff`, but it makes the fitted values nearly constant within clusters and so drives `phase1_deff` up toward the cluster size. Under a clustered phase 1 that first term can dominate, and the design loses to a plain sample even though the residual design effect looks favorable.

Every result carries the single-phase comparison, because two-phase sampling is not always an improvement. Skipping phase 1 and measuring directly can be both cheaper and more precise when the screener costs too much or the stratifier predicts too little, and the print output says so when that happens.

### Response rates

Both phases can lose sample to nonresponse, and the two enter separately. `resp_rate` is the phase-1 response or successful-classification rate, and a `resp_rate` column is the phase-2 completion rate in each stratum. Response divides its own component, exactly as the design effects do, so a stratum that responds poorly is subsampled differently rather than the whole design being inflated by one factor.

```{r twophase-resp}
screened <- transform(frame, resp_rate = c(0.9, 0.7, 0.85, 0.6))

n_twophase(screened, phase1_cost = 1, budget = 50000, resp_rate = 0.8)
```

Everything here counts units **issued**, so a cost quoted per completed interview has to be converted first: `c_contact + resp_rate * c_complete`. Issued and expected-responding counts are reported separately, because they are different quantities and neither is an effective sample size. There is no single effective size for a two-phase design, since the two variance components carry different design effects.

Phase 2 can only draw from what phase 1 managed to classify, so every subsampling fraction is capped at the phase-1 response rate.

One caveat worth stating plainly: dividing by a response rate assumes response is ignorable *within the strata you supplied*. That is a substantive claim about the strata, not a property of the arithmetic, and no sample size removes nonresponse bias. Where the assumption is uncomfortable, modeling the nonresponse as a follow-up phase is the alternative the design itself offers.

### When the phase-1 sample is already fixed

Often it is. The screener has run, or phase 1 is an existing survey, a panel, or a listing operation sized by field capacity rather than by this design. Phone surveys built on a previous face-to-face round are the common case. Supply `n_phase1` and the only decision left is how deep to subsample.

```{r twophase-fixed}
n_twophase(frame, phase1_cost = 1, budget = 50000, n_phase1 = 20000)
```

The relative allocation is unchanged, because it is $S_h\sqrt{d_{2h}/c_h}$ in every mode. What moves is the overall scale, now pinned by the budget left after paying for phase 1 rather than by the variance-cost trade-off. Since the optimizing choice of `n_phase1` is the best member of this family, fixing it can only match or lose to leaving it free, and the difference is what a phase-1 size you did not choose is costing you.

Two things can go wrong, and both give a specific error. The budget may not reach phase 2 at all. Or a target CV may sit below the **phase-1 variance floor**: what that `n_phase1` leaves when every classified unit is carried into phase 2, which is

$$\frac{1}{\mu}\sqrt{\frac{1}{n_a}\left(\frac{d_1A}{r_1} + \sum_h \frac{d_{2h}W_hS_h^2}{r_{2h}r_1}\right) - \frac{d_1A + \sum_h d_{2h}W_hS_h^2}{N}}.$$

The between-stratum term is the part already paid for by the phase-1 size. It is not the whole floor, though, because subsampling stops at the classification rate rather than at 1, so the phase-2 residual survives at $\nu_h = r_1$ and is added to it. The subtracted term is the finite population correction, and it is what makes the floor vanish at a census: with $n_a = N$ and no losses the two halves cancel exactly. It carries the design effects, since those inflate an SRSWOR variance here as everywhere, but not the response rates, because measuring every unit removes all the variance while failing to reach some of them does not.

### Whole units and assurance

`$operational` carries the field design in whole units, and it is what the printed block reports: every count there is a whole unit off the same solution, so the phase sizes, the stratum takes and the responding figures reconcile. Budget mode floors and then buys back units in order of variance reduction per unit cost, staying inside the budget. CV mode rounds up, overshooting precision rather than missing it. `summary()` sets the continuous optimum beside the fielded design and gives the single-phase comparator in full.

Planning at the *expected* respondent count leaves roughly half of all designs short. `assurance` reports the issue for which the required respondents arrive with at least that probability, from the binomial distribution of respondents:

```{r twophase-assurance}
screened <- transform(frame, resp_rate = c(0.9, 0.7, 0.85, 0.6))

n_twophase(screened, phase1_cost = 1, budget = 50000,
           resp_rate = 0.8, assurance = 0.90)
```

If the assured phase-2 issue exceeds the pool phase 1 supplies, that is a signal to enlarge phase 1 rather than to over-issue.

### Nonresponse follow-up

Following up a subsample of nonrespondents is the same allocation problem with two strata. The respondents are already measured, so they cost nothing more and are all kept, which is `unit_cost = 0` with `take_all = TRUE`. Only the nonrespondents are subsampled:

```{r twophase-nrfu}
theta <- 0.5   # phase-1 response rate

nrfu <- data.frame(
  stratum   = c("respondents", "nonrespondents"),
  N         = c(theta, 1 - theta),
  sd        = c(1, 1),
  unit_cost = c(0, 200),
  take_all  = c(TRUE, FALSE)
)

n_twophase(nrfu, phase1_cost = 50, budget = 100000,
           mu = 1, single_cost = 50 / theta)
```

The optimal follow-up fraction reduces to $\sqrt{c_1/(c_2\theta)}$ here, the standard result. Two details matter in this configuration. `single_cost`: a `unit_cost` of 0 means "already measured", not "free", so the baseline for the comparison has to be named explicitly as the cost of one completed interview without follow-up. And `resp_rate` should stay at its default of 1, because here the strata *are* response status, so classification succeeds for every unit and setting a phase-1 rate as well would count the same loss twice.

## Change and panel designs

The families above size one occasion. A survey that runs more than once is sized against a different quantity: the change between two occasions, and, if the same units are followed, how many to recruit so that enough of them are still responding later.

### Sizing a change

`n_change()` and `prec_change()` are the `n_*`/`prec_*` pair for a change in a mean or a proportion. Two independent rounds cost twice a single occasion:

```{r change-independent}
n_change(p = c(0.30, 0.36), moe = 0.02)
```

Measuring the same units twice is cheaper, because the change is a difference and the unit-level correlation cancels part of it. Two arguments describe that, and only their product buys anything: `overlap` is the share of the first occasion's responding sample measured again, and `overlap_cor` the correlation among those shared units. A full panel of uncorrelated measurements saves nothing.

```{r change-overlap}
n_change(p = c(0.30, 0.36), moe = 0.02, overlap = 0.75, overlap_cor = 0.5)
```

The pair round-trips like every other:

```{r change-roundtrip}
res <- n_change(p = c(0.30, 0.36), moe = 0.02, overlap = 0.75, overlap_cor = 0.5)
prec_change(res)$moe
```

`overlap` and `overlap_cor` come from different places, and it is worth keeping them apart. The overlap is a property of the design and is fixed once the rotation is declared. The correlation is a property of the variable and has to come from a previous round of the same survey.

### Overlap from a rotation schedule

`design_overlap()` supplies the design half. Give it the occasions a unit spends in and out of sample over its whole life, as a compact spec naming the spells in order from in sample, and it returns the overlap at every lag the schedule reaches:

```{r overlap-cps}
cps <- design_overlap("4-8-4")
cps[c(1, 12)]
```

Those are the two published CPS figures: 75 percent of the sample is shared between consecutive months, 50 percent between the same month a year apart. **A schedule is a finite life, not a repeating pattern**, and the difference is invisible in the answer. Read `"4-8-4"` as "four in, eight out, repeat" and only the group finishing its stint leaves each month, which gives 87.5 percent. The finite life loses a group at the end of each of its two spells, keeps six of eight, and gives the published 75 percent. A compact spec with an even number of spells could only mean a cycle and is refused.

Each lag is a different design question, so name the one the change spans rather than passing the profile whole:

```{r overlap-lags}
c(consecutive = n_change(p = c(0.30, 0.36), moe = 0.02,
                         overlap = cps[1], overlap_cor = 0.5)$n,
  annual = n_change(p = c(0.30, 0.36), moe = 0.02,
                    overlap = cps[12], overlap_cor = 0.5)$n)
```

Two notations for a schedule are in print, and both are accepted. `"4-8-4"` counts occasions per spell, which is how the CPS pattern is written. `"1-1-0-0-1-1"` carries one flag per occasion, which is how Lynn (2012) writes the same kind of design. They are told apart by the `0`, which no spell can be. A string of all 1s is a valid sentence in both and a different design in each, so it is refused rather than resolved by precedence. `"1-1-1"` is three consecutive occasions as a pattern and one in, one out, one in as spells. Their consecutive overlaps are 2/3 and 0. Write `"3"` or `"1-0-1"`.

`plot()` draws the schedule as the chart these designs are published as, one row per cohort and one column per time period:

```{r overlap-chart, fig.width=7, fig.height=4.5}
plot(design_overlap("1-1-0-0-1-1"))
```

A cohort enters at every period drawn, so the total row climbs through the launch and settles at the sample the overlaps divide by, which is marked. Everything left of that mark is the gradual start a rotating design has: reaching the intended sample takes a full life unless the first period's sample is deliberately split into cohorts of unequal life, which one schedule does not express.

One conversion is the planner's to make. `design_overlap()` counts the units the design *issues* at both occasions, while `overlap` in the sizing and power functions is the share of the first occasion's *respondents* measured again. The two are the same number at full response. Below it they are not, and converting between them needs an assumption about how response persists across occasions, which the package does not make: under response independent between occasions at rate `r`, an issued overlap `f` leaves about `f * r` among respondents, rising towards `f` as respondents become likelier to respond again.

### Recruiting a panel that loses units

A panel loses units at every wave, so the sample that carries the analysis is smaller than the one recruited. `n_panel()` sizes the recruitment. It does not restate the arguments of `n_prop()` or `n_mean()`. It takes one of their results and reads it as both the responding sample to reach and the estimand to report precision for.

The rates below are the UK LFS: 73 percent response at recruitment, then quarterly retention of a surviving cohort.

```{r panel-fixed}
target <- n_prop(p = 0.5, moe = 0.031)
lfs <- n_panel(target, retention = c(0.878, 0.963, 0.936, 0.956),
               resp_rate = 0.728)
lfs
```

The printed block is the answer: what to issue, what it leaves, and the precision at each wave. `summary()` gives what the answer rests on, including the rates assumed, where the life's loss falls, the cv and expected cases at each wave, and for a rotating design the standing sample and the launch.

`resp_rate` and `retention` are separate arguments because a panel's loss is concentrated at recruitment. Here 61 percent of the whole five-wave loss happens at wave 1, and an average rate spread over the waves would under-issue.

```{r panel-waves}
lfs$waves[, c("wave", "q", "loss_share", "n_resp", "moe")]
```

`target_wave` sizes for a precision reached earlier than the last wave, with the later waves still reported.

The same rates can be run as a rotating panel, where an equal cohort enters every occasion and the estimate pools every cohort alive. That returns a different quantity, so it has a different name:

```{r panel-rotating}
rot <- n_panel(target, retention = c(0.878, 0.963, 0.936, 0.956),
               resp_rate = 0.728, design = "rotating")
c(entrants = rot$n_entrants, in_sample = rot$n_in_sample)
```

`n_entrants` is what the design takes on each occasion once it is running. `n_in_sample` is what its five live cohorts hold between them, which is also the cumulative recruitment that reaching a steady state takes. On these rates they differ by a factor of five, and reading one as the other is the mistake the two names exist to prevent.

Every figure so far is an expected value, which leaves about half of all panels short of the target. `assurance` reports the recruitment that clears it with a stated probability instead:

```{r panel-assurance}
n_panel(target, retention = c(0.878, 0.963, 0.936, 0.956),
        resp_rate = 0.728, assurance = 0.90)$n_assured
```

For a fixed panel that inverts one binomial tail. For a rotating one the respondents at an occasion are a sum of binomials at different cumulative probabilities, one per live cohort, which is a Poisson-binomial and is computed exactly rather than approximated.

`prec_panel()` runs the pair in the other direction: given a recruitment already budgeted, it reports the responding sample and the precision left at each wave, and says so when that falls short of what the target needs.

### Bringing a rotating design up

The recruitment above describes the design at a steady state, which it reaches once every stage of the life is represented at one occasion. How it gets there is a design decision. `start = "gradual"` recruits one cohort an occasion, so the sample climbs over a full life. `start = "immediate"` splits the first occasion into equal panels planned for life lengths from the full life down to one occasion. Every panel begins at its first interview, and together they hold the whole sample at once.

```{r panel-launch}
n_panel(target, retention = c(0.878, 0.963, 0.936, 0.956),
        resp_rate = 0.728, design = "rotating",
        start = "immediate")$launch[, c("period", "n_in_sample", "n_resp", "moe")]
```

`design_overlap()` gives the same mature membership-overlap profile under either launch. A gradual launch has higher realized overlap while it fills because no full set of cohorts has yet rotated through. An immediate launch is in membership equilibrium at its first occasion and **not** in response equilibrium. Every unit there is at wave 1, so the occasion holds at least as many respondents as the design ever holds again, and strictly more as soon as any wave retains less than all of the one before. Here that is 1172 against the design's 1001, temporarily more precise, converging down as the interview mix matures. The two coincide only where nothing is lost after recruitment. A gradual launch approaches the same respondent figure from below, at 234. `$launch_waves` decomposes each occasion into the interviews represented in the sample, which is where both launch paths show.

Either way the early occasions rest on a different response composition from the rest of the series, and that, rather than the sample size, is what makes them hard to compare with it.

`design_schedule()` turns that launch into an operational manifest without drawing a sample. The horizon policy is explicit: `"continuing"` records interviews owed after the reporting window, `"truncate_lives"` ends those lives at the window, and `"close_intake"` stops recruiting early enough for every cohort to finish. Rounding is explicit too and is applied to each panel before startup panels are summed.

```{r design-schedule}
rot_launch <- n_panel(
  target,
  retention = c(0.878, 0.963, 0.936, 0.956),
  resp_rate = 0.728,
  design = "rotating",
  start = "immediate"
)
field_plan <- design_schedule(
  rot_launch,
  life = design_overlap("5"),
  horizon = 8,
  horizon_policy = "continuing",
  refreshment = "entrant_register",
  rounding = "ceiling"
)
field_plan
field_plan$components[, c("cohort", "entry_wave", "operational_issue",
                          "panels", "frame_role")]
```

A schedule's length is set by the reporting horizon rather than by the design, so `print()` states the issue profile as the runs it is made of and stays the same size whether the window is six occasions or fifty-two. `summary()` gives the occasion-by-occasion tables. The dense `$schedule` table marks every component-panel combination active or inactive at each in-horizon occasion. `$tail_commitments` contains only the promised interviews after the horizon. An entrant register means the later frames contain new eligible units. If later cohorts instead use whole frame vintages, declare `refreshment = "whole_vintage"` and construct their combined analysis weights outside this planning object.

```{r panel-launch-chart, fig.width=7, fig.height=5}
plot(design_overlap("5"), start = "immediate")
```

### Averaging occasions, and the trade-off that makes

A repeated survey rarely publishes only single occasions and changes between them. An annual average built from quarterly rounds is an estimate in its own right, and it is the one place a rotation costs rather than pays. `prec_pooled()` and `n_pooled()` size it, on the equal-weight mean of the occasion estimates.

```{r pooled-basic}
prec_pooled(var = 100, n = 500, occasions = 4)
prec_pooled(var = 100, n = 500, occasions = 4, overlap = 0.75, cor_decay = 0.8)
```

The second design shares three quarters of each round with the next and is markedly less precise on the annual figure. Overlap induces a positive covariance between occasions. That covariance is subtracted when you difference two occasions and added when you average them. It is positive whenever the overlap exceeds the sampling fraction `n / N`, which always holds here because there is no finite population correction. `?prec_pooled` explains the boundary below it, where occasions share fewer units than chance and the two directions reverse. The correlation is stated per lag because a panel's correlation falls away with distance. Here `cor_decay = 0.8` gives the AR(1) shape, with correlation `0.8^m` at lag `m`.

Three estimands and one design lever, priced together. The overlaps come from the schedule rather than being asserted:

```{r pooled-tradeoff}
horizon <- 8
arms <- function(life) {
  ov <- design_overlap(as.character(life), max_lag = horizon - 1)
  data.frame(
    span        = attr(ov, "life"),
    interviews  = sum(attr(ov, "schedule")),
    overlap_1   = round(as.numeric(ov)[1], 3),
    se_level    = round(prec_mean(var = 100, n = 1000, N = 1e5)$se, 4),
    se_change   = round(prec_change(var = 100, n = 1000, N = 1e5,
                                    overlap = ov[1], overlap_cor = 0.8)$se, 4),
    se_pooled   = round(prec_pooled(var = 100, n = 1000, N = 1e5,
                                    occasions = horizon, overlap = ov,
                                    cor_decay = 0.8)$se, 4)
  )
}
do.call(rbind, lapply(c(2, 4, 6, 8), arms))
```

Reading down the columns: lengthening the life sharpens the change and blunts the annual average, while the single-occasion level does not move at all. The level is a function of that occasion's size alone, so it is a horizontal reference line rather than a third position, and the trade-off is between the other two. A design serving both is sized by taking the larger of `n_change()` and `n_pooled()`, since neither dominates.

Two columns carry the burden because one cannot. `span` is the calendar distance from a unit's first interview to its last and `interviews` is how many times it is actually asked. They coincide for an uninterrupted schedule and separate for a rotation with a rest period, where `"4-8-4"` spans sixteen occasions and asks eight.

```{r pooled-burden}
gapped <- design_overlap("4-8-4")
c(span = attr(gapped, "life"), interviews = sum(attr(gapped, "schedule")))
```

The issued-versus-respondent conversion from earlier applies here too, and `prec_pooled()` enforces it rather than documenting it: a `design_overlap()` result is a statement about issued samples, so it is accepted directly at full response and refused below it, where the number you want is the respondent overlap you expect.

One refusal is worth recognising. A lag's covariance changes sign once its overlap falls below the sampling fraction, and a covariance whose sign varies across lags need not be a covariance at all. `prec_pooled()` checks the assembled matrix, not the variance read off it, because a kernel that fails can still return a positive number:

```{r pooled-guard, error=TRUE}
prec_pooled(var = 100, n = 67000, occasions = 8, N = 1e5,
            overlap = design_overlap("4", max_lag = 7), cor_decay = 0.8)
```

That is a municipality-scale sampling fraction, not an exotic corner, and the message names the fraction, the lag that turns, and whether the rotation would also exhaust its own population before the horizon is out.

## Power analysis

Precision analysis answers "how precisely can we estimate a level?" Power analysis answers a different question: "can we detect a change or difference between two groups or time points?"

### Sample size for a detectable change

A survey measured vaccination coverage at 70%. How many households are
needed in a follow-up to detect a 5 percentage point increase with 80%
power?

```{r power-n}
power_prop(p1 = 0.70, p2 = 0.75, power = 0.80)
```

With a design effect:

```{r power-n-deff}
power_prop(p1 = 0.70, p2 = 0.75, power = 0.80, deff = 2.0)
```

### Three solve modes

`power_prop()` and `power_mean()` each solve for whichever of `n`, `power`, or the effect size is left unspecified:

```{r power-modes}
# Solve for n (default when p2 and power given)
power_prop(p1 = 0.70, p2 = 0.75, power = 0.80, deff = 2.0)

# Solve for power (set power = NULL)
power_prop(p1 = 0.70, p2 = 0.75, n = 1500, power = NULL, deff = 2.0)

# Solve for MDE (omit p2)
power_prop(p1 = 0.70, n = 1500, deff = 2.0)
```

### Panel surveys

The power family takes the same `overlap` and `overlap_cor` as `n_change()`, and `design_overlap()` supplies the first of them from a rotation schedule. If the follow-up resamples a fraction of the original households and the between-round correlation is known, the required sample size drops:

```{r power-panel}
power_prop(p1 = 0.70, p2 = 0.75, power = 0.80, deff = 2.0,
           overlap = 0.50, overlap_cor = 0.6)
```

### Continuous outcomes

The same three solve modes work for means:

```{r power-mean}
# Sample size to detect a difference of 5 (within-group var = 200)
power_mean(200, effect = 5)

# MDE with n = 400 per group
power_mean(200, n = 400)
```

### Difference-in-differences

`power_did()` handles two-group, two-period DiD designs. Specify treated and control group outcomes as `c(baseline, endline)` vectors:

```{r power-did}
# Proportion outcome: treated improves, control stable
power_did(treat = c(0.50, 0.55), control = c(0.50, 0.48),
          outcome = "prop", effect = 0.07)

# Mean outcome with panel overlap
power_did(treat = c(50, 55), control = c(50, 52),
          outcome = "mean", var = 100, effect = 3,
          overlap = 0.5, overlap_cor = 0.6)
```

### Unequal groups and alternative methods

Both `power_prop()` and `power_mean()` support unequal group sizes via `ratio` or explicit `n = c(n1, n2)`:

```{r power-unequal}
# 2:1 allocation ratio
power_prop(p1 = 0.30, p2 = 0.35, ratio = 2)

# Unequal group variances
power_mean(c(80, 120), effect = 5)
```

For rare proportions, arcsine and log-odds transforms are more accurate:

```{r power-methods}
power_prop(p1 = 0.15, p2 = 0.18, alternative = "one.sided",
           method = "arcsine")

power_prop(p1 = 0.15, p2 = 0.18, alternative = "one.sided",
           method = "logodds")
```

### Power curve

`plot()` draws the power-vs-sample-size curve with reference lines at the solved point:

```{r power-plot, fig.width=7, fig.height=4}
pw <- power_prop(p1 = 0.70, p2 = 0.75, power = 0.80, deff = 2.0)
plot(pw)
```

## Sensitivity analysis

How sensitive is a sample size to assumptions about the design effect, response rate, or prevalence? The `predict()` method evaluates any `svyplan` result at new parameter combinations.

### Varying design parameters

```{r predict-n}
x <- n_prop(p = 0.3, moe = 0.05, deff = 1.5)
predict(x, expand.grid(
  deff = c(1.0, 1.5, 2.0, 2.5),
  resp_rate = c(0.8, 0.9, 1.0)
))
```

### Cluster budget scenarios

```{r predict-cluster}
cl <- n_cluster(stage_cost = c(500, 50), icc = 0.05, budget = 100000)
predict(cl, data.frame(budget = c(50000, 100000, 150000, 200000)))
```

### Power curves

```{r predict-power}
pw <- power_prop(p1 = 0.30, p2 = 0.35, n = 500, power = NULL)
predict(pw, data.frame(n = seq(200, 1000, 200)))
```

### Precision by sample size

```{r predict-prec}
pr <- prec_prop(p = 0.3, n = 400)
predict(pr, data.frame(n = c(100, 200, 400, 800, 1600)))
```

## Putting it all together

A typical planning sequence for a national household survey ties several `svyplan` functions together. A `svyplan()` profile captures the shared design assumptions:

1. Set shared design parameters once with `svyplan()`
2. Define indicators and precision targets per domain
3. Compute sample sizes with `n_multi()`
4. Evaluate achieved precision with `prec_multi()`
5. Run sensitivity analysis with `predict()`
6. If frame data is available, estimate variance components with `varcomp()`
7. Optimize the multistage allocation with `n_cluster()` or `n_multi_cluster()`
8. Determine strata boundaries with `strata_bound()`
9. Check that the design has adequate power with `power_prop()` / `power_mean()`

```{r workflow}
# 1: shared design assumptions
design <- svyplan(deff = 2.0, resp_rate = 0.85)

# 2-3: multi-indicator sample size
targets <- data.frame(
  name = c("stunting", "vaccination", "anemia"),
  p    = c(0.25, 0.70, 0.12),
  moe  = c(0.05, 0.05, 0.03),
  deff = c(2.0, 1.5, 2.5)
)
result <- n_multi(targets)
result

# 4: achieved precision for each indicator
prec_multi(result)

# 5: how sensitive is the binding indicator to the response rate?
predict(
  n_prop(p = 0.25, moe = 0.05, plan = design),
  data.frame(resp_rate = c(0.7, 0.8, 0.9, 1.0))
)

# 9: can we detect a 5pp decline in stunting with this sample?
power_prop(
  p1 = 0.25, p2 = 0.20,
  n = as.integer(result), power = NULL,
  plan = design
)
```

Sample-size, cluster, power, and strata-boundary results support `as.integer()` and `as.double()`. For sample-size results, these return the rounded-up and exact values of *n*, respectively, which makes it straightforward to pass results between functions or into other packages.

## Scope

`svyplan` answers one question in several forms: how many units does a planned design need, and what precision does a given design achieve. Some neighbouring topics are deliberately outside that boundary. They are listed here so their absence reads as a decision rather than an oversight.

**Totals.** There is no `n_total()`. A total is a mean rescaled by *N*, so the CV target is identical and a margin of error on the total is `moe / N` on the mean. `?n_mean` has a worked section on sizing for a total.

**Ratio and regression estimation.** Ratio estimators and regression estimators are analysis-stage techniques. Designs are not usually sized around the precision of a ratio, and sizing for a regression coefficient is a general statistical problem rather than a survey one, covered by packages such as `pwr`. The survey-specific part, the variance inflation from a complex design, is already the `deff` argument.

**Two-phase designs beyond stratification and follow-up.** `n_twophase()` allocates a two-phase sample: a large cheap phase 1, then a subsample measured on the expensive variable. Double sampling for stratification and nonresponse follow-up are the same problem in that function, differing only in which strata are marked `take_all`. What is *not* covered is the third member of the family, double sampling for a regression or ratio estimator, where the phase-1 variable is continuous rather than a stratifier. Its variance effect is expressible today as a design effect below 1, since a regression estimator multiplies the variance by $1 - \rho^2$, but the cost trade-off between the phases is not built in.

**More than three stages.** Two- and three-stage designs are supported. Surveys rarely go deeper, and the cost and variance models for a fourth stage would carry more assumptions than they would earn.

**Attrition, and choosing a rotation.** `n_panel()` is arithmetic on declared rates. It says what recruitment yields under a response rate and a retention curve, not where those rates come from. Attrition modelling, attrition weighting, and any adjustment for informative dropout are analysis-stage work and outside the boundary. Two consequences are worth stating. A unit lost at one wave is treated as lost for good, so a panel whose nonrespondents return later does better than planned here, and planning this way over-issues. Also, `design_overlap()` reads a schedule that has already been chosen. It does not solve the inverse problem of finding a schedule that delivers a target overlap while limiting how many times a unit may be interviewed.

**Targeting nonresponse follow-up.** Deciding *how much* follow-up to do is answered by `n_twophase()`. Deciding *which* cases to pursue is not, and is out of scope: prioritizing by response propensity, or adapting effort as fieldwork proceeds, needs paradata from the field that no planning stage has.

**PPS selection inside joint allocation.** `varcomp()` estimates variance components under PPS first-stage selection, but the joint allocator (`n_alloc()` with `measures` and `targets`) uses an anticipated variance model. Feeding it components estimated under PPS is therefore an approximation, and the more unequal the sizes, the rougher it is. A two-stage joint allocation can identify certainty PSUs from a register. It does not support `take_all` in a fixed-take multistage frame because a census of PSUs still leaves within-PSU sampling.

**Domains not identifiable on the frame.** Targets can be set per domain when domain membership is known before fielding, as it is for region or urban/rural. For domains realized only after contact, such as age or sex subgroups, the realized subsample is random and has to be inflated by hand from the expected domain share. The same line runs through `design_df()`: a domain that is a union of whole strata gets an exact degrees-of-freedom count in `$domains`, while one cutting across strata does not, because counting the PSUs that contain its members needs frame information a plan does not carry.

**Converting between listing units and analysis units.** Sizing works in the units the precision target refers to. Converting a size in eligible persons to a number of households to list, given a mean household size and an eligibility rate, is arithmetic the user does explicitly. The MICS/DHS section above and the README both show the calculation.
