---
title: "Stratification and sample allocation"
description: >
  Construct candidate strata, allocate a sample under practical constraints,
  and check the fieldwork allocation.
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Stratification and sample allocation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(svyplan)
```

Stratification divides the population before sampling. An allocation then
decides how many units to select from each stratum. A domain is different. It
is a population for which an estimate is required, and it can contain one or
more strata.

This article covers one-indicator allocation. See [Multiple indicators and
domains](multi-indicator.html) for requirements that combine into one total,
and [Joint allocation](joint-allocation.html) when one allocation must satisfy
several estimands and domains simultaneously.

## Allocate a fixed sample

The frame has one row per stratum. `N` is the population size, `sd` is the
anticipated within-stratum standard deviation, and `mean` supplies the scale
for a CV calculation:

```{r allocation-frame}
frame <- data.frame(
  stratum = c("Urban", "Peri-urban", "Rural"),
  N = c(4000, 3000, 3000),
  sd = c(10, 15, 8),
  mean = c(50, 60, 55)
)

allocation <- n_alloc(frame, n = 600, alloc = "neyman")
allocation
```

Neyman allocation places more sample where $N_hS_h$ is larger. The result
keeps a continuous optimum in `$detail$n` and a whole-unit fieldwork proposal
in `$detail$n_int`:

```{r allocation-detail}
allocation$detail[, c("stratum", "N", "n", "n_int", "weight")]
```

Use the operational column when issuing a sample. Its planned `weight` is the
population-to-sample ratio for the allocation, not a final analysis weight
adjusted for realized selection and response.

## Check an operational allocation

`prec_alloc()` evaluates any proposed set of stratum sample sizes. Passing the
planning result evaluates its continuous design. Supply the integer column to
check the whole-unit fieldwork design:

```{r allocation-check}
prec_alloc(allocation)
prec_alloc(allocation, n = allocation$detail$n_int)
```

The detail table shows how much each stratum contributes to overall variance.
`.share` sums to 1:

```{r precision-detail}
assessment <- prec_alloc(allocation, n = allocation$detail$n_int)
assessment$detail[, c("stratum", "n", ".se", ".cv", ".share")]
```

## Solve for precision or budget

Instead of fixing total `n`, ask for a target CV:

```{r allocation-cv}
n_alloc(frame, cv = 0.04, alloc = "neyman")
```

For budget allocation, add a cost per sampled unit and use the optimal
cost-adjusted rule:

```{r allocation-budget}
cost_frame <- transform(frame, unit_cost = c(1, 1.5, 2.5))
n_alloc(cost_frame, budget = 900, alloc = "optimal")
```

The three common rules answer different questions:

| Rule | Relative allocation | Use |
|:---|:---|:---|
| `"proportional"` | $N_h$ | Equal sampling fractions. |
| `"neyman"` | $N_hS_h$ | Minimize variance with equal unit costs. |
| `"optimal"` | $N_hS_h/\sqrt{c_h}$ | Minimize variance with unequal unit costs. |

`alloc = "power"` implements Bankier allocation [@bankier1988], with weights
proportional to the population CV `sd / abs(mean)` times
`alloc_measure^alloc_q`. The allocation measure defaults to `N * abs(mean)`,
and with that default `alloc_q = 1` gives Neyman allocation. At `alloc_q = 0`
the weights are the population CVs whatever the measure. Means must be
nonzero.
The population CV describes variation among units and differs from the `cv`
precision target, which describes the relative standard error of an estimate.

Supply a positive `alloc_measure` column to use another stratum measure:

```{r allocation-bankier}
bankier_frame <- transform(frame, alloc_measure = N)
n_alloc(bankier_frame, n = 600, alloc = "power", alloc_q = 0.5)
```

With a custom measure, `alloc_q = 1` need not give Neyman allocation.
Stratum-specific response rates and design effects multiply the weights by
`sqrt(deff / resp_rate)`. Bankier allocation does not generally equalize domain
CVs. Use explicit domain targets when comparable precision is required.

## Add operational constraints

Minimum stratum sizes, maximum planned weights, and census strata can be
declared with `min_n_stratum`, `max_weight`, and `take_all`:

```{r allocation-constraints}
constrained_frame <- transform(
  frame,
  unit_cost = c(1, 1.5, 1),
  max_weight = c(25, 20, NA),
  take_all = c(FALSE, FALSE, TRUE)
)

constrained <- n_alloc(
  constrained_frame,
  budget = 3500,
  alloc = "optimal",
  min_n_stratum = 40
)
constrained
```

The constraints change the optimum, so assess the returned operational design
as a unit. Independently rounding continuous stratum values can violate a
budget or precision requirement.

### Fieldwork assumptions by stratum

`deff` and `resp_rate` can be scalar values or frame columns. Stratum-specific
response rates change both the number issued and the best allocation:

```{r stratum-response}
response_frame <- transform(
  frame,
  resp_rate = c(0.92, 0.80, 0.65)
)

response_plan <- n_alloc(response_frame, n = 600, alloc = "neyman")
response_plan$detail[, c("stratum", "n", "n_int")]
```

The response adjustment sets an expected responding count. It does not remove
nonresponse bias. Use [two-phase planning](two-phase.html) when a subsample of
nonrespondents will receive additional follow-up.

## Require precision for domains

When each domain is a union of complete strata, `domains` requests a target
for every domain:

```{r allocation-domains}
domain_frame <- data.frame(
  province = c("North", "North", "South", "South"),
  stratum = c("Urban", "Rural", "Urban", "Rural"),
  N = c(2000, 3000, 1800, 3200),
  sd = c(12, 18, 10, 16),
  mean = c(55, 48, 58, 50)
)

domain_plan <- n_alloc(
  domain_frame,
  domains = "province",
  cv = 0.04,
  alloc = "power",
  alloc_q = 0.3
)
domain_plan
domain_plan$domains
```

These domains are known on the frame. For a domain that appears naturally
within one sample and is not identified on the frame, size the sample from the
domain's expected share, as described in
[Multiple indicators and domains](multi-indicator.html). That calculation
applies to single-stage indicator tables only, because a population share
alone does not describe how domain members spread across clusters.

<span id="strata-boundaries"></span>

## Construct candidate strata

When a continuous auxiliary variable is available on the frame,
`strata_bound()` constructs candidate boundaries. The objective is the
precision of the auxiliary variable itself, so its usefulness for a study
variable depends on how closely the two are related.

```{r boundary-basic}
set.seed(12345)
x <- rlnorm(5000, meanlog = 6, sdlog = 1.2)

boundaries <- strata_bound(
  x,
  n_strata = 4,
  cv = 0.05,
  method = "cumrootf"
)
boundaries
```

Four methods are available:

| Method | Character |
|:---|:---|
| `"cumrootf"` | Fast Dalenius–Hodges cumulative root frequency rule. |
| `"geo"` | Fast geometric progression, useful for skewed values. |
| `"lh"` | LH-inspired coordinate search. |
| `"kozak"` | Kozak random search, exact by enumeration on small problems. |

The last two are heuristics. Their results are candidates rather than
certified global optima.

The Kozak search starts from more points than the LH search, so it costs more
run time, but more starting points do not guarantee a better partition. At a
fixed sample size, compare the CV each method reaches on the frame at hand:

```{r boundary-methods}
set.seed(2024)
kozak_boundaries <- strata_bound(x, n_strata = 4, n = 300, method = "kozak")
kozak_boundaries

method_cv <- c(
  sapply(c("cumrootf", "geo", "lh"), function(method) {
    strata_bound(x, n_strata = 4, n = 300, method = method)$cv
  }),
  kozak = kozak_boundaries$cv
)
round(method_cv, 4)
```

On this variable the LH and Kozak searches reach the same smallest CV, the
cumulative root frequency rule comes close, and the geometric rule falls well
behind.

The search allocates each candidate partition by Neyman allocation unless
`alloc` says otherwise. With `alloc = "power"`, it uses the Bankier rule
[@bankier1988]

$$
n_h \propto \frac{S_h}{|\bar{X}_h|}\left(N_h|\bar{X}_h|\right)^q,
$$

where $S_h$ and $\bar{X}_h$ are the standard deviation and mean of `x` in
stratum $h$, $N_h$ is the stratum population, and $q$ is `alloc_q`. The
population CV $S_h/|\bar{X}_h|$ and the measure $N_h|\bar{X}_h|$ are
recomputed for every candidate partition, so the allocation rule shapes the
cut points:

```{r boundary-bankier}
strata_bound(x, n_strata = 4, n = 300, alloc = "power", alloc_q = 0.5)
```

At `alloc_q = 1` the rule is Neyman allocation. At `alloc_q = 0` the sample is
proportional to the stratum population CVs. The rule does not generally
equalize stratum CVs. Because it divides by each stratum mean, `x` should be a
positive size measure. To use an allocation measure other than
$N_h|\bar{X}_h|$, construct the strata first and then add an `alloc_measure`
column to the frame passed to `n_alloc()`.

`predict()` applies the learned boundaries to new observations:

```{r classify}
predict(
  boundaries,
  newdata = c(100, 500, 1500, 5000),
  labels = paste0("S", 1:4)
)
```

The strata table connects directly to `n_alloc()`:

```{r boundary-handoff}
allocated_boundaries <- n_alloc(boundaries$strata, cv = 0.05)
allocated_boundaries
```

Both functions use the same variance calculation. Design assumptions carry
through both, passed directly or through a `svyplan()` profile:

```{r boundary-plan}
plan <- svyplan(deff = 1.8, resp_rate = 0.85)
planned_boundaries <- strata_bound(x, n_strata = 4, cv = 0.05, plan = plan)
c(
  strata_bound = planned_boundaries$n,
  n_alloc = n_alloc(planned_boundaries$strata, cv = 0.05, plan = plan)$n
)
```

The difference between the two totals is whole-unit rounding, not a change of
model. `strata_bound()` reports the whole-unit allocation in `$strata$n`,
whereas `n_alloc()` reports the continuous optimum.

A scalar `deff` does not move the cut points. At a fixed sample size it
multiplies the variance of every candidate partition by the same factor, so
the same partition remains best and only the reported CV changes:

```{r boundary-deff}
rbind(
  deff_1.0 = strata_bound(x, n_strata = 4, n = 300)$boundaries,
  deff_1.8 = strata_bound(x, n_strata = 4, n = 300, deff = 1.8)$boundaries
)
```

Under a `cv` target, what `deff` changes is the `n` the target requires.

Compare the stratum sampling fractions before deciding whether the proposed
boundaries create workable fieldwork loads:

```{r boundary-plot, fig.cap="Sampling fractions under the candidate four-stratum design. Compare each stratum with the dashed overall fraction to identify where sampling is most intensive.", fig.alt="Bar chart of the sampling fraction in each of four constructed strata. A dashed horizontal line marks the overall sampling fraction."}
plot(boundaries)
```

## More design workflows

### Multi-indicator surveys

[Multiple indicators and domains](multi-indicator.html) explains how indicator
requirements combine, including separate quotas and naturally occurring
domains.

### Joint indicator and domain targets

[Joint allocation](joint-allocation.html) shows how to satisfy several
indicator and overlapping domain requirements with one allocation.

### Stratified two-stage designs

The [stratified-cluster workflow](multistage.html#stratified-two-stage-designs)
estimates components within strata, constructs the `n_alloc()` frame, and
checks the whole-unit design and PSU availability bounds.

### Two-phase designs

[Two-phase planning](two-phase.html) covers screening, phase-specific response,
operational precision, assurance, and nonresponse follow-up.

The allocation formulas and distinctions used here follow standard
survey-sampling treatments [@cochran1977; @valliant2018].

## References
