---
title: "Allocation, domains, and two-phase designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Allocation, domains, and two-phase designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(svyplan)
```
This article covers multi-indicator requirements, stratification, constrained allocation, and two-phase sampling. Start with [Getting started](svyplan.html), or continue with [multistage planning](multistage.html), [repeated surveys](repeated-surveys.html), and [power and sensitivity](power.html).

## Multi-indicator surveys

Household surveys like the Demographic and Health Surveys (DHS) or MICS track many indicators simultaneously. Each indicator has its own expected prevalence, precision target, and design effect. `n_multi()` finds the sample size that satisfies all targets at once.

```{r n-multi}
targets <- data.frame(
  name = c("stunting", "vaccination", "anemia"),
  p    = c(0.25, 0.70, 0.12),
  moe  = c(0.05, 0.05, 0.03),
  deff = c(2.0, 1.5, 2.5)
)

n_multi(targets)
```

The binding indicator (marked with `*`) drives the overall sample size. The other indicators are estimated with better precision than requested.

Proportion rows accept the same four methods as `n_prop()`, either for the whole table through `prop_method` or per row through a `prop_method` column. A `df` column sets each row's interval quantile, so one table can mix an ordinary indicator with a rare one estimated from few clusters:

```{r n-multi-methods}
targets_rare <- data.frame(
  name = c("stunting", "cocaine_use"),
  p    = c(0.25, 0.02),
  moe  = c(0.05, 0.01),
  prop_method = c("wald", "beta"),
  df   = c(NA, 25)
)

n_multi(targets_rare)
```

### Achieved precision

`prec_multi()` computes the achieved precision for each indicator at the chosen *n*:

```{r prec-multi}
plan <- n_multi(targets)
prec_multi(plan)
```

### Domain targets

Surveys often need adequate precision within subpopulations (urban/rural, regions). Add domain columns to the targets data frame and `n_multi()` optimizes each domain separately:

```{r n-multi-domain}
targets_dom <- data.frame(
  name       = rep(c("stunting", "vaccination", "anemia"), each = 2),
  residence  = rep(c("urban", "rural"), 3),
  p          = c(0.18, 0.32, 0.80, 0.60, 0.08, 0.16),
  moe        = c(0.05, 0.05, 0.05, 0.05, 0.03, 0.03),
  deff       = c(1.5, 2.5, 1.2, 1.8, 2.0, 3.0)
)

n_multi(targets_dom, domains = "residence")
```

Domain columns are specified explicitly via the `domains` parameter, and `domain_sampling` decides how the per-domain requirements combine into the single `n` that is reported. The default `"separate"` treats the domains as disjoint quotas drawn independently, the usual case for regional domains in one national survey, and reports their sum: every quota has to be met, so nothing smaller delivers the design. `"natural"` treats them as analytic domains arising at their own rate inside one sample, requires a `share` column, and reports `max(.n / share)`, the size whose expected yield meets every quota.

Neither reading is the largest single domain requirement. That number is reported separately as `n_domain_max`, and it is what `binding` refers to. Use `$domains` to see each domain's own sample size.

### Relative precision in MICS and DHS

MICS is a household survey program run by UNICEF, and DHS is one funded by USAID and implemented by ICF. Both run standardized questionnaires and sample designs across many countries so that indicators on children, women, health and nutrition stay comparable across borders, and both publish the sampling rules their country teams work to. Those rules are where a planner meets the two relative precision targets in this package, and the two programs do not use the same one.

MICS states a requirement as a **relative margin of error**, the margin of error as a fraction of the indicator, which is what `rmoe` takes. DHS states one as a **relative standard error**, the standard error over the indicator, which its sampling manual also calls the coefficient of variation and which is what `cv` takes. They are not interchangeable. Under Wald, `rmoe = q * cv`, where `q` is the normal quantile by default and the t quantile when `df` is supplied. The identity does not hold under the other interval methods. `rmoe` is available on the scalar functions and as an indicator column.

The sampling rules are published. UNICEF's [MICS tools](https://mics.unicef.org/tools) carry the sample size template for the current MICS7 round, and the DHS rules are in ICF International (2012), *Demographic and Health Survey Sampling and Household Listing Manual*, Calverton, Maryland ([PDF](https://dhsprogram.com/pubs/pdf/DHSM4/DHS6_Sampling_Manual_Sept2012_DHSM4.pdf)).

`rmoe` is `moe / p`, so it fixes the same interval `moe` does while scaling with the estimate the way `cv` does. That places the package's four precision quantities on two axes: `se` and `cv` describe the sampling variance, while `moe` and `rmoe` describe the interval drawn around it.

|          | Absolute | Relative |
|:---------|:---------|:---------|
| Sampling | `se`     | `cv`     |
| Interval | `moe`    | `rmoe`   |

The sampling row is free of `alpha` and of the interval method, whereas the interval row depends on both. A rare outcome shows the split at its widest:

```{r rmoe-two-axes}
sapply(c("wald", "wilson", "logodds", "beta"), function(m) {
  r <- prec_prop(p = 0.02, n = 900, deff = 2, df = 25, method = m)
  c(se = r$se, cv = r$cv, moe = r$moe, rmoe = r$rmoe)
})
```

`se` and `cv` do not move across the four columns, but `moe` and `rmoe` do. So `rmoe = q * cv`, where `q` is the interval quantile the design uses (`qnorm(1 - alpha / 2)`, or the `t` quantile on `df` degrees of freedom when `df` is supplied, as here), holds under Wald alone. Under the other three, reading a relative margin of error off the CV that way overstates the precision by 5 to 10 percent in this design, which is why a relative margin of error is a target in its own right rather than a rescaled `cv`.

```{r n-multi-rme}
# MICS-style: 12% relative MOE per domain, domain-specific deff and prevalence
targets_mics <- data.frame(
  name      = rep(c("stunting", "vaccination"), each = 3),
  region    = rep(c("North", "Central", "South"), 2),
  p         = c(0.35, 0.25, 0.40, 0.60, 0.75, 0.55),
  rmoe      = 0.12,
  deff      = c(2.5, 2.0, 3.0, 1.5, 1.2, 1.8),
  resp_rate = c(0.90, 0.85, 0.95, 0.90, 0.85, 0.95)
)

n_multi(targets_mics, domains = "region")
```

The MICS template reports sample size in **households**. svyplan returns the number of **individuals** in the target population. To convert, divide by the expected number of eligible individuals per household: `n_hh = ceiling(n / (pb * hh_size))`, where `pb` is the share of the target population in the total population and `hh_size` is the average household size.

### Minimum sample size per domain

The `min_n_domain` parameter sets a floor for the per-domain sample size (`n_alloc()` uses `min_n_stratum` for its per-stratum floor):

```{r n-multi-min-n}
n_multi(targets_dom, domains = "residence", min_n_domain = 300)
```

### Mixing proportions and means

Targets can mix proportion and mean indicators. Use `var` and `mu` columns for means (leave `p` as `NA`), and `p` for proportions (leave `var`/`mu` as `NA`):

```{r n-multi-mixed}
targets_mixed <- data.frame(
  name = c("vaccination", "expenditure"),
  p    = c(0.70, NA),
  var  = c(NA, 40000),
  mu   = c(NA, 800),
  cv   = c(0.05, 0.10),
  deff = c(1.5, 2.0)
)

n_multi(targets_mixed)
```

## Strata boundaries

When a continuous auxiliary variable is available on the frame,
`strata_bound()` constructs cut points intended to reduce the sample size or
CV for a given number of strata. The iterative choices below are local
heuristics, so their result is a candidate solution rather than a certified
global minimum.

The boundaries are built from the distribution of `x` alone, and the variance
they minimize is that of the total or mean of `x`. Nothing models the study
variable's conditional mean or variance given `x`, so for any `y` other than
`x` the result is a proxy, good in proportion to how closely `y` tracks `x`.
The reported CV is likewise the CV for `x`. Stratifying an expenditure frame
to measure a poverty rate is the case this serves well, whereas stratifying
on a variable unrelated to the indicator is not.

```{r strata-bound}
set.seed(12345)
x <- rlnorm(5000, meanlog = 6, sdlog = 1.2)

strata_bound(x, n_strata = 4, n = 300, method = "cumrootf")
```

Four methods are available:

| Method       | Algorithm                                 | Speed    | Best for       |
|:-------------|:------------------------------------------|:---------|:---------------|
| `"cumrootf"` | Dalenius-Hodges cumulative root frequency | Fast     | General use    |
| `"geo"`      | Geometric progression                     | Fast     | Skewed data    |
| `"lh"`       | LH-inspired coordinate optimization       | Moderate | Fast local search |
| `"kozak"`    | Kozak-inspired random-restart local search| Slow     | More starting points |

The `"kozak"` heuristic explores multiple starts and can find a better local
solution on highly skewed distributions. Neither iterative method guarantees a
global optimum or claims to reproduce the published algorithm exactly:

```{r strata-kozak}
strata_bound(x, n_strata = 4, cv = 0.05, method = "kozak")
```

### Allocation methods

Allocation within strata is controlled by the `alloc` parameter:

- `"proportional"`: $n_h \propto N_h$
- `"neyman"` (default): $n_h \propto N_h \cdot S_h$ (minimizes national CV)
- `"optimal"`: $n_h \propto N_h \cdot S_h / \sqrt{c_h}$ (accounts for costs)
- `"power"`: Bankier (1988) compromise, $n_h \propto S_h \cdot N_h^{q}$,
  where the exponent `alloc_q` $\in [0, 1]$ controls the trade-off between
  national precision (`alloc_q` $= 1$, Neyman) and equal subnational CVs
  (`alloc_q` $= 0$)

```{r strata-power-alloc}
# Bankier power allocation (compromise, alloc_q = 0.5)
strata_bound(x, n_strata = 4, n = 300, alloc = "power", alloc_q = 0.5)
```

### Classifying new observations

`predict()` applies the learned boundaries to new data, returning a
factor:

```{r strata-predict}
sb <- strata_bound(x, n_strata = 4, n = 200, method = "cumrootf")
new_x <- c(100, 500, 1500, 5000)
predict(sb, newdata = new_x, labels = paste0("S", 1:4))
```

### Visualizing allocation

`plot()` shows the sampling fraction per stratum. Under Neyman allocation, high-variance strata are sampled more intensively. The dashed line marks the overall fraction:

```{r strata-plot, fig.width=7, fig.height=4}
plot(sb)
```

### Handing boundaries to the allocator

`$strata` carries the `N`, `sd`, and `mean` columns `n_alloc()` needs, so the
two halves of a stratified design connect directly:

```{r strata-handoff}
sb_cv <- strata_bound(x, n_strata = 4, cv = 0.05, method = "lh")
n_alloc(sb_cv$strata, cv = 0.05)
```

Both functions evaluate the same variance, so `cv` means the same thing to
each and `n` is a fielded sample in both. Design assumptions carry across
unchanged, whether passed directly or through a `svyplan()` profile:

```{r strata-deff}
plan <- svyplan(deff = 1.8, resp_rate = 0.85)
sb_deff <- strata_bound(x, n_strata = 4, cv = 0.05, plan = plan)
sb_deff$n
n_alloc(sb_deff$strata, cv = 0.05, plan = plan)$n
```

The remaining gap between the two totals is integer rounding, not a
difference in model: `strata_bound()` reports the whole-unit allocation in
`$strata$n`, while `n_alloc()` reports the continuous optimum.

A scalar `deff` does not move the cut points. It scales the variance of every
candidate boundary set equally, so what it changes is the `n` a target
requires, not where the strata divide.

## Stratified allocation

Given a sampling frame with stratum population sizes (`N`) and stratum standard deviations (`sd`), `n_alloc()` distributes a total sample across strata. This is the second half of a stratified design: `strata_bound()` finds the cut points, and `n_alloc()` allocates within strata.

```{r n-alloc-basic}
frame <- data.frame(
  N = c(4000, 3000, 3000),
  sd = c(10, 15, 8),
  mean = c(50, 60, 55)
)

# Fixed total n with Neyman allocation
n_alloc(frame, n = 600, alloc = "neyman")
```

Three solve modes are available: fixed total `n`, target `cv`, or `budget` constraint (when `unit_cost` is provided in the frame).

Constraints and domain-level targets are often the more practical use cases:

```{r n-alloc-constraints}
frame_constraints <- transform(
  frame,
  unit_cost = c(1, 1.5, 1),
  max_weight = c(25, 20, NA),
  take_all = c(FALSE, FALSE, TRUE)
)

n_alloc(frame_constraints, budget = 3500, alloc = "optimal", min_n_stratum = 40)
```

```{r n-alloc-domains}
frame_domains <- data.frame(
  province = c("North", "North", "South", "South"),
  stratum = c("Urban", "Rural", "Urban", "Rural"),
  N = c(2000, 3000, 1800, 3200),
  sd = c(12, 18, 10, 16),
  mean = c(55, 48, 58, 50)
)

# Minimum total n such that each province meets the CV target
n_alloc(frame_domains, domains = "province",
        cv = 0.04, alloc = "power", alloc_q = 0.3)
```

`prec_alloc()` computes the achieved precision for any allocation:

```{r prec-alloc}
prec_alloc(frame, n = c(200, 250, 150))
```

`$detail` carries the same numbers per stratum, so you can see where the
precision is actually being bought. `.share` is the stratum's contribution to
the design variance of the overall mean and sums to 1:

```{r prec-alloc-detail}
prec_alloc(frame, n = c(200, 250, 150))$detail[, c("stratum", "n", ".se", ".cv", ".share")]
```

With `domains`, `$domains` reports the same table `n_alloc()` returns, so a
design and its assessment compare row for row.

### Fieldwork that differs across strata

`deff` and `resp_rate` take one value per stratum, either as vectors or as
frame columns, for designs whose fieldwork is not uniform. A response rate
that varies changes where the sample should go, not just how much of it comes
back:

```{r n-alloc-per-stratum}
uneven <- data.frame(
  stratum   = c("Urban", "Peri-urban", "Rural"),
  N         = c(4000, 3000, 3000),
  sd        = c(10, 15, 8),
  mean      = c(50, 60, 55),
  resp_rate = c(0.92, 0.80, 0.65)
)
n_alloc(uneven, n = 600, alloc = "neyman")$detail[, c("stratum", "n", "n_int")]
```

A value shared by every stratum is a constant factor that cancels out of a
proportional weighting, so scalar `deff` and `resp_rate` behave exactly as
before.

### Joint indicator and domain targets

When one allocation must satisfy several estimands, use an atomic frame plus
long `measures` and `targets` tables. Atomic means that every frame row has a
single value for every domain classification in use. The example below crosses
region and residence, allowing those two domain systems to overlap without
requiring separate public functions or an `alloc` selector.

```{r n-alloc-joint-data}
joint_frame <- data.frame(
  stratum = c("NU", "NR", "SU", "SR"),
  region = c("North", "North", "South", "South"),
  residence = c("Urban", "Rural", "Urban", "Rural"),
  N = c(1000, 2000, 1500, 1000),
  unit_cost = c(1, 1.2, 1.5, 1)
)

joint_measures <- data.frame(
  stratum = rep(joint_frame$stratum, 2),
  name = rep(c("vaccination", "income"), each = 4),
  p = c(0.50, 0.40, 0.60, 0.30, rep(NA, 4)),
  mean = c(rep(NA, 4), 50, 55, 60, 45),
  sd = c(rep(NA, 4), 10, 12, 15, 9)
)

joint_targets <- data.frame(
  name = c("vaccination", "vaccination", "income"),
  domain = c(".overall", "region", "residence"),
  level = c(NA, "North", "Urban"),
  cv = c(0.05, 0.08, NA),
  moe = c(NA, NA, 2)
)
```

```{r n-alloc-joint-fit}
joint_fit <- n_alloc(
  joint_frame,
  measures = joint_measures,
  targets = joint_targets,
  min_n_stratum = 2
)
joint_fit
joint_fit$constraints
```

The optimizer minimizes variable cost subject to all three requirements.
`joint_fit$detail$n` is the continuous optimum and
`joint_fit$detail$n_int` is the deterministic feasible operational
recommendation. The latter is locally cleaned but is not claimed to be a
globally optimal integer solution. Both precision tables are retained:

```{r n-alloc-joint-assess}
prec_alloc(joint_fit)
prec_alloc(joint_fit, n = joint_fit$detail$n_int)
```

A modified named allocation can be assessed in any order because names are
matched exactly to `frame$stratum`.

The precision a joint allocation plans for can be checked against the
precision it delivers, by executing the design repeatedly against a known
population and comparing empirical CVs with `$constraints$.achieved`. Two
things are worth fixing before reading such a check. Compare against
`$constraints$.achieved` rather than against the target, since integerization
leaves the realized design a little better than what was asked for, and set
the acceptance band in advance from the number of replicates, since an
empirical CV from a few hundred draws carries its own error. The recipe needs
a sampling engine to execute the design, so it belongs with one rather than
here.

#### Fixed-take multistage joint allocation

The same public interface supports two- and three-stage designs when every
later-stage take is fixed. The frame owns stage populations, takes, and costs.
The measures table owns indicator-specific stage homogeneity parameters. Only
`n_psu` is optimized. Public `n`, planning bounds, and precision assessments
remain in ultimate-unit units.

This three-stage example samples PSUs, then a fixed number of SSUs per PSU,
then a fixed number of ultimate units per SSU:

```{r n-alloc-joint-three-stage}
joint_frame_3stage <- within(joint_frame, {
  unit_cost <- NULL
  N_psu <- c(100, 160, 120, 90)
  N_ssu <- c(600, 1200, 900, 500)
  n_per_psu <- c(8, 10, 12, 7)
  n_per_ssu <- c(4, 5, 3, 4)
  cost_psu <- c(300, 400, 450, 350)
  cost_ssu <- c(25, 30, 35, 28)
  cost_tsu <- c(5, 6, 7, 5)
})

joint_measures_3stage <- transform(
  joint_measures,
  icc_psu = rep(c(0.03, 0.05, 0.08, 0.04), 2),
  icc_ssu = rep(c(0.10, 0.08, 0.12, 0.06), 2),
  var_ratio_psu = 1,
  var_ratio_ssu = 1
)

joint_fit_3stage <- n_alloc(
  joint_frame_3stage,
  measures = joint_measures_3stage,
  targets = joint_targets,
  min_n_stratum = 20
)
joint_fit_3stage$detail[, c(
  "stratum", "n_psu_int", "n_per_psu", "n_per_ssu", "n_int"
)]
prec_alloc(joint_fit_3stage, n = joint_fit_3stage$detail$n_int)
```

The exact operational identity is
`n_int = n_psu_int * n_per_psu * n_per_ssu`. The standard multistage forms
use `N_psu`, while three-stage frames also require the stratum's total `N_ssu`.
Fixed takes must be whole numbers. `take_all` is unavailable because taking
every PSU alone does not establish a census of ultimate units.

#### Certainty PSUs from a register

For a two-stage joint allocation, a PSU register can replace `N_psu`. It has
one row per PSU with `stratum` and `N`, and its totals must match `frame$N`.
The allocation identifies PSUs whose implied PPS inclusion probability reaches
one. A `certainty` column can add PSUs to that part of the design.

```{r n-alloc-joint-certainty}
certainty_frame <- joint_frame_3stage[, setdiff(names(joint_frame_3stage),
                                                  c("N_ssu", "n_per_ssu", "cost_tsu"))]
certainty_frame$N_psu <- NULL

psu_register <- do.call(rbind, lapply(seq_len(nrow(certainty_frame)), function(i) {
  N <- certainty_frame$N[i]
  size <- rep(10, 20)
  size[1] <- floor(N / 3)
  size[2] <- N - sum(size[-2])
  data.frame(stratum = certainty_frame$stratum[i], N = size)
}))

certainty_fit <- n_alloc(
  certainty_frame,
  measures = joint_measures_3stage[, setdiff(names(joint_measures_3stage),
                                              c("icc_ssu", "var_ratio_ssu"))],
  targets = joint_targets,
  psu = psu_register,
  min_n_stratum = 20
)
certainty_fit$detail[, c("stratum", "n_psu_certain", "n_psu_draw", "n_int")]
```

`$psu` carries the classification, threshold, and source for every PSU. The
operational allocation uses whole takes in certainty PSUs and whole PSUs in
the remainder. `prec_alloc(certainty_fit)` reproduces its precision exactly.

### Stratified two-stage designs

Most household surveys stratify first and cluster within each stratum,
sampling enumeration areas and then households. Adding a `icc_psu`
column to the frame switches `n_alloc()` to this design. The per-stratum
homogeneity can come straight from `varcomp()` with `strata`, whose
output columns are named to match the frame:

```{r n-alloc-cluster-varcomp}
set.seed(3)
listing <- data.frame(
  region = rep(c("North", "South"), each = 600),
  ea = rep(1:60, each = 20),
  income = rnorm(1200, rep(c(50, 70), each = 600), 15) +
    rep(rnorm(60, 0, 6), each = 20)
)
vc_region <- varcomp(income ~ ea, data = listing, strata = ~region)
vc_region
```

```{r n-alloc-cluster}
frame_2stage <- merge(
  data.frame(stratum = c("North", "South"), N = c(40000, 60000)),
  as.data.frame(vc_region)[, c("stratum", "sd", "mean", "icc_psu", "var_ratio_psu")],
  by = "stratum"
)
frame_2stage$cost_psu <- c(400, 550)
frame_2stage$cost_ssu <- c(45, 60)

res_2stage <- n_alloc(frame_2stage, cv = 0.02)
res_2stage
res_2stage$detail[, c("stratum", "n_int", "n_per_psu", "n_psu_int")]
```

Each stratum gets its cost-optimal cluster take (fix it instead with a
`n_per_psu` column), and the PSU counts fall out of the element
allocation. Budget mode, `min_n_stratum`, and `max_weight` work the same
way here. `take_all` does not apply: taking every PSU leaves the
within-PSU take in force, so it does not enumerate a stratum, and it is
refused rather than half-honoured.

An `N_psu` column caps each stratum at the PSUs it actually has, both in
the continuous allocation and in the whole-unit design. Without it the
allocation can ask for more clusters than a stratum contains:

```{r n-alloc-cluster-npsu}
frame_bounded <- frame_2stage
frame_bounded$N_psu <- c(40, 900)

res_bounded <- n_alloc(frame_bounded, n = 3000)
res_bounded$detail[, c("stratum", "n_psu", "n_psu_int", "N_psu",
                       ".psu_frac", ".bound_source")]
```

North holds only 40 enumeration areas, so its allocation stops there and
`.bound_source` names the constraint that stopped it. Asking for precision
the remaining PSUs cannot deliver is an error rather than a silent
overshoot, and it says the ceiling came from PSU availability under a
with-replacement first stage rather than from the population.

`N_psu` is a feasibility constraint only. It does not switch on a
first-stage finite population correction, so precision still uses a
with-replacement first stage and stays conservative when a design takes an
appreciable share of the available PSUs. `print()` says so when that
happens, and `.psu_frac` reports the share.

## Two-phase designs

In a two-phase (or double) sample, phase 1 draws a large sample and measures a variable the frame does not carry, and phase 2 subsamples those units and measures the variable of interest on them. What phase 1 measures decides which design it is, a stratification variable giving double sampling for stratification and response status giving nonresponse follow-up. `n_twophase()` allocates both phases at once, either to reach a target CV or to minimize it under a fixed budget.

The frame is one row per phase-2 stratum, similarly to what is used in `n_alloc()` :

```{r twophase-strat}
frame <- data.frame(
  stratum   = c("A", "B", "C", "D"),
  N         = c(3500, 2500, 2500, 1500),
  sd        = c(12, 25, 8, 40),
  mean      = c(40, 70, 35, 90),
  unit_cost = c(2, 5, 1, 9)
)

n_twophase(frame, phase1_cost = 1, budget = 50000)
```

The `nu` column is the fraction of each phase-1 stratum carried forward. Its shape is Neyman-like, proportional to $S_h/\sqrt{c_h}$, but the overall scale is set by the *between*-stratum variance: the weaker the stratification, the closer every fraction moves to 1, meaning keep everything phase 1 found.

### Two design effects, not one

Neither phase has to be a simple random sample, and the two carry *separate* design effects because they apply to different things. `phase1_deff` inflates the between-stratum component, the part phase 1 is responsible for. A `deff` column inflates each stratum's within-stratum residual, the part phase 2 has to measure. `single_deff` describes the comparator, which need not be fielded like phase 2.

```{r twophase-deff}
clustered <- transform(frame, deff = c(1.2, 2.5, 0.8, 1.7))

n_twophase(clustered, phase1_cost = 1, budget = 50000,
           phase1_deff = 2.2, single_deff = 2.9)
```

Inflating the combined variance by one design effect instead would be a different and wrong model. The two also pull against each other, and the trap runs in the direction that looks attractive: a stratifier built purely from between-cluster structure shrinks the residual, lowering the stratum `deff`, but it makes the fitted values nearly constant within clusters and so drives `phase1_deff` up toward the cluster size. Under a clustered phase 1 that first term can dominate, and the design loses to a plain sample even though the residual design effect looks favorable.

Every result carries the single-phase comparison, because two-phase sampling is not always an improvement. Skipping phase 1 and measuring directly can be both cheaper and more precise when the screener costs too much or the stratifier predicts too little, and the print output says so when that happens.

### Response rates

Both phases can lose sample to nonresponse, and the two enter separately. `resp_rate` is the phase-1 response or successful-classification rate, and a `resp_rate` column is the phase-2 completion rate in each stratum. Response divides its own component, exactly as the design effects do, so a stratum that responds poorly is subsampled differently rather than the whole design being inflated by one factor.

```{r twophase-resp}
screened <- transform(frame, resp_rate = c(0.9, 0.7, 0.85, 0.6))

n_twophase(screened, phase1_cost = 1, budget = 50000, resp_rate = 0.8)
```

Everything here counts units **issued**, so a cost quoted per completed interview has to be converted first: `c_contact + resp_rate * c_complete`. Issued and expected-responding counts are reported separately, because they are different quantities and neither is an effective sample size. There is no single effective size for a two-phase design, since the two variance components carry different design effects.

Phase 2 can only draw from what phase 1 managed to classify, so every subsampling fraction is capped at the phase-1 response rate.

One caveat worth stating plainly: dividing by a response rate assumes response is ignorable *within the strata you supplied*. That is a substantive claim about the strata, not a property of the arithmetic, and no sample size removes nonresponse bias. Where the assumption is uncomfortable, modeling the nonresponse as a follow-up phase is the alternative the design itself offers.

### When the phase-1 sample is already fixed

Often it is. The screener has run, or phase 1 is an existing survey, a panel, or a listing operation sized by field capacity rather than by this design. Phone surveys built on a previous face-to-face round are the common case. Supply `n_phase1` and the only decision left is how deep to subsample.

```{r twophase-fixed}
n_twophase(frame, phase1_cost = 1, budget = 50000, n_phase1 = 20000)
```

The relative allocation is unchanged, because it is $S_h\sqrt{d_{2h}/c_h}$ in every mode. What moves is the overall scale, now pinned by the budget left after paying for phase 1 rather than by the variance-cost trade-off. Since the optimizing choice of `n_phase1` is the best member of this family, fixing it can only match or lose to leaving it free, and the difference is what a phase-1 size you did not choose is costing you.

Two things can go wrong, and both give a specific error. The budget may not reach phase 2 at all. Or a target CV may sit below the **phase-1 variance floor**: what that `n_phase1` leaves when every classified unit is carried into phase 2, which is

$$\frac{1}{\mu}\sqrt{\frac{1}{n_a}\left(\frac{d_1A}{r_1} + \sum_h \frac{d_{2h}W_hS_h^2}{r_{2h}r_1}\right) - \frac{d_1A + \sum_h d_{2h}W_hS_h^2}{N}}.$$

The between-stratum term is the part already paid for by the phase-1 size. It is not the whole floor, though, because subsampling stops at the classification rate rather than at 1, so the phase-2 residual survives at $\nu_h = r_1$ and is added to it. The subtracted term is the finite population correction, and it is what makes the floor vanish at a census: with $n_a = N$ and no losses the two halves cancel exactly. It carries the design effects, since those inflate an SRSWOR variance here as everywhere, but not the response rates, because measuring every unit removes all the variance while failing to reach some of them does not.

### Whole units and assurance

`$operational` carries the field design in whole units, and it is what the printed block reports: every count there is a whole unit off the same solution, so the phase sizes, the stratum takes and the responding figures reconcile. Budget mode floors and then buys back units in order of variance reduction per unit cost, staying inside the budget. CV mode rounds up, overshooting precision rather than missing it. `summary()` sets the continuous optimum beside the fielded design and gives the single-phase comparator in full.

Planning at the *expected* respondent count leaves roughly half of all designs short. `assurance` reports the issue for which the required respondents arrive with at least that probability, from the binomial distribution of respondents:

```{r twophase-assurance}
screened <- transform(frame, resp_rate = c(0.9, 0.7, 0.85, 0.6))

n_twophase(screened, phase1_cost = 1, budget = 50000,
           resp_rate = 0.8, assurance = 0.90)
```

If the assured phase-2 issue exceeds the pool phase 1 supplies, that is a signal to enlarge phase 1 rather than to over-issue.

### Nonresponse follow-up

Following up a subsample of nonrespondents is the same allocation problem with two strata. The respondents are already measured, so they cost nothing more and are all kept, which is `unit_cost = 0` with `take_all = TRUE`. Only the nonrespondents are subsampled:

```{r twophase-nrfu}
theta <- 0.5   # phase-1 response rate

nrfu <- data.frame(
  stratum   = c("respondents", "nonrespondents"),
  N         = c(theta, 1 - theta),
  sd        = c(1, 1),
  unit_cost = c(0, 200),
  take_all  = c(TRUE, FALSE)
)

n_twophase(nrfu, phase1_cost = 50, budget = 100000,
           mu = 1, single_cost = 50 / theta)
```

The optimal follow-up fraction reduces to $\sqrt{c_1/(c_2\theta)}$ here, the standard result. Two details matter in this configuration. `single_cost`: a `unit_cost` of 0 means "already measured", not "free", so the baseline for the comparison has to be named explicitly as the cost of one completed interview without follow-up. And `resp_rate` should stay at its default of 1, because here the strata *are* response status, so classification succeeds for every unit and setting a phase-1 rate as well would count the same loss twice.
