---
title: "Joint allocation across indicators and domains"
description: >
  Build frame, measure, and target tables for one allocation that must satisfy
  several precision requirements.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Joint allocation across indicators and domains}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(svyplan)
```

A joint allocation assigns one sample across strata while satisfying several
indicator and domain requirements. Use this interface when separate
single-indicator allocations would compete for the same fieldwork sample.

The simpler [allocation guide](allocation.html) introduces strata, costs, and
operational rounding. [Multiple indicators and domains](multi-indicator.html)
shows how several requirements determine a total before allocation.

## Three input tables

The joint interface separates population structure, anticipated measurements,
and publication requirements:

| Table | One row represents | Main contents |
|:---|:---|:---|
| `frame` | An atomic stratum | Population size, domain labels, costs, and design bounds. |
| `measures` | One indicator in one stratum | Anticipated proportion or mean and standard deviation. |
| `targets` | One required estimate | Indicator, domain and level, and precision target. |

For element-level designs, a proportion in `measures$p` uses the
finite-population variance `N * p * (1 - p) / (N - 1)` for its stratum, with
`N > 1`. This matches the convention in `prec_prop()`. Explicit variances and
standard deviations are used as supplied. The multistage designs introduced
below retain the working proportion variance `p * (1 - p)`.

An atomic stratum has one value for every domain classification. Crossing
region with residence therefore produces four frame rows:

```{r joint-data}
joint_frame <- data.frame(
  stratum = c("NU", "NR", "SU", "SR"),
  region = c("North", "North", "South", "South"),
  residence = c("Urban", "Rural", "Urban", "Rural"),
  N = c(1000, 2000, 1500, 1000),
  unit_cost = c(1, 1.2, 1.5, 1)
)

joint_measures <- data.frame(
  stratum = rep(joint_frame$stratum, 2),
  name = rep(c("vaccination", "income"), each = 4),
  p = c(0.50, 0.40, 0.60, 0.30, rep(NA, 4)),
  mean = c(rep(NA, 4), 50, 55, 60, 45),
  sd = c(rep(NA, 4), 10, 12, 15, 9)
)

joint_targets <- data.frame(
  name = c("vaccination", "vaccination", "income"),
  domain = c(".overall", "region", "residence"),
  level = c(NA, "North", "Urban"),
  cv = c(0.05, 0.08, NA),
  moe = c(NA, NA, 2)
)
```

The targets request an overall vaccination CV, a separate CV for vaccination
in the North, and a margin of error for urban income. The optimizer minimizes
variable cost while meeting all three:

```{r joint-fit}
joint_fit <- n_alloc(
  joint_frame,
  measures = joint_measures,
  targets = joint_targets,
  min_n_stratum = 2
)
joint_fit
joint_fit$constraints
```

`$constraints` shows the target and achieved precision for every requirement.
Requirements closest to their limits determine the allocation. The remaining
requirements can be better than requested because they share the same sample.

## Continuous and fieldwork allocations

The continuous optimum and the whole-unit recommendation are both retained:

```{r joint-detail}
joint_fit$detail[, c("stratum", "n", "n_int", "weight")]
```

The integer allocation is constructed to remain feasible. Check both designs
directly:

```{r joint-check}
prec_alloc(joint_fit)
prec_alloc(joint_fit, n = joint_fit$detail$n_int)
```

A modified named allocation can be passed in any row order because names are
matched to `frame$stratum`. When comparing a simulation with the plan, compare
the empirical precision with `$constraints$.achieved`, not only with the
requested target, because integerization can make the fieldwork design
slightly more precise.

## Fixed-take multistage allocation

The same interface supports two-stage designs when the number of ultimate
units sampled per primary sampling unit (PSU) is fixed. The frame holds PSU
populations, takes, and stage costs. The measures table holds the
indicator-specific PSU homogeneity. Only the number of PSUs is optimized.

```{r multistage-data}
cluster_frame <- within(joint_frame, {
  unit_cost <- NULL
  N_psu <- c(100, 160, 120, 90)
  n_per_psu <- c(8, 10, 12, 7)
  cost_psu <- c(300, 400, 450, 350)
  cost_ssu <- c(25, 30, 35, 28)
})

cluster_measures <- transform(
  joint_measures,
  icc_psu = rep(c(0.03, 0.05, 0.08, 0.04), 2),
  var_ratio_psu = 1
)
```

```{r multistage-fit}
cluster_fit <- n_alloc(
  cluster_frame,
  measures = cluster_measures,
  targets = joint_targets,
  min_n_stratum = 20
)

cluster_fit$detail[, c(
  "stratum", "n_psu_int", "n_per_psu", "n_int", "weight"
)]
prec_alloc(cluster_fit, n = cluster_fit$detail$n_int)
```

Public `n` remains the ultimate-unit count. The operational identity is
`n_int = n_psu_int * n_per_psu`. A three-stage frame adds `N_ssu`,
`n_per_ssu`, `cost_tsu`, and the corresponding indicator-specific SSU
homogeneity. `take_all` is unavailable in fixed-take multistage mode because a
census of PSUs does not imply a census of ultimate units.

## Certainty PSUs from a register

For a two-stage design, a PSU register can replace `N_psu`. It has one row per
PSU with its stratum and size, and its stratum totals must match `frame$N`.
`n_alloc()` identifies PSUs whose implied probability-proportional-to-size
inclusion probability reaches 1. A `certainty` column can mark additional
certainty PSUs.

```{r certainty-data}
certainty_frame <- cluster_frame
certainty_frame$N_psu <- NULL

psu_register <- do.call(rbind, lapply(seq_len(nrow(certainty_frame)), function(i) {
  population <- certainty_frame$N[i]
  size <- rep(10, 20)
  size[1] <- floor(population / 3)
  size[2] <- population - sum(size[-2])
  data.frame(stratum = certainty_frame$stratum[i], N = size)
}))
```

The remainder's PSUs are selected with probability proportional to size, so
its `icc_psu` is the size-weighted share of the stratum variance that lies
between PSUs. `varcomp()` returns it when `prob` is each PSU's share of its
stratum's units. Use every PSU of the stratum, since which ones end up
certainty depends on the allocation. Here `census` stands for a census or
pilot with one row per unit:

```{r certainty-icc}
set.seed(2)
psu_register$psu_id <- seq_len(nrow(psu_register))
census <- data.frame(
  stratum = rep(psu_register$stratum, psu_register$N),
  psu_id = rep(psu_register$psu_id, psu_register$N)
)
psu_effect <- rnorm(nrow(psu_register))
census$vaccination <- rbinom(
  nrow(census), 1, plogis(-0.2 + 0.5 * psu_effect[census$psu_id])
)
census$income <- 50 + 3 * psu_effect[census$psu_id] +
  rnorm(nrow(census), 0, 11)
census$share <- ave(census$psu_id, census$psu_id, FUN = length) /
  ave(census$psu_id, census$stratum, FUN = length)

icc_of <- function(name) {
  vc <- varcomp(reformulate("psu_id", name), strata = ~stratum,
                prob = ~share, data = census)
  data.frame(stratum = vc$strata$stratum, name = name,
             icc_psu = vc$strata$icc_psu)
}
certainty_measures <- merge(
  joint_measures, rbind(icc_of("vaccination"), icc_of("income"))
)
certainty_measures
```

Without `prob`, `varcomp()` assumes PSUs of equal chance and measures the
spread of PSU totals, which carries the spread of PSU sizes. On a register
as unequal as this one that is not the homogeneity the model needs:

```{r certainty-icc-srs}
varcomp(income ~ psu_id, strata = ~stratum, data = census)$strata$icc_psu
```

`varcomp()` needs at least two PSUs in a stratum. A stratum with a single PSU
has no variance between PSUs, so leave its `icc_psu` as `NA` and `n_alloc()`
uses 0.

```{r certainty-fit}
certainty_fit <- n_alloc(
  certainty_frame,
  measures = certainty_measures,
  targets = joint_targets,
  psu = psu_register,
  min_n_stratum = 20
)

certainty_fit$detail[, c(
  "stratum", "n_psu_certain", "n_psu_draw", "n_int"
)]
```

`n_psu_certain` counts the certainty PSUs and `n_psu_draw` the remainder PSUs
the operational design draws, so
`n_int = n_certain_int + n_psu_draw * n_per_psu`. `$psu` records each PSU's classification, threshold, and source.
`prec_alloc(certainty_fit)` reproduces the planned precision of that fieldwork
design.

Some planning tools also move small groups of remainder PSUs into the
certainty part, so that each group yields at least two PSUs for variance
estimation. That grouping is a selection decision, and `n_alloc()` does not
make it. To plan with it, form the groups from the PSU sizes and
`$psu$.threshold`, set `certainty = TRUE` in the register for the PSUs to
promote, and solve again. Those PSUs report `.certainty_source` as
`"supplied"`, and the printed plan counts them. Expect the certainty count to
differ from such a tool's until you do.

## Explore controls that remain movable

For a fixed-take multistage result, `predict()` can re-solve the stored problem
over candidate takes:

```{r take-sensitivity}
predict(cluster_fit, data.frame(n_per_psu = c(6, 10, 14, 20)))
```

Changing precision targets or the measures table defines a different planning
problem. Edit those inputs and call `n_alloc()` again. See `?predict.svyplan`
for the supported controls and output columns for each joint-allocation mode.

For a single indicator with independently chosen cluster counts and takes,
continue with [Planning multistage surveys](multistage.html).
