---
title: "Multistage survey planning"
description: >
  Estimate variance components, choose cluster counts and takes, and assess
  design effects and degrees of freedom.
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Multistage survey planning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(svyplan)
```

The package supports two- and three-stage cluster designs. A primary sampling
unit (PSU) is selected at the first stage, a secondary sampling unit (SSU) at
the second, and an ultimate unit is the person, household, or other unit on
which the outcome is measured. See [Getting started](svyplan.html) for the
core sizing and precision workflow and [sample allocation](allocation.html)
for stratified designs.

## Multistage cluster designs

For two-stage or three-stage designs, the question is not just "how many?" but
"how many clusters and how many units per cluster?" The answer depends on the
cost structure and the degree of clustering.

### Variance components from frame data

If frame data with cluster identifiers are available, `varcomp()` estimates
between-cluster and within-cluster variance components:

```{r varcomp}
set.seed(1)
n_clust <- 80
n_hh <- 15
cluster_id <- rep(seq_len(n_clust), each = n_hh)
cluster_mean <- rnorm(n_clust, mean = 50, sd = 10)
expenditure <- cluster_mean[cluster_id] + rnorm(n_clust * n_hh, sd = 20)
frame <- data.frame(cluster = cluster_id, expenditure = expenditure)

vc <- varcomp(expenditure ~ cluster, data = frame)
vc
as.data.frame(vc)
```

For this two-stage design, `icc` is the measure of homogeneity
$\delta = V_b/(V_b + V_w)$ in the sampling variance decomposition described
by @valliant2018. It feeds directly into `n_cluster()` and the cluster-planning
mode of `design_effect()`. The associated design-effect approximation is
$k[1 + (b - 1)\delta]$, where `var_ratio` supplies $k$ and $b$ is the number
of units sampled per PSU.

This measure is related to conventional intraclass correlations and Kish's
rate of homogeneity (*roh*), but the values are not generally identical.
When supplying an ICC from another source, check its definition and the
associated design-effect formula. The component ratio above lies between
0 and 1 for nonnegative components with a positive sum. Values near either
boundary are special cases for the closed-form optimizer, so `n_cluster()`
rejects values numerically too close to either boundary.

### Optimal two-stage allocation

Given per-stage costs and the variance structure, `n_cluster()` finds the
allocation that either minimizes cost for a target CV, or minimizes CV for a
given budget:

```{r n-cluster-2stage}
# Minimize cost to achieve CV = 0.05
n_cluster(stage_cost = c(500, 50), icc = vc, cv = 0.05)

# Maximize precision within a budget
n_cluster(stage_cost = c(500, 50), icc = vc, budget = 100000)
```

Passing `icc = vc` (the varcomp object) automatically extracts `icc`,
`unit_relvar`, and `var_ratio`.

### Three-stage designs

Add a third cost element and a second icc for three-stage designs:

```{r n-cluster-3stage}
n_cluster(
  stage_cost  = c(1000, 200, 20),
  icc = c(0.01, 0.05),
  cv    = 0.05
)
```

### Response rate in cluster designs

A cluster design can lose units at more than one stage, and the two losses are
not interchangeable, so they are named for the stage they act on.
`resp_rate_psu` is the share of selected clusters that can be worked at all, as
when whole PSUs are lost to insecurity, flooding, or an inaccessible road.
Losing a PSU removes its entire contribution, so it deflates every variance
term alike and the first-stage sample is inflated by `1 / resp_rate_psu` to
cover it.

In budget mode, reducing the PSU response rate alone leaves the allocation
unchanged and worsens the achieved CV. Later-stage response rates can change
the optimal takes even with a fixed budget. In CV mode, compensate for lost
PSUs by increasing the first-stage issue:

```{r n-cluster-resp}
n_cluster(stage_cost = c(500, 50), icc = vc, cv = 0.05, resp_rate_psu = 0.90)
```

Specify response separately for each sampled stage:

| Argument | Affected unit | Interpretation |
|:---|:---|:---|
| `resp_rate_psu` | PSU, in either design | Probability that a selected cluster can be worked. |
| `resp_rate_ssu` | SSU, in a three-stage design | Response at the second stage, conditional on the PSU being available. |
| `resp_rate` | Ultimate unit | Completion at the last stage, conditional on the earlier stages being available. |

For a two-stage design, the SSU is the ultimate unit, so its response belongs
in `resp_rate`. Supplying `resp_rate_ssu` for that design returns an error.

The three are not interchangeable. Losing whole PSUs deflates every variance
term alike, whereas losing ultimate units shrinks the realized cluster and so
changes the clustering penalty too. That is why the later-stage rates move the
cost-optimal take while the PSU rate does not:

```{r n-cluster-resp-stages}
# a DHS-shaped design: EAs, households within EAs, one person per household
n_cluster(stage_cost = c(500, 100, 50), icc = c(0.02, 0.05), cv = 0.05,
          resp_rate_psu = 0.95, resp_rate_ssu = 0.90, resp_rate = 0.85)
```

Sizes and costs stay gross. `$n` counts the units to issue and `$cost` pays for
them, while the variance reads what they realize. Lumping the later rates into
`resp_rate_psu` overstates the variance, so the plan meets the target at a
higher cost than needed and with the wrong take, and the excess cost grows with
the between-PSU homogeneity.

### Fixed overhead costs

The standard cost model is purely linear: `C = c1*n_psu +
c2*n_psu*n_per_psu`. Real surveys also have a fixed overhead for training,
infrastructure, or setup. The `fixed_cost` parameter adds this term. In budget
mode, only `budget - fixed_cost` is available for variable costs. This reduces
the number of PSUs while leaving the continuous cost-optimal take per PSU
unchanged. In CV mode, the required allocation stays the same and
`fixed_cost` is added to total cost:

```{r n-cluster-fixed}
# Budget mode: fixed overhead reduces the allocatable budget
n_cluster(stage_cost = c(500, 50), icc = vc, budget = 100000, fixed_cost = 5000)

# CV mode: same allocation, cost increases by fixed_cost
n_cluster(stage_cost = c(500, 50), icc = vc, cv = 0.05, fixed_cost = 5000)
```

The same parameter is available when `n_cluster()` sizes a table of
indicators.

### Evaluating cluster allocations

Compare a candidate fieldwork allocation with the continuous optimum, keeping
the same variance components and costs. Suppose operations can field 100 PSUs
with 10 ultimate units each:

```{r prec-cluster}
alloc <- n_cluster(stage_cost = c(500, 50), icc = vc, budget = 100000)
candidate <- c(n_psu = 100, n_per_psu = 10)
candidate_cost <- candidate[1] * (500 + 50 * candidate[2])

data.frame(
  design = c("Continuous optimum", "100 PSUs with 10 units each"),
  n_psu = c(alloc$n[1], candidate[1]),
  n_per_psu = c(alloc$n[2], candidate[2]),
  cost = c(alloc$cost, candidate_cost),
  cv = c(prec_cluster(alloc)$cv, prec_cluster(n = candidate, icc = vc)$cv)
)
```

The whole-unit candidate spends the same 100,000 budget. Its CV shows the
precision cost of that operational choice relative to the continuous optimum.
This comparison assumes that at least 100 PSUs and 10 units per selected PSU
are available. Frame limits must also be checked before selecting records.

Vary an uncertain cost before committing to that plan:

```{r cluster-cost-sensitivity}
predict(alloc, data.frame(cost_psu = c(400, 500, 600)))
```

The budget stays fixed while the preferred cluster counts and takes change.
A higher cost of opening a PSU favors taking more units in each visited PSU.

Vary the budget to see what each increment buys:

```{r cluster-budget-sensitivity}
predict(alloc, data.frame(budget = c(50000, 100000, 150000, 200000)))
```

The cost-optimal take does not depend on the budget, so every increment goes
to more PSUs. Without a finite population correction, the CV falls with the
square root of the budget, and doubling the budget cuts it by a factor of
about 1.4.

## Stratified two-stage designs

When a survey stratifies first and clusters within each stratum, use
`n_alloc()` to choose stratum-specific cluster counts and takes. Estimate the
variance components within strata from a pilot or previous survey:

```{r stratified-components}
set.seed(3)
listing <- data.frame(
  region = rep(c("North", "South"), each = 600),
  ea = rep(1:60, each = 20),
  income = rnorm(1200, rep(c(50, 70), each = 600), 15) +
    rep(rnorm(60, 0, 6), each = 20)
)
vc_region <- varcomp(income ~ ea, data = listing, strata = ~region)
vc_region
```

Here `ea` identifies enumeration areas, the PSUs. Join the components to a
planning frame with one row per stratum. `N` counts ultimate population units,
while `cost_psu` is paid per selected PSU and `cost_ssu` per ultimate unit:

```{r stratified-allocation}
frame_2stage <- merge(
  data.frame(stratum = c("North", "South"), N = c(40000, 60000)),
  as.data.frame(vc_region)[, c(
    "stratum", "sd", "mean", "icc_psu", "var_ratio_psu"
  )],
  by = "stratum"
)
frame_2stage$cost_psu <- c(400, 550)
frame_2stage$cost_ssu <- c(45, 60)

stratified_plan <- n_alloc(frame_2stage, cv = 0.02)
stratified_plan$detail[, c(
  "stratum", "n_per_psu", "n_int", "n_psu_int", "n_per_psu_int"
)]
stratified_plan$operational[c("n", "cost", "cv")]
```

`n_per_psu` is the continuous cost-optimal take. The field design selects
`n_psu_int` whole PSUs at the whole take `n_per_psu_int`, which gives the
`n_int` ultimate units, and `$operational` reports that design's cost and
precision. The function chooses cost-optimal takes unless they are fixed by a
frame column. `take_all` is unsupported in this mode, because visiting all
PSUs while subsampling within them does not enumerate the stratum.

The take is part of the design, so evaluating the field design, or a
variation of it, needs both stage sizes. With them, `prec_alloc()`
reproduces the operational precision:

```{r stratified-field-precision}
field <- stratified_plan$detail
prec_alloc(
  stratified_plan,
  n = field$n_int,
  n_per_psu = field$n_per_psu_int
)$cv
```

### Bound the number of available PSUs

Suppose the frame contains only 40 PSUs in the North and 900 in the South:

```{r stratified-availability}
frame_bounded <- frame_2stage
frame_bounded$N_psu <- c(40, 900)
bounded_plan <- n_alloc(frame_bounded, n = 3000)
bounded_plan$detail[, c(
  "stratum", "n", "n_int", "n_psu_int", "n_per_psu_int", "N_psu",
  ".bound_source"
)]
bounded_plan$operational[c("n", "cost", "cv")]
```

The North allocation reaches its PSU ceiling, and `.bound_source` names the
constraint. The remaining sample comes from the South. In whole units the
North design stays within its 40 PSUs, and the field total stays within one
take of the requested 3,000.

On its own, `N_psu` is an availability bound. Under the default
`fpc = "unit"` it adds no first-stage finite population correction, so
precision keeps the with-replacement first-stage approximation even when a
stratum selects most of its PSUs. `fpc = "stage"` applies the stage-by-stage
correction, which needs `N_psu`, and `N_ssu` as well at three stages:

```{r stratified-availability-fpc}
c(
  unit = bounded_plan$cv,
  stage = n_alloc(frame_bounded, n = 3000, fpc = "stage")$cv
)
```

An unattainable precision target returns an error rather than requesting
unavailable PSUs.

## Assess the design

### Design effects

`design_effect()` builds the design effect you expect a plan to produce, from
the design features you are choosing. The clustering component uses the same
variance model as `n_cluster()` and `prec_cluster()`, so the two always agree:

```{r design-effect}
design_effect(icc = 0.05, n_per_psu = 20)
```

Components are selected by the arguments you supply and multiply together.
Unequal weighting comes from planned weights or from a planned allocation's
stratum sizes and takes, and the stratification gain (the only component that
can fall below 1) from the stratum standard deviations and means:

```{r design-effect-components}
frame <- data.frame(
  N = c(50000, 120000), n = c(600, 400), sd = c(12, 20), mean = c(55, 48)
)
deff <- design_effect(icc = 0.05, n_per_psu = 20, strata = frame)
deff
```

The result is a plain number wherever one is expected, and `effective_n()` is
its mirror:

```{r design-effect-use}
n_prop(p = 0.3, moe = 0.05, deff = deff)
effective_n(deff, n = 1000)
```

You can also read the features off a plan you already built, including an
`n_alloc()` allocation:

```{r design-effect-plan}
design_effect(n_cluster(stage_cost = c(500, 50), icc = 0.05, cv = 0.05))
```

These are planning quantities. After data collection, compute the design effect
the sample actually achieved with `survey::svymean(..., deff = TRUE)`, which
uses the realized weights, strata, and clusters.

### Planned degrees of freedom

For a design-based variance estimator, degrees of freedom depend on the
number of independently sampled PSUs and the number of strata. In an
unstratified cluster design with $m$ sampled PSUs, the planned count is
$m - 1$:

```{r design-df}
cluster_plan <- n_cluster(
  stage_cost = c(500, 50),
  icc = 0.05,
  budget = 100000
)
design_df(cluster_plan)
```

`design_df()` uses the whole-unit PSU count that will be fielded. Pass its
result to interval calculations that accept `df`, such as the Korn–Graubard
proportion method. In a stratified design, the count is sampled PSUs minus
strata.

This is a planned count. Certainty PSUs do not contribute between-PSU
variance, and a domain that cuts across strata needs the realized number of
PSUs containing domain members. A plan without that frame information cannot
derive those adjustments exactly. See `?design_df` for limits by design class.

### Multi-indicator multistage

For surveys with several indicators in a cluster design, hand `n_cluster()`
a table of indicators carrying `icc_psu` columns, along with stage costs:

```{r n-multi-cluster}
targets_ms <- data.frame(
  name   = c("stunting", "vaccination"),
  p      = c(0.25, 0.70),
  cv     = c(0.10, 0.08),
  icc_psu = c(0.05, 0.02)
)

n_cluster(indicators = targets_ms, stage_cost = c(500, 50), budget = 80000)
```

For several indicators and domains allocated across explicit strata, continue
with [Joint allocation](joint-allocation.html). The variance and cost models
used here follow @valliant2018.

## References
