---
title: "Planning for multiple indicators and domains"
description: >
  Combine indicator requirements, distinguish separate from natural domains,
  and translate analysis units into a fieldwork sample.
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Planning for multiple indicators and domains}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(svyplan)
```

Surveys commonly publish several indicators and estimates for several
subpopulations. This article shows how their requirements combine into one
sample size. See [Precision targets](precision.html) if `moe`, `rmoe`, and `cv`
are new to you.

## Several indicators in one sample

Each row of the input table is one indicator with its anticipated value,
precision target, and design assumptions:

```{r multi-basic}
targets <- data.frame(
  name = c("stunting", "vaccination", "anemia"),
  p = c(0.25, 0.70, 0.12),
  moe = c(0.05, 0.05, 0.03),
  deff = c(2.0, 1.5, 2.5)
)

multi_plan <- n_multi(targets)
multi_plan
```

`n_multi()` calculates each requirement and takes the largest. The binding
indicator, marked with `*` in the output, determines the common sample size.
The other indicators meet or exceed their requested precision.

Assess the precision each indicator receives at that size with
`prec_multi()`:

```{r multi-precision}
prec_multi(multi_plan)
```

Proportions and means can share a table. Use `p` for a proportion and `var`
with `mu` for a mean:

```{r multi-mixed}
mixed_targets <- data.frame(
  name = c("vaccination", "expenditure"),
  p = c(0.70, NA),
  var = c(NA, 40000),
  mu = c(NA, 800),
  cv = c(0.05, 0.10),
  deff = c(1.5, 2.0)
)

n_multi(mixed_targets)
```

An indicator table can also supply `resp_rate`, `df`, `prop_method`, or
`min_cases` by row. See `?n_multi` for the complete column contract.

## Domains sampled as separate quotas

A domain is a population for which the survey needs a separate estimate.
When regions are explicit sampling quotas, their requirements add because
every quota must be fielded:

```{r separate-domains}
regional_targets <- data.frame(
  name = rep(c("stunting", "vaccination"), each = 3),
  region = rep(c("North", "Central", "South"), 2),
  p = c(0.35, 0.25, 0.40, 0.60, 0.75, 0.55),
  moe = 0.05,
  deff = c(2.5, 2.0, 3.0, 1.5, 1.2, 1.8)
)

regional_plan <- n_multi(
  regional_targets,
  domains = "region",
  domain_sampling = "separate"
)
regional_plan
regional_plan$domains
```

Within each region, one indicator binds. The reported overall `n` is the sum
of the regional requirements. `n_domain_max` is only the largest single
regional requirement, not the national total.

Use `min_n_domain` when operations require a minimum quota independently of
the statistical target:

```{r domain-floor}
n_multi(
  regional_targets,
  domains = "region",
  domain_sampling = "separate",
  min_n_domain = 300
)
```

## Domains that occur naturally in one sample

Some domains, such as age or sex groups, are not separate sampling quotas.
They appear within a common sample at an anticipated rate. Set
`domain_sampling = "natural"` and provide that rate in `share`:

```{r natural-domains}
natural_targets <- data.frame(
  name = "coverage",
  group = c("common", "rare"),
  p = 0.50,
  moe = c(0.05, 0.10),
  share = c(0.90, 0.10)
)

natural_plan <- n_multi(
  natural_targets,
  domains = "group",
  domain_sampling = "natural"
)
natural_plan

transform(
  natural_plan$domains,
  overall_requirement = .n / .share
)
```

The common group needs the larger sample *within its domain*, which is
`n_domain_max`. The rare group nevertheless determines the overall sample
because only 10% of the sample is expected to belong to it. In natural-domain
mode, the binding overall requirement is the largest `.n / .share`, not the
largest raw `.n`.

This calculation targets the expected domain yield. It does not guarantee
that a realized random sample will contain the required count. Natural-domain
sizing is unavailable for multistage indicator tables because a population
share alone does not describe how domain members are distributed across
clusters. In that setting, size domains separately or use a design-specific
model based on the frame.

Domains known on the frame can instead be represented as explicit strata or
allocation constraints. Continue with [sample allocation](allocation.html) or
[joint allocation](joint-allocation.html) for those designs.

## Relative precision in MICS and DHS

The distinction between `rmoe` and `cv` matters in major household survey
programs. MICS sample-size tools express some requirements as a relative
margin of error, which maps to `rmoe` [@khan2019mics]. DHS documentation uses
relative standard error, also called the coefficient of variation, which maps
to `cv` [@icf2012]. These quantities are not interchangeable outside the Wald
relationship described in [Precision targets](precision.html).

Here is an illustrative MICS-style table with a 12% relative margin of error:

```{r relative-moe}
relative_targets <- data.frame(
  name = rep(c("stunting", "vaccination"), each = 3),
  region = rep(c("North", "Central", "South"), 2),
  p = c(0.35, 0.25, 0.40, 0.60, 0.75, 0.55),
  rmoe = 0.12,
  deff = c(2.5, 2.0, 3.0, 1.5, 1.2, 1.8),
  resp_rate = c(0.90, 0.85, 0.95, 0.90, 0.85, 0.95)
)

relative_plan <- n_multi(relative_targets, domains = "region")
relative_plan
```

These values illustrate the interface. They are not a country-specific survey
design.

## Convert analysis units to listing units

The unit returned by a calculation is the unit described by its indicator and
design inputs. For this example, assume both indicators refer to the same
eligible population and all eligible people in participating households are
approached. The anticipated design effects must cover the intended household
sampling design. If fieldwork samples households, preserve each region's
requirement when converting to household counts.

Here the indicator table's `resp_rate` describes individual completion
conditional on household participation. The issued person counts therefore
already allow for individual nonresponse. Use the expected number of eligible
people per participating household:

```{r household-conversion}
household_plan <- as.data.frame(relative_plan)
household_plan$people_to_issue <- ceiling(household_plan$.n)
target_population_share <- 0.18
mean_household_size <- 4.5
eligible_per_household <- target_population_share * mean_household_size

household_plan$households_to_complete <- ceiling(
  household_plan$people_to_issue / eligible_per_household
)
household_plan[c("region", "people_to_issue", "households_to_complete")]
```

Allow separately for household nonresponse, which has not been included in
this example's indicator table:

```{r household-response}
household_response <- 0.90
household_plan$households_to_issue <- ceiling(
  household_plan$households_to_complete / household_response
)
household_plan[c("region", "households_to_complete", "households_to_issue")]
```

State whether an eligibility or response adjustment is already present before
converting units. Applying either adjustment twice will overstate the
fieldwork requirement. Eligibility and household size can also vary by region.
These conversions plan expected yields, rather than guaranteeing enough
eligible respondents in every region.

If indicators refer to different eligible populations, convert each
indicator's required person count to households using its own expected yield
and response rates, then take the largest household requirement within each
region. Taking the largest person count first can miss the indicator that
requires the most households. If only some eligible people are selected
within a household, include that subsampling in the expected yield and use
design assumptions appropriate to that selection procedure.

## References
