---
title: "Power for comparisons and changes"
description: >
  Solve for per-group sample size, power, or a minimum detectable effect under
  independent, overlapping, and difference-in-differences designs.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Power for comparisons and changes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(svyplan)
```

Precision planning asks how well a survey can estimate one quantity. Power
planning asks whether a design can detect a specified difference. State the
comparison first, whether two independent groups, two occasions with overlap,
or a treated-versus-control change.

The `n` reported by `power_prop()` and `power_mean()` is the gross sample size
per group. For `power_did()`, it is the gross sample size per arm per wave. See
[Repeated surveys](repeated-surveys.html) for the distinction between issued
and respondent overlap.

## Detect a difference between proportions

How many units per group are needed to detect a change from 70% to 75% with
80% power and a two-sided 5% significance level?

```{r power-size}
power_prop(
  p1 = 0.70,
  p2 = 0.75,
  power = 0.80,
  alpha = 0.05
)
```

Add design and response assumptions in the same units as the sample size:

```{r power-design}
power_prop(
  p1 = 0.70,
  p2 = 0.75,
  power = 0.80,
  deff = 2,
  resp_rate = 0.85
)
```

The design effect multiplies variance, while the response rate converts the
gross issued size into its expected responding size. Neither adjustment
addresses nonresponse bias.

## Three solve modes

The proportion and mean functions solve for whichever of sample size, power,
or effect size is unspecified:

```{r solve-modes}
# Gross sample size per group
power_prop(p1 = 0.70, p2 = 0.75, power = 0.80, deff = 2)

# Power at 1,500 issued units per group
power_prop(p1 = 0.70, p2 = 0.75, n = 1500, power = NULL, deff = 2)

# Minimum detectable second proportion
power_prop(p1 = 0.70, n = 1500, deff = 2)
```

The minimum detectable effect (MDE) is the smallest difference that reaches
the requested power under the planning assumptions. A one-sided test changes
the critical value and must be justified before looking at results.

For continuous outcomes, supply the anticipated within-group variance:

```{r mean-power}
# Detect a difference of 5 when each group has variance 200
power_mean(var = 200, effect = 5, power = 0.80)

# MDE at 400 issued units per group
power_mean(var = 200, n = 400)
```

## Unequal groups

`ratio` is group 1 size divided by group 2 size (`n1 / n2`). For example,
`ratio = 2` allocates twice as many units to group 1:

```{r unequal-groups}
power_prop(p1 = 0.30, p2 = 0.35, ratio = 2)
power_mean(var = c(80, 120), effect = 5, ratio = 2)
```

When sample sizes are already fixed, supply a length-two `n` vector in group 1,
group 2 order.

## Overlapping occasions

If some respondents are observed at both occasions, their positive outcome
correlation reduces the variance of change:

```{r overlap-power}
power_prop(
  p1 = 0.70,
  p2 = 0.75,
  power = 0.80,
  deff = 2,
  overlap = 0.50,
  overlap_cor = 0.60
)
```

`overlap` is the share of the first occasion's responding sample observed
again. A rotation object describes issued overlap, so convert it using an
explicit response-persistence assumption when response is incomplete.

## Difference-in-differences

`power_did()` plans the interaction contrast

$$
(Y_{T,1} - Y_{T,0}) - (Y_{C,1} - Y_{C,0}),
$$

where $T$ and $C$ denote treated and control arms and 0 and 1 denote baseline
and endline. Supply each arm as `c(baseline, endline)`:

```{r did-power}
# Proportion: treated rises by 5 points while control falls by 2
power_did(
  treat = c(0.50, 0.55),
  control = c(0.50, 0.48),
  outcome = "prop",
  effect = 0.07
)

# Mean outcome with half of each arm observed again
power_did(
  treat = c(50, 55),
  control = c(50, 52),
  outcome = "mean",
  var = 100,
  effect = 3,
  overlap = 0.50,
  overlap_cor = 0.60
)
```

Here `n` is per arm per wave. Unequal treated and control arms are supported.
Supply `n = c(n_treat, n_control)` to evaluate fixed sizes, or
`ratio = n_treat / n_control` when solving for sizes. Within each arm, the
calculation assumes the same issued size at baseline and endline, and it does
not accept four independently chosen cell sizes.

## Alternative proportion methods

Near a boundary, arcsine and log-odds calculations provide transformed-scale
alternatives to the untransformed Wald calculation:

```{r transformed-power}
power_prop(
  p1 = 0.15,
  p2 = 0.18,
  alternative = "one.sided",
  method = "arcsine"
)

power_prop(
  p1 = 0.15,
  p2 = 0.18,
  alternative = "one.sided",
  method = "logodds"
)
```

The methods test differences on different scales and can produce different
operating characteristics. Choose the method to match the planned analysis,
not after comparing which result is smallest.

## Inspect a power curve

The curve shows how many issued units each independent group needs before the
design reaches the requested power:

```{r power-curve, fig.cap="Power to detect a 5 percentage point difference with a design effect of 2 and complete response. The intersection marks the 80 percent power requirement per group.", fig.alt="Power curve against gross sample size per group. Dashed lines identify 80 percent power and the corresponding required sample size for detecting a change from 70 to 75 percent with a design effect of 2."}
power_plan <- power_prop(
  p1 = 0.70,
  p2 = 0.75,
  power = 0.80,
  deff = 2
)
plot(power_plan)
```

For exact values, use `predict()` on supported power results:

```{r power-grid}
power_at_n <- power_prop(
  p1 = 0.70,
  p2 = 0.75,
  n = 500,
  power = NULL,
  deff = 2
)
predict(power_at_n, data.frame(n = seq(500, 2500, by = 500)))
```

## A consistent planning workflow

Suppose the same independent survey design will estimate baseline coverage and
compare baseline with a later round. Store shared assumptions once:

```{r workflow-profile}
design <- svyplan(deff = 2, resp_rate = 0.85)
```

First, size each occasion for a 5 percentage point margin of error and check
the expected precision:

```{r workflow-precision}
occasion <- n_prop(p = 0.70, moe = 0.05, plan = design)
occasion
prec_prop(occasion)
```

Then evaluate the power to detect a change from 70% to 75% when both occasions
issue that many units:

```{r workflow-power}
power_prop(
  p1 = 0.70,
  p2 = 0.75,
  n = as.integer(occasion),
  power = NULL,
  plan = design
)
```

If that power is insufficient, solve for the per-occasion requirement and use
the larger of the precision and power sizes:

```{r workflow-decision}
power_requirement <- power_prop(
  p1 = 0.70,
  p2 = 0.75,
  power = 0.80,
  plan = design
)

max(
  precision = as.integer(occasion),
  power = as.integer(power_requirement)
)
```

The design effect and response rate are planning guesses. Check how the power
requirement moves across plausible values before fixing the sample:

```{r workflow-sensitivity}
predict(
  power_requirement,
  expand.grid(
    deff = c(1.5, 2.0, 2.5),
    resp_rate = c(0.75, 0.85, 0.95)
  )
)
```

The requirement is proportional to the design effect and inversely
proportional to the response rate, so the grid brackets the sample the field
budget must cover.

Meeting a precision target does not automatically meet a power target. This
workflow keeps the design effect, response rate, sample-size units, and group
definition consistent across both calculations. Surveys with several
indicators combine their requirements as shown in
[Multiple indicators and domains](multi-indicator.html). A clustered design
uses variance components and stage costs in place of the scalar design
effect, as in [Multistage survey planning](multistage.html), and a
stratified design adds the steps in
[Stratification and sample allocation](allocation.html).
