---
title: "Power and sensitivity analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Power and sensitivity analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(svyplan)
```
This article covers power calculations, parameter grids, and an end-to-end planning workflow. See [Getting started](svyplan.html) for the core sizing and precision workflow, [repeated surveys](repeated-surveys.html) for overlap definitions, and [multistage planning](multistage.html) for cluster design inputs.

## Power analysis

Precision analysis answers "how precisely can we estimate a level?" Power analysis answers a different question: "can we detect a change or difference between two groups or time points?"

### Sample size for a detectable change

A survey measured vaccination coverage at 70%. How many households are
needed in a follow-up to detect a 5 percentage point increase with 80%
power?

```{r power-n}
power_prop(p1 = 0.70, p2 = 0.75, power = 0.80)
```

With a design effect:

```{r power-n-deff}
power_prop(p1 = 0.70, p2 = 0.75, power = 0.80, deff = 2.0)
```

### Three solve modes

`power_prop()` and `power_mean()` each solve for whichever of `n`, `power`, or the effect size is left unspecified:

```{r power-modes}
# Solve for n (default when p2 and power given)
power_prop(p1 = 0.70, p2 = 0.75, power = 0.80, deff = 2.0)

# Solve for power (set power = NULL)
power_prop(p1 = 0.70, p2 = 0.75, n = 1500, power = NULL, deff = 2.0)

# Solve for MDE (omit p2)
power_prop(p1 = 0.70, n = 1500, deff = 2.0)
```

### Panel surveys

The power family takes the same `overlap` and `overlap_cor` as `n_change()`, and `design_overlap()` supplies the first of them from a rotation schedule. If the follow-up resamples a fraction of the original households and the between-round correlation is known, the required sample size drops:

```{r power-panel}
power_prop(p1 = 0.70, p2 = 0.75, power = 0.80, deff = 2.0,
           overlap = 0.50, overlap_cor = 0.6)
```

### Continuous outcomes

The same three solve modes work for means:

```{r power-mean}
# Sample size to detect a difference of 5 (within-group var = 200)
power_mean(200, effect = 5)

# MDE with n = 400 per group
power_mean(200, n = 400)
```

### Difference-in-differences

`power_did()` handles two-group, two-period DiD designs. Specify treated and control group outcomes as `c(baseline, endline)` vectors:

```{r power-did}
# Proportion outcome: treated improves, control stable
power_did(treat = c(0.50, 0.55), control = c(0.50, 0.48),
          outcome = "prop", effect = 0.07)

# Mean outcome with panel overlap
power_did(treat = c(50, 55), control = c(50, 52),
          outcome = "mean", var = 100, effect = 3,
          overlap = 0.5, overlap_cor = 0.6)
```

### Unequal groups and alternative methods

Both `power_prop()` and `power_mean()` support unequal group sizes via `ratio` or explicit `n = c(n1, n2)`:

```{r power-unequal}
# 2:1 allocation ratio
power_prop(p1 = 0.30, p2 = 0.35, ratio = 2)

# Unequal group variances
power_mean(c(80, 120), effect = 5)
```

Near a boundary, the arcsine and log-odds calculations provide transformed-scale alternatives to the untransformed Wald calculation. Their operating characteristics still depend on the design and planning values:

```{r power-methods}
power_prop(p1 = 0.15, p2 = 0.18, alternative = "one.sided",
           method = "arcsine")

power_prop(p1 = 0.15, p2 = 0.18, alternative = "one.sided",
           method = "logodds")
```

### Power curve

`plot()` draws the power-vs-sample-size curve with reference lines at the solved point:

```{r power-plot, fig.width=7, fig.height=4}
pw <- power_prop(p1 = 0.70, p2 = 0.75, power = 0.80, deff = 2.0)
plot(pw)
```

## Sensitivity analysis

How sensitive is a sample size to assumptions about the design effect, response rate, or prevalence? The `predict()` method evaluates any `svyplan` result at new parameter combinations.

### Varying design parameters

```{r predict-n}
x <- n_prop(p = 0.3, moe = 0.05, deff = 1.5)
predict(x, expand.grid(
  deff = c(1.0, 1.5, 2.0, 2.5),
  resp_rate = c(0.8, 0.9, 1.0)
))
```

### Cluster budget scenarios

```{r predict-cluster}
cl <- n_cluster(stage_cost = c(500, 50), icc = 0.05, budget = 100000)
predict(cl, data.frame(budget = c(50000, 100000, 150000, 200000)))
```

### Power curves

```{r predict-power}
pw <- power_prop(p1 = 0.30, p2 = 0.35, n = 500, power = NULL)
predict(pw, data.frame(n = seq(200, 1000, 200)))
```

### Precision by sample size

```{r predict-prec}
pr <- prec_prop(p = 0.3, n = 400)
predict(pr, data.frame(n = c(100, 200, 400, 800, 1600)))
```

## Putting it all together

A typical planning sequence for a national household survey ties several `svyplan` functions together. A `svyplan()` profile captures the shared design assumptions:

1. Set shared design parameters once with `svyplan()`
2. Define indicators and precision targets per domain
3. Compute sample sizes with `n_multi()`
4. Evaluate achieved precision with `prec_multi()`
5. Run sensitivity analysis with `predict()`
6. If frame data is available, estimate variance components with `varcomp()`
7. Optimize the multistage allocation with `n_cluster()` or `n_multi_cluster()`
8. Determine strata boundaries with `strata_bound()`
9. Check that the design has adequate power with `power_prop()` / `power_mean()`

```{r workflow}
# 1: shared design assumptions
design <- svyplan(deff = 2.0, resp_rate = 0.85)

# 2-3: multi-indicator sample size
targets <- data.frame(
  name = c("stunting", "vaccination", "anemia"),
  p    = c(0.25, 0.70, 0.12),
  moe  = c(0.05, 0.05, 0.03),
  deff = c(2.0, 1.5, 2.5)
)
result <- n_multi(targets)
result

# 4: achieved precision for each indicator
prec_multi(result)

# 5: how sensitive is the binding indicator to the response rate?
predict(
  n_prop(p = 0.25, moe = 0.05, plan = design),
  data.frame(resp_rate = c(0.7, 0.8, 0.9, 1.0))
)

# 9: can we detect a 5pp decline in stunting with this sample?
power_prop(
  p1 = 0.25, p2 = 0.20,
  n = as.integer(result), power = NULL,
  plan = design
)
```

Sample-size, cluster, power, and strata-boundary results support `as.integer()` and `as.double()`. For sample-size results, these return the rounded-up and exact values of *n*, respectively, which makes it straightforward to pass results between functions or into other packages.
