---
title: "Precision targets for proportions, means, and ratios"
description: >
  Choose an estimand, express its precision target, and interpret the planned
  interval.
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Precision targets for proportions, means, and ratios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(svyplan)
```

This article helps you choose the quantity to estimate and the way to state
its precision. Begin with [Getting started](svyplan.html) for the basic
sample-size and precision workflow.

## Four ways to state precision

`svyplan` reports four related quantities:

| | Absolute | Relative |
|:---|:---|:---|
| Sampling uncertainty | standard error (`se`) | coefficient of variation (`cv`) |
| Interval half-width | margin of error (`moe`) | relative margin of error (`rmoe`) |

For a proportion, `moe = 0.05` means 5 percentage points. It does not mean 5%
of the proportion. The latter is a relative margin of error. At `p = 0.70`,
`rmoe = 0.05` is an absolute margin of error of 0.035.

```{r precision-targets}
n_prop(p = 0.70, moe = 0.035)
n_prop(p = 0.70, rmoe = 0.05)
n_prop(p = 0.70, cv = 0.05)
```

The coefficient of variation describes the sampling variance. The two margin
of error targets also depend on the confidence level and interval method. For
a two-sided Wald interval, `rmoe` is the interval quantile times `cv`.

## Proportions and interval methods

The default Wald calculation is the familiar normal approximation:

```{r proportion-basic}
n_prop(p = 0.30, moe = 0.05)
```

Four interval methods are available:

| Method | Planning interpretation |
|:---|:---|
| `"wald"` | Symmetric normal approximation (the default). |
| `"wilson"` | Wilson score interval, bounded to the parameter space. |
| `"logodds"` | Interval formed on the log-odds scale and transformed back. |
| `"beta"` | Korn–Graubard interval with optional design degrees of freedom. |

Their differences are most visible near 0 or 1 or with few sampled clusters:

```{r proportion-methods}
sapply(c("wald", "wilson", "logodds", "beta"), function(method) {
  result <- prec_prop(
    p = 0.02,
    n = 900,
    deff = 2,
    df = 25,
    method = method
  )
  c(se = result$se, cv = result$cv, moe = result$moe, rmoe = result$rmoe)
})
```

The sampling standard error and CV do not change across columns. The interval
half-widths do because each method constructs a different interval. Choose the
method that matches the intended reporting procedure. Use `df = design_df(...)`
when the planned variance estimator has few degrees of freedom.

The log-odds limits bound the proportion, or equivalently its odds. They do
not bound an odds ratio, and the method does not size a logistic-regression
coefficient.

## Means and totals

For a mean, supply a population variance or standard deviation estimated from
a previous survey, pilot, or comparable population:

```{r mean-basic}
n_mean(var = 40000, moe = 30)
n_mean(sd = 200, mu = 800, cv = 0.05)
```

A target CV or relative margin of error also needs the anticipated mean
because both are relative to its magnitude.

A total is a mean multiplied by a known population size. It therefore does
not need a separate sizing function. A CV target is unchanged and an absolute
margin of error for a total must be divided by `N` before sizing the mean:

```{r total}
population <- 10000
total_moe <- 500000

n_mean(
  var = 2500,
  moe = total_moe / population,
  N = population
)
```

This conversion assumes that the population size used to define the total is
known for planning.

## Ratios of two totals

`n_ratio()` plans for

$$
R = T_y / T_x,
$$

where the numerator and denominator are observed on the same sampled units.
Examples include consumption per person and yield per hectare. Supply the
anticipated ratio, the coefficient of variation of each component, and their
unit-level correlation:

```{r ratio-basic}
n_ratio(
  r = 420,
  cv_num = 1.20,
  cv_den = 0.45,
  component_cor = 0.65,
  cv = 0.05
)
```

This estimand is different from the mean of unit ratios $y_i/x_i$, which is
planned with `n_mean()`. A coverage rate among an eligible group is usually a
domain proportion. First calculate the sample required *within* the eligible
domain, then use its expected share to calculate the overall sample. The
[multiple-indicator and domain guide](multi-indicator.html) demonstrates that
step.

The ratio calculation uses the relative variance of the linearized variable
$e = y - Rx$. A stronger positive association between the components can make
the ratio more stable:

```{r ratio-correlation}
ratio_plan <- n_ratio(
  r = 420,
  cv_num = 1.20,
  cv_den = 0.45,
  component_cor = 0.65,
  cv = 0.05
)

predict(
  ratio_plan,
  data.frame(component_cor = c(0.30, 0.65, 0.90))
)
```

Margin-of-error targets are stated on the ratio's own scale. Here `moe = 20`
asks for an interval of plus or minus 20 currency units per person, and
`rmoe` would state the same target relative to `r`. Design effect and
response rate enter as for a mean:

```{r ratio-targets}
n_ratio(
  r = 420,
  cv_num = 1.20,
  cv_den = 0.45,
  component_cor = 0.65,
  moe = 20,
  deff = 1.3,
  resp_rate = 0.85
)
```

`prec_ratio()` evaluates a ratio design at a fixed sample size. Passing it the
sizing result carries the same assumptions and returns the target CV:

```{r ratio-precision}
prec_ratio(ratio_plan)
```

When the two component CVs are nearly equal and their correlation is near 1,
the assumed ratio is nearly constant and the required size falls toward zero.
That size depends on a correlation known to several decimals, and
`n_ratio()` warns about it:

```{r ratio-degenerate, warning = TRUE}
n_ratio(
  r = 420,
  cv_num = 1.20,
  cv_den = 1.19,
  component_cor = 0.9999,
  cv = 0.05
)
```

### Using a known auxiliary total

A ratio can also be used to estimate a different quantity. If the auxiliary
total $T_x$ is known from the frame, the estimator
$\widehat{T}_y = T_x\widehat{R}$ targets the outcome total. For example, known
total cultivated area can turn estimated yield per hectare into an estimated
harvest total. Multiplication by a known constant preserves relative
precision, while an absolute margin of error must be multiplied by $T_x$.

The first-order variance uses residuals $e_i = y_i - R x_i$ and the planned
sample design [@cochran1977]. A generic regression variance reduction such as
$1-\rho^2$ is not automatically the variance reduction of this ratio estimator,
because that reduction depends on the assumed relation between the two
variables. The calculation also omits higher-order ratio bias. Document those
approximations when using `n_ratio()` for a known-total auxiliary estimator.

The same residuals give the planning input when the ratio estimator sharpens
the estimate of the mean of $y$. Fit the ratio to pilot data, take the
residual variance, and pass it to `n_mean(var = )` with the mean of the
analysis variable:

```{r ratio-residual}
set.seed(2)
area <- rlnorm(200, 3, 0.5)
harvest <- 2.5 * area + rnorm(200, 0, 15)
pilot_ratio <- mean(harvest) / mean(area)
residual <- harvest - pilot_ratio * area

n_mean(var = var(harvest), mu = mean(harvest), cv = 0.02)
n_mean(var = var(residual), mu = mean(harvest), cv = 0.02)
n_ratio(
  r = pilot_ratio,
  cv_num = sd(harvest) / mean(harvest),
  cv_den = sd(area) / mean(area),
  component_cor = cor(area, harvest),
  cv = 0.02
)
```

The first call ignores the auxiliary variable. The second sizes the ratio
estimator of the mean. With a CV target it gives the same size as
`n_ratio()` on the pilot's component CVs and correlation, because the two
estimators have the same relative variance.

If the auxiliary total is estimated in an earlier sampling phase, it is no
longer a fixed multiplier. Its uncertainty and covariance must be included.
See the [two-phase method boundary](two-phase.html#method-boundary).

### Ratios in a cluster design

For a multistage survey, estimate the variance components of the linearized
variable. The relevant homogeneity is not generally the homogeneity of the
numerator or denominator:

```{r clustered-ratio}
set.seed(1)
psu <- rep(1:30, each = 20)
x <- rlnorm(600, 1, 0.4)
y <- 3 * x + rnorm(600, 0, 0.8)

ratio <- mean(y) / mean(x)
linearized <- y - ratio * x
components <- suppressWarnings(varcomp(linearized, stage_id = list(psu)))

n_cluster(
  stage_cost = c(500, 50),
  icc = components$icc,
  unit_relvar = var(linearized) / mean(y)^2,
  var_ratio = components$var_ratio,
  cv = 0.01
)
```

The linearized variable has mean zero by construction, so `varcomp()` warns
that its relative variances are infinite. Its `icc` and `var_ratio` remain
usable because they do not depend on the mean and `unit_relvar` is supplied
separately on the ratio estimator's scale.

## Precision across sample sizes

`prec_prop()` and the other `prec_*()` functions evaluate a fixed sample
size. `predict()` over `n` shows how precision changes as the sample grows:

```{r precision-by-n}
fixed_n <- prec_prop(p = 0.30, n = 400)
predict(fixed_n, data.frame(n = c(100, 200, 400, 800, 1600)))
```

Each fourfold increase in `n` halves the standard error and the margin of
error.

## Solve for a reportable level

`prec_prop()` can solve for the smallest proportion that a fixed sample can
estimate at a target relative precision:

```{r inverse-proportion}
floor <- prec_prop(n = 1500, cv = 0.10, N = 2e6)
floor
floor$params$p
```

Every larger proportion has a smaller CV under the same design, and every
smaller one is too rare to report at that precision. This makes the result a
prospective reporting threshold. The printed `expected_cases` is the expected
number of positive cases at that proportion, so the precision threshold can
be read beside the `min_cases` floor of `n_prop()`.

`predict()` traces the threshold across sample sizes, with the target CV held
fixed:

```{r inverse-proportion-grid}
predict(floor, expand.grid(n = c(500, 1500, 5000)))
```

A `cv` target fixes the sampling variance and gives the same threshold under
every interval method. An `rmoe` target fixes the relative half-width of the
interval the survey will report, so its threshold moves with `method`:

```{r inverse-proportion-rmoe}
prec_prop(n = 1500, rmoe = 0.20, method = "wald", N = 2e6)$params$p
prec_prop(n = 1500, rmoe = 0.20, method = "beta", N = 2e6)$params$p
```

Two properties bound the `rmoe` search. The Wilson and Korn–Graubard
half-widths do not shrink to zero as `p` approaches 1, so their relative
margin of error has a positive floor. A target below that floor is refused,
and the error names the floor:

```{r inverse-proportion-floor, error = TRUE}
prec_prop(n = 1500, rmoe = 0.001, method = "wilson", N = 2e6)
```

The back-transformed log-odds half-width turns upward once the logit spread
exceeds `logit(p)`. The turn lies near `p = 0.999` at `n = 1500` and near
`p = 0.97` at `n = 30`. Under `"logodds"` the search stops at that turn and
returns the lower root.

`prec_mean()` takes `cv` the same way and returns the magnitude of the mean.
The standard error of a mean does not involve the mean, so this inversion is
a division rather than a search:

```{r inverse-mean}
prec_mean(var = 25, n = 1500, cv = 0.10)$params$mu
```

## Expected confidence intervals

`confint()` extracts the expected interval from supported size and precision
results:

```{r expected-intervals}
confint(n_prop(p = 0.30, moe = 0.05))
confint(n_mean(var = 100, mu = 50, moe = 2))
confint(ratio_plan)
```

The `level` argument sets the confidence level, and a precision result
gives its interval the same way:

```{r expected-intervals-level}
confint(fixed_n)
confint(fixed_n, level = 0.99)
```

A repeated-survey result has an interval too. For the average level pooled
across two occasions:

```{r expected-intervals-pooled}
confint(prec_pooled(var = 100, n = 300, occasions = 2, mu = 50))
```

The center for a mean must be supplied because variance and margin of error do
not determine it. For a proportion, Wald endpoints are truncated to [0, 1]
when needed. The other methods stay within the parameter space by
construction. Cluster, allocation, power, and multi-indicator results have
no single interval and are not supported by `confint()`.

These calculations follow standard survey-sampling variance approximations
[@cochran1977; @valliant2018]. Document the source and uncertainty of every
planning value, then compare plausible alternatives with `predict()`.

## References
