---
title: "Planning repeated surveys and panels"
description: >
  Relate precision for levels, changes, and pooled estimates to overlap,
  panel retention, rotation, and fieldwork schedules.
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Planning repeated surveys and panels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(svyplan)
```

A repeated survey can target a level at one occasion, a change between
occasions, or an average across occasions. Sharing units across occasions has
different consequences for those estimands. This article connects those
precision questions to panel recruitment and fieldwork schedules.

See [Getting started](svyplan.html) for single-occasion sizing and
[Power](power.html) for detecting a change.

## Terms used here

| Term | Meaning |
|:---|:---|
| Occasion | One calendar round of the survey. |
| Wave | A unit's interview number within its participation period. |
| Cohort | Units entering the design together. |
| Retention | Conditional probability of remaining from one wave to the next. |
| Issued overlap | Share of selected units common to two occasions. |
| Respondent overlap | Share of the first occasion's respondents measured again. |

Keep issued and respondent overlap separate whenever response is incomplete.

## Precision for change

`n_change()` and `prec_change()` plan a change in a mean or proportion. Start
with independent samples at the two occasions:

```{r change-independent}
n_change(p = c(0.30, 0.36), moe = 0.02)
```

Measuring some of the same responding units at both occasions reduces the
variance of a difference when their measurements are positively correlated:

```{r change-overlap}
change_plan <- n_change(
  p = c(0.30, 0.36),
  moe = 0.02,
  overlap = 0.75,
  overlap_cor = 0.50
)
change_plan
prec_change(change_plan)
```

`overlap` is a design property. `overlap_cor` is specific to the outcome and
should come from an earlier round or other defensible evidence. A full panel
with zero outcome correlation provides no variance reduction for change.

## Precision for an average across occasions

An annual average of quarterly estimates is a different estimand. Positive
covariance is subtracted in a difference but added in an average:

```{r pooled-basic}
prec_pooled(var = 100, n = 500, occasions = 4)
prec_pooled(
  var = 100,
  n = 500,
  occasions = 4,
  overlap = 0.75,
  cor_decay = 0.80
)
```

In the second calculation, adjacent occasions share 75% of units and outcome
correlation decays as $0.8^m$ at lag $m$. Under the ordinary case where
overlap exceeds the sampling fraction, sharing units sharpens change estimates
but reduces the information gained by averaging occasions. A single-occasion
level depends on that occasion's sample size and is unchanged.

The length of participation is the design lever that sets the overlap. Take
the overlaps from
[rotations](#derive-overlap-from-a-rotation) that keep a unit in sample for
2, 4, 6, or 8 consecutive occasions, and compare the standard errors at the
same occasion sample over an eight-occasion horizon:

```{r pooled-tradeoff}
horizon <- 8
arms <- function(life) {
  rot <- design_rotation(as.character(life))
  ov <- design_overlap(rot, max_lag = horizon - 1)
  data.frame(
    span = attr(rot, "life"),
    interviews = attr(rot, "n_occasion"),
    overlap_1 = round(as.numeric(ov)[1], 3),
    se_level = round(prec_mean(var = 100, n = 1000, N = 1e5)$se, 4),
    se_change = round(prec_change(
      var = 100, n = 1000, N = 1e5,
      overlap = ov[1], overlap_cor = 0.80
    )$se, 4),
    se_pooled = round(prec_pooled(
      var = 100, n = 1000, N = 1e5,
      occasions = horizon, overlap = ov, cor_decay = 0.80
    )$se, 4)
  )
}
do.call(rbind, lapply(c(2, 4, 6, 8), arms))
```

Reading down the columns, longer participation sharpens the change
estimate and blunts the pooled average. The single-occasion level does not
move, because it depends only on that occasion's sample size. It is a
reference line rather than a third position on the trade-off, which runs
between change and the average. `span` counts the occasions from a unit's
first interview to its last, and `interviews` counts how often it is asked.
They coincide here because each rotation is uninterrupted.

The table assumes full response. That is the only case in which
`prec_pooled()` accepts a `design_overlap()` result directly. Below full
response it refuses the issued overlap and asks for the respondent overlap
you expect.

Neither estimand dominates, so a design serving both change and pooled
estimates takes the larger of the `n_change()` and `n_pooled()` requirements:

```{r pooled-requirement}
change_required <- n_change(
  var = 100,
  moe = 1.5,
  overlap = 0.75,
  overlap_cor = 0.80
)
pooled_required <- n_pooled(
  var = 100,
  moe = 1.5,
  occasions = 4,
  overlap = 0.75,
  cor_decay = 0.80
)

requirements <- c(
  change = as.double(change_required),
  pooled = as.double(pooled_required)
)
requirements
max(requirements)
```

`?prec_pooled` documents the finite-population boundary where overlap falls
below the sampling fraction. The function checks the complete covariance
matrix rather than returning a variance from an invalid combination.

```{r pooled-diagnostic, error = TRUE}
prec_pooled(
  var = 100,
  n = 67000,
  occasions = 8,
  N = 1e5,
  overlap = design_overlap(design_rotation("4"), max_lag = 7),
  cor_decay = 0.80
)
```

This diagnostic identifies the lag at which the covariance changes sign and
whether the rotation would also exhaust the population over the horizon.

## Derive overlap from a rotation

`design_rotation()` records the occasions a cohort is in and out of sample.
The string `"4-8-4"` means four occasions in sample, eight out, and four more
in sample before leaving permanently:

```{r cps-overlap}
cps_rotation <- design_rotation("4-8-4")
cps_overlap <- design_overlap(cps_rotation)
cps_overlap[c(1, 12)]
```

The results are the 75% month-to-month and 50% year-to-year issued overlaps of
the Current Population Survey design [@cps2019]. The specification describes
one cohort's finite participation pattern. It is not an endlessly repeating
cycle.

A rotation records two measures of respondent burden. The `life` attribute is
the span from a unit's first interview to its last, and `n_occasion` counts
the interviews. They separate when the pattern has a rest period:

```{r rotation-burden}
c(
  span = attr(cps_rotation, "life"),
  interviews = attr(cps_rotation, "n_occasion")
)
```

The `"4-8-4"` pattern spans sixteen occasions and asks each unit eight times.

Use the overlap for the lag being estimated:

```{r overlap-by-lag}
c(
  consecutive = n_change(
    p = c(0.30, 0.36), moe = 0.02,
    overlap = cps_overlap[1], overlap_cor = 0.50
  )$n,
  annual = n_change(
    p = c(0.30, 0.36), moe = 0.02,
    overlap = cps_overlap[12], overlap_cor = 0.50
  )$n
)
```

Two rotation notations are supported. Spell notation such as `"4-8-4"`
counts consecutive occasions in and out. Binary notation such as
`"1-1-0-0-1-1"` records status at every occasion. A string containing only
ones is ambiguous, so write `"3"` for three consecutive occasions or include
zeros in a binary pattern.

Use the chart to check when cohorts return and when the design reaches its
steady sample before calculating overlap at a particular lag:

```{r rotation-chart, fig.cap="A two-in, two-out, two-in rotation reaches its steady sample at occasion six. The cohort rows show which interviews contribute to overlap between occasions.", fig.alt="Cohort rotation chart with one row per cohort and one column per occasion. Each cohort is interviewed twice, rests for two occasions, and is interviewed twice more. The total row reaches its steady sample at occasion six."}
plot(design_overlap(design_rotation("1-1-0-0-1-1")))
```

`design_overlap()` describes issued samples. The sizing and power functions
define `overlap` among respondents. They coincide at full response. Otherwise,
conversion requires an assumption about response persistence. Under response
independent across occasions at rate $r$, an issued overlap $f$ yields about
$fr$ respondent overlap and persistent response moves it closer to $f$.

## Recruit a panel under attrition

`n_panel()` takes a scalar target from `n_prop()` or `n_mean()` and calculates
how many units to recruit so enough respondents remain at a target wave. The
rates below are illustrative: 72.8% response at recruitment followed by four
wave-to-wave retention rates.

```{r fixed-panel}
target <- n_prop(p = 0.50, moe = 0.031)
panel <- n_panel(
  target,
  retention = c(0.878, 0.963, 0.936, 0.956),
  resp_rate = 0.728
)
panel
panel$waves[, c("wave", "q", "loss_share", "n_resp", "moe")]
```

The output reports the units to recruit, expected respondents at each wave,
and their expected precision. Initial response and later retention are
separate because spreading one average rate across waves changes the required
recruitment. The model treats loss as permanent. A study that allows units to
return after a missed wave needs a different retention model.

`assurance` requests a probability of retaining the required responding
sample rather than planning only at its expectation:

```{r panel-assurance}
n_panel(
  target,
  retention = c(0.878, 0.963, 0.936, 0.956),
  resp_rate = 0.728,
  assurance = 0.90
)$n_assured
```

`prec_panel()` runs the calculation in the other direction for a recruitment
already fixed by budget or operations.

## Rotating panels and startup

For a rotating panel, an equal cohort enters each occasion and the estimate
combines the live cohorts:

```{r rotating-panel}
rotating_panel <- n_panel(
  target,
  retention = c(0.878, 0.963, 0.936, 0.956),
  resp_rate = 0.728,
  design = "rotating"
)
c(
  entrants_per_occasion = rotating_panel$n_entrants,
  standing_issue = rotating_panel$n_in_sample
)
```

`n_entrants` is the new cohort issued at each mature occasion.
`n_in_sample` is the standing issued sample across all live cohorts. Do not use
one as a substitute for the other.

The mature design can start gradually or immediately. A gradual start recruits
one cohort per occasion. An immediate start fills the sample on the first
occasion with cohorts assigned different remaining participation lengths:

```{r panel-startup}
immediate_panel <- n_panel(
  target,
  retention = c(0.878, 0.963, 0.936, 0.956),
  resp_rate = 0.728,
  design = "rotating",
  start = "immediate"
)
immediate_panel$launch[, c("period", "n_in_sample", "n_resp", "moe")]
```

Membership is at its mature level immediately, but response composition is
not, because every unit starts at wave 1. `$launch_waves` shows the wave
composition as the design settles.

## Build a fieldwork schedule

`design_schedule()` turns the panel and rotation into an operational manifest
without selecting records:

```{r fieldwork-schedule}
field_plan <- design_schedule(
  immediate_panel,
  rotation = design_rotation("5"),
  horizon = 8,
  horizon_policy = "continuing",
  refreshment = "entrant_register",
  rounding = "ceiling"
)
field_plan
field_plan$components[, c(
  "cohort", "entry_wave", "operational_issue", "panels", "frame_role"
)]
```

The horizon policy states whether interviews promised after the reporting
window continue, cohort participation is truncated, or intake closes early.
An entrant register assumes later frames contain newly eligible units. A
whole-vintage refreshment can also be declared, but the planning object does
not infer analysis weights across frame vintages.

The immediate-start chart helps schedule replacement cohorts. It describes
issued membership. Use the launch table above for the response composition
and precision during startup:

```{r startup-chart, fig.cap="An immediate start fills the issued sample from the first occasion using cohorts with different remaining participation lengths. New cohorts replace departing cohorts over time.", fig.alt="Immediate-start panel chart showing cohorts with different remaining participation lengths in the first occasion and subsequent cohorts entering as earlier cohorts leave."}
plot(design_overlap(design_rotation("5")), start = "immediate")
```

`design_rotation()` records a rotation already chosen. It does not optimize a
pattern for target overlap and respondent burden. Compare candidate patterns
by passing their lag-specific overlaps to the precision functions above.

## References
