---
title: "Planning two-phase surveys and nonresponse follow-up"
description: >
  Allocate screening and measurement phases, account for stage-specific
  response, and compare with a single-phase design.
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Planning two-phase surveys and nonresponse follow-up}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(svyplan)
```

In a two-phase design, phase 1 collects a relatively inexpensive variable
that is absent from the frame. Phase 2 subsamples the classified phase-1 units
and measures the outcome of interest. Double sampling for stratification and
subsampling nonrespondents for follow-up both have this structure.

See [Stratification and sample allocation](allocation.html) for a one-phase
allocation and [Joint allocation](joint-allocation.html) for several
indicators and domains.

## Allocate two phases

The frame has one row per phase-2 stratum. `N` is the stratum population,
`sd` and `mean` describe the outcome measured in phase 2, and `unit_cost` is
the incremental phase-2 cost:

```{r two-phase-frame}
frame <- data.frame(
  stratum = c("A", "B", "C", "D"),
  N = c(350000, 250000, 250000, 150000),
  sd = c(12, 25, 8, 40),
  mean = c(40, 70, 35, 90),
  unit_cost = c(2, 5, 1, 9)
)

two_phase <- n_twophase(
  frame,
  phase1_cost = 1,
  budget = 50000,
  N = sum(frame$N)
)
two_phase
```

The population contains one million units. The frame's `N` column determines
relative stratum shares. The separate function argument `N = sum(frame$N)`
supplies the population size for the phase-1 finite population correction and
sample-size bound. It is not inferred from the column. If only relative sizes
are known, leave that argument at its default `Inf` and document the
large-population approximation.

The output distinguishes units issued at phase 1, the phase-2 allocation, and
the expected precision. `nu` is the fraction of the phase-1 units in a stratum
issued to phase 2. Its relative pattern reflects variability and phase-2
costs. Its overall scale also depends on how much variation phase-1
stratification explains.

Every result includes a single-phase comparator because a second phase is not
automatically more efficient. Screening can cost more than it saves when the
phase-1 variable explains little or costs too much to measure.

Compare the designs at the same budget:

```{r single-phase-comparison}
data.frame(
  design = c("Two phases", "Single phase"),
  cv = c(two_phase$cv, two_phase$single_phase$cv),
  cost = c(two_phase$cost, two_phase$single_phase$cost)
)
```

The design with the smaller CV uses the budget more efficiently under these
assumptions. The sizes need different interpretations, because phase 1 screens
units whereas the single-phase comparator measures the outcome directly.

## Design effects and response at each phase

The arguments refer to different pieces of the variance:

| Input and location | Applies to |
|:---|:---|
| Argument `phase1_deff` | Between-stratum component estimated through phase 1. |
| Frame column `deff` | Within-stratum residual measured in phase 2. |
| Argument `resp_rate` | Phase-1 response or successful classification. |
| Frame column `resp_rate` | Phase-2 completion within each stratum. |
| Arguments `single_deff` and `single_resp_rate` | Single-phase comparator. |

Design effects multiply their variance components. Each variance component is
divided by its corresponding response rate, so lower response reduces
effective information and increases variance for a given issued sample.
Applying one common factor to the whole variance represents a different design.

```{r two-phase-design}
screened <- transform(
  frame,
  deff = c(1.2, 2.5, 0.8, 1.7),
  resp_rate = c(0.90, 0.70, 0.85, 0.60)
)

two_phase_adjusted <- n_twophase(
  screened,
  phase1_cost = 1,
  budget = 50000,
  N = sum(frame$N),
  phase1_deff = 2.2,
  resp_rate = 0.80,
  single_deff = 2.9,
  single_resp_rate = 0.80
)
two_phase_adjusted
```

All counts are issued units. Convert a phase cost quoted per completed
interview before supplying it. For example, a cost with contact and completion
components is `contact_cost + resp_rate * completion_cost` per unit issued.

Response-rate adjustments assume response is ignorable within the supplied
strata. They do not remove nonresponse bias.

## Check the fieldwork design

`$operational` contains reconciled whole-unit phase sizes and stratum takes.
Calculate expected responding counts from its issue counts and the response
rates. In budget mode, the construction remains inside the budget. In
target-CV mode, it rounds toward meeting the target.

```{r operational}
two_phase_adjusted$operational
c(
  continuous_cv = prec_twophase(two_phase_adjusted)$cv,
  fieldwork_cv = two_phase_adjusted$operational$cv
)
```

`prec_twophase()` applied to the result evaluates the continuous design. The
operational CV describes the whole-unit allocation to field. Its expected
respondents are:

```{r operational-respondents}
c(
  phase1 = two_phase_adjusted$operational$n["n_phase1"] * 0.80,
  phase2 = sum(two_phase_adjusted$operational$n_int * screened$resp_rate)
)
```

Use the operational result as a unit. Rounding each phase or stratum
independently can produce an infeasible plan.

## Fix an existing phase-1 sample

Sometimes phase 1 is already complete or set by operational capacity. Supply
its issued size with `n_phase1`:

```{r fixed-phase-one}
fixed_phase_one <- n_twophase(
  frame,
  phase1_cost = 1,
  budget = 50000,
  N = sum(frame$N),
  n_phase1 = 20000
)
fixed_phase_one
```

The phase-2 relative allocation retains its variance-cost form, while its
scale is determined by the budget remaining after phase 1. Fixing phase 1 can
only match or be less efficient than allowing the optimizer to choose it
within the same model.

A fixed phase-1 sample imposes a precision floor. Even carrying every
successfully classified unit into phase 2 cannot recover information that
phase 1 did not collect. If a requested CV lies below that floor,
`n_twophase()` returns an error describing the infeasible target. The detailed
formula and component definitions are documented in `?n_twophase`.

## Plan for a probability of meeting the target

Realized respondent counts can fall below their expectation. `assurance`
reports additional issue counts that meet each required responding count with
at least the requested probability under a binomial response model. Phase-2
assurance is **marginal by stratum and conditional on the phase-1 pool**. It
does not guarantee that all stratum targets are met simultaneously, and random
phase-1 composition can leave a stratum short of its expected pool.

To isolate phase-1 response uncertainty, return to the unclustered `frame`,
with 80% phase-1 response and complete phase-2 response:

```{r assurance}
assured_plan <- n_twophase(
  frame,
  phase1_cost = 1,
  cv = 0.03,
  N = sum(frame$N),
  resp_rate = 0.80,
  assurance = 0.90
)
assured_plan$operational[c(
  "n", "cost", "assured_phase1", "assured", "pool", "assured_cost"
)]
```

`n` and `cost` describe the expected-response plan. `assured_phase1` is the
phase-1 issue with assurance, `assured` gives the phase-2 issues by stratum,
and `assured_cost` is their combined cost. Here complete phase-2 response
requires no extra phase-2 issue. With incomplete phase-2 response, assurance
can increase both phases and can exceed the budget of the expected design.

If the assured phase-2 issue exceeds the pool supplied by phase 1, enlarge
phase 1 or revise the response assumptions. A simultaneous guarantee needs an
explicit treatment of all strata and the random phase-1 pool. A marginal
level of 0.90 alone does not provide one.

## Nonresponse follow-up

Subsampling nonrespondents is the same allocation structure with two strata.
Initial respondents are already measured, so they have zero additional cost
and are all retained. Only initial nonrespondents are subsampled:

```{r nonresponse-follow-up}
initial_response <- 0.50

follow_up_frame <- data.frame(
  stratum = c("respondents", "nonrespondents"),
  N = c(initial_response, 1 - initial_response),
  sd = c(1, 1),
  unit_cost = c(0, 200),
  take_all = c(TRUE, FALSE)
)

follow_up <- n_twophase(
  follow_up_frame,
  phase1_cost = 50,
  budget = 100000,
  mu = 1,
  single_cost = 50 / initial_response
)
follow_up
```

`single_cost` states the cost of one completed interview without follow-up. A
zero phase-2 cost for respondents means “already measured,” not “free.” Leave
the function-level `resp_rate` at 1 here. The strata already encode response
status, so another phase-1 response adjustment would count the same loss twice.

The design decides how much follow-up effort to allocate. Selecting particular
cases using response propensities or live fieldwork information is a separate
operational decision.

## Method boundary

`n_twophase()` covers a phase-1 categorical variable used for stratification
or response status. It does not implement two-phase regression or ratio
estimators based on a continuous auxiliary variable. The variance of those
estimators has a component from each phase, so a single-phase calculation
does not carry over.

Suppose phase 1 is a simple random sample of $n_1$ units from a population of
$N$, measuring the auxiliary variable $x$, and phase 2 is a simple random
subsample of $n_2$ of those units, measuring the study variable $y$. Let
$S_y^2$ be the population variance of $y$, and $S_e^2$ the variance of the
residuals from the regression of $y$ on $x$. To first order, the regression
estimator of the mean of $y$ has variance [@brus2022, section 11.2]

$$
S_y^2\left(\frac{1}{n_1}-\frac{1}{N}\right)
+ S_e^2\left(\frac{1}{n_2}-\frac{1}{n_1}\right).
$$

The familiar multiplier $1-\rho^2$ on $S_y^2$, with $\rho$ the correlation
between $x$ and $y$, applies alone only when the population mean of $x$ is
known. Here that mean is estimated from phase 1. With $S_e^2 = (1-\rho^2)S_y^2$
and negligible sampling fractions, the variance becomes

$$
S_y^2\left\{\frac{1-\rho^2}{n_2} + \frac{\rho^2}{n_1}\right\},
$$

so a design effect for this estimator would depend on both phase sizes. Record
this distinction rather than treating a single-phase `n_ratio()` calculation
as a two-phase allocation. The cost and precision trade-off between the two
phases is not implemented.

## References
